---
title: "Lecture 101"
author: "granolarr by Dr Stefano De Sabbata<br/>School of Geography, Geology, and the Env.<br/>University of Leicester<br/><a href=\"https://github.com/r-for-geographic-data-science/granolarr\">github.com/sdesabbata/r-for-geographic-data-science</a><br/><a href=\"mailto:s.desabbata@le.ac.uk\">s.desabbata&commat;le.ac.uk</a> &vert; <a href=\"https://twitter.com/maps4thought\">&commat;maps4thought</a><br/>text licensed under <a href=\"https://creativecommons.org/licenses/by-sa/4.0/\">CC BY-SA 4.0</a>, code licensed under <a href=\"https://www.gnu.org/licenses/gpl-3.0.html\">GNU GPL v3.0</a>"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    template: ../utils/IOSlides/RfGDS_Template.html
    css: ../utils/RMarkdown/rmarkdown_classes.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("RGDS_HOME"))
rm(list = ls())
```



# Introduction to R



## About this module

This module will provide you with the fundamental skills in

- basic programming in R
- data wrangling
- data analysis
- reproducibility

basis for

- *Geospatial Data Analysis*
- *Geospatial Databases and Information Retrieval*



## R programming language
    
One of the most widely used programming languages and an effective tool for *(geospatial)* data science

:::::: {.cols data-latex=""}

::: {.col style="width: 50%" data-latex="{0.5\textwidth}"}

- data wrangling
- statistical analysis
- machine learning
- data visualisation and maps
- processing spatial data
- geographic information analysis

:::

::: {.col style="width: 50%; text-align: right;" data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide', fig.height = 3, fig.width = 3}
ifelse(!dir.exists(file.path(Sys.getenv("RGDS_HOME"), "tmp")), dir.create(file.path(Sys.getenv("RGDS_HOME"), "tmp")), FALSE)

if (!dir.exists(file.path("./tmp", "E06000016"))) {
  unzip(paste0(Sys.getenv("RGDS_HOME"), "/data/", "e06000016.zip"), exdir = "tmp")
}

library(sp)
library(rgdal)
leic_2011OAC <- readOGR("tmp/E06000016/E06000016_Leicester.shp")

library(tmap)
tm_shape(leic_2011OAC) +
  tm_polygons(
    "supgrpname", 
    legend.show = FALSE,
    border.alpha = 0) +
  tm_layout(
    bg.color = "#FFF0E2", 
    frame = FALSE) +
  tm_credits(
    "Contains public sectorinformation licensed\nunder the Open Government Licence v3.0", 
    size = 0.5, 
    position = c("right", "bottom")
  )

unlink("tmp/E06000016", recursive = TRUE)
if(length(list.files(path = "./tmp", include.dirs = TRUE)) == 0){
  unlink("tmp", recursive = TRUE)
}
```

:::

::::::



## Schedule

The lectures and practical sessions have been designed to follow the schedule below



:::::: {.cols data-latex=""}

::: {.col data-latex="{0.5\textwidth}"}

- **1 R coding**
  - 100 Introduction
  - 110 R programming
- **2 Data wrangling**
  - 200 Selection and manipulation
  - 210 Table operations
  - 220 Reproducibility
  
:::

::: {.col data-latex="{0.5\textwidth}"}

- **3 Data analysis**
  - 300 Exploratory data analysis
  - 310 Comparing data
  - 320 Regression models
- **4 Machine learning**
  - 400 Unsupervised
  - 410 Supervised

:::
::::::



## Reference books

<div style="font-size:80%;">
Suggested reading

- *Programming Skills for Data Science: Start Writing Code to Wrangle, Analyze, and Visualize Data with R* by Michael Freeman and Joel Ross, Addison-Wesley, 2019. See book [webpage](https://www.pearson.com/us/higher-education/program/Freeman-Programming-Skills-for-Data-Science-Start-Writing-Code-to-Wrangle-Analyze-and-Visualize-Data-with-R/PGM2047488.html) and [repository](https://programming-for-data-science.github.io/).
- *R for Data Science* by Garrett Grolemund and Hadley Wickham, O'Reilly Media, 2016. See [online book](https://r4ds.had.co.nz/).
- *Discovering Statistics Using R* by Andy Field, Jeremy Miles and Zoë Field, SAGE Publications Ltd, 2012. See book [webpage](https://www.discoveringstatistics.com/books/discovering-statistics-using-r/).
- *Machine Learning with R: Expert techniques for predictive modeling* by Brett Lantz, Packt Publishing, 2019. See book [webpage](https://subscription.packtpub.com/book/big_data_and_business_intelligence/9781788295864).

Further reading

- *The Art of R Programming: A Tour of Statistical Software Design* by Norman Matloff, No Starch Press, 2011. See book [webpage](https://nostarch.com/artofr.htm)
- *An Introduction to R for Spatial Analysis and Mapping* by Chris Brunsdon and Lex Comber, Sage, 2015. See book [webpage](https://uk.sagepub.com/en-gb/eur/an-introduction-to-r-for-spatial-analysis-and-mapping/book241031)
- *Geocomputation with R* by Robin Lovelace, Jakub Nowosad, Jannes Muenchow, CRC Press, 2019. See [online book](https://bookdown.org/robinlovelace/geocompr/).
</div>


## R

Created in 1992 by Ross Ihaka and Robert Gentleman at the University of Auckland, New Zealand

- Free, open-source implementation of *S*
    - statistical programming language 
    - Bell Labs

<br/>

- Functional programming language
- Supports (and commonly used as) procedural (i.e., imperative) programming
- Object-oriented
- Interpreted (not compiled)



## Interpreting values

When values and operations are inputted in the *Console*, the interpreter returns the results of its interpretation of the expression

```{r, echo=TRUE}
2
"String value"
# comments are ignored
```



## Basic types

R provides three core data types

- numeric 
    - both integer and real numbers
- character 
    - i.e., text, also called *strings*
- logical
    - `TRUE` or `FALSE`


## Numeric operators

R provides a series of basic numeric operators

<div>

|Operator|Meaning         |Example  |Output     |
|--------|----------------|---------|-----------|
|+       |Plus            |`5 + 2`  |`r 5 + 2`  |
|-       |Minus           |`5 - 2`  |`r 5 - 2`  |
|`*`     |Product         |`5 * 2`  |`r 5 * 2`  |
|/       |Division        |`5 / 2`  |`r 5 / 2`  |
|%/%     |Integer division|`5 %/% 2`|`r 5 %/% 2`|
|%%      |Module          |`5 %% 2` |`r 5 %% 2` |
|^       |Power           |`5^2`    |`r 5^2`    |

</div>
<br/>

```{r, echo=TRUE}
5 + 2
```



## Logical operators


R provides a series of basic logical operators to test

<div>

|Operator|Meaning           |Example           |Output              |
|--------|------------------|------------------|--------------------|
|==      |Equal             |`5 == 2`          |`r 5 == 2`          |
|!=      |Not equal         |`5 != 2`          |`r 5 != 2`          |
|> (>=)  |Greater (or equal)|`5 > 2`           |`r 5 > 2`           |
|< (<=)  |Less (or equal)   |`5 <= 2`          |`r 5 <= 2`          |
|!       |Not               |`!TRUE`           |`r !TRUE`           |
|&       |And               |`TRUE & FALSE`    |`r TRUE & FALSE`    |
|\|      |Or                |`TRUE | FALSE`    |`r TRUE | FALSE`    |

</div>
<br/>

```{r, echo=TRUE}
5 >= 2
```



## Summary

An introduction to R

- Basic types
- Basic operators

**Next**: Core concepts

- Variables
- Functions
- Libraries

```{r cleanup, include=FALSE}
rm(list = ls())
```
