---
title: "Lecture 102"
author: "Dr Stefano De Sabbata<br/><small>School of Geography, Geology, and the Env., University of Leicester<br/><a href=\"https://github.com/sdesabbata/r-for-geographic-data-science\" style=\"color: white\">github.com/sdesabbata/r-for-geographic-data-science</a><br/><a href=\"mailto:s.desabbata@le.ac.uk\" style=\"color: white\">s.desabbata&commat;le.ac.uk</a> &vert; <a href=\"https://twitter.com/maps4thought\" style=\"color: white\">&commat;maps4thought</a><br/>text licensed under <a href=\"https://creativecommons.org/licenses/by-sa/4.0/\" style=\"color: white\">CC BY-SA 4.0</a>, code licensed under <a href=\"https://www.gnu.org/licenses/gpl-3.0.html\" style=\"color: white\">GNU GPL v3.0</a></small>"
institute: ""
date: "<small>2022-10-04</small>"
output:
  xaringan::moon_reader:
    self_contained: true
    css: ["css/sdesabbata-uol.css", "css/sdesabbata-uol-fonts.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "macros/my_remark_macros.js"
---
class: inverse, center, middle

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("RGDS_HOME"))
rm(list = ls())
```



# Reproducible data science



---

## Recap

**Prev**: Introduction

- 101 Lecture: Introduction to R
- 102 Lecture: Core concepts
- 103 Lecture: Tidyverse
- 104 Practical session

**Now**: Data types

- vectors
- factors
- matrices, arrays
- lists


---

## Data science

Data science is

---

## Reproduciblity

In quantitative research, an analysis or project are considered to be **reproducible** if:

- *"the data and code used to make a finding are available and they are sufficient for an independent researcher to recreate the finding."* [Christopher Gandrud, *Reproducible Research with R and R Studio*](https://www.crcpress.com/Reproducible-Research-with-R-and-R-Studio/Gandrud/p/book/9781498715379)

That is becoming more and more important in science:

- as programming and scripting are becoming integral in most disciplines
- as the amount of data increases



---

## Why?

In **scientific research**:

- verificability of claims through replication
- incremental work, avoid duplication

For your **working practice**:

- better working practices
    - coding
    - project structure
    - versioning
- better teamwork
- higher impact (not just results, but code, data, etc.)



---

## Reproducibility and software engineering

Core aspects of **software engineering** are:

- project design
- software **readibility**
- testing
- **versioning**

As programming becomes integral to research, similar necessities arise among scientists and data analysts.



---

## Reproducibility and "big data"

There has been a lot of discussions about **"big data"**...

- volume, velocity, variety, ...

Beyond the hype of the moment, as the **amount** and **complexity** of data increases

- the time required to replicate an analysis using point-and-click software becomes unsustainable
- room for error increases

Workflow management software (e.g., ArcGIS ModelBuilder) is one answer, reproducible data analysis based on script languages like R is another.



---

## Reproducibility in GIScience

[Singleton *et al.*](https://www.tandfonline.com/doi/abs/10.1080/13658816.2015.1137579) have discussed the issue of reproducibility in GIScience, identifying the following best practices:

1. Data should be accessible within the public domain and available to researchers.
2. Software used should have open code and be scrutable.
3. Workflows should be public and link data, software, methods of analysis and presentation with discursive narrative
4. The peer review process and academic publishing should require submission of a workflow model and ideally open archiving of those materials necessary for
replication.
5. Where full reproducibility is not possible (commercial software or sensitive data) aim to adopt aspects attainable within circumstances



---

## Document everything

In order to be reproducible, every step of your project should be documented in detail

- data gathering
- data analysis
- results presentation

Well documented R scripts are an excellent way to document your project. 



---

## Document well

Create code that can be **easily understood** by someone outside your project, including yourself in six-month time!

- use a style guide (e.g. [tidyverse](http://style.tidyverse.org/)) consistently
- also add a **comment** before any line that could be ambiguous or particularly difficult or important
- add a **comment** before each code block, describing what the code does
- add a **comment** at the beginning of a file, including
    - date
    - contributors
    - other files the current file depends on
    - materials, sources and other references 



---

## Workflow

Relationships between files in a project are not simple:

- in which order are file executed?
- when to copy files from one folder to another, and where?

A common solution is using **make files**

- commonly written in *bash* on Linux systems
- they can be written in R, using commands like
    - *source* to execute R scripts
    - *system* to interact with the operative system



---

## Mark.R

Section of the [*R for Geographic Data Science*](https://github.com/sdesabbata/r-for-geographic-data-science) project make file [Make.R](https://github.com/sdesabbata/r-for-geographic-data-science/blob/main/src/Make.R) that generates the current slides for the lecture session 221

```{}
cat("\n\n>>> Rendering 221-reproducibility.Rmd <<<\n\n")
rmarkdown::render(
  paste0(
    Sys.getenv("RGDS_HOME"), 
    "/src/slides/221-reproducibility.Rmd"
  ), 
  quiet = TRUE, 
  output_dir = paste0(
    Sys.getenv("RGDS_HOME"), 
    "/docs/slides"
  )
)
```


---

## Future-proof formats

Complex formats (e.g., .docx, .xlsx, .shp, ArcGIS .mxd)

- can become obsolete
- are not always portable
- usually require proprietary software

Use the simplest format to **future-proof** your analysis.<br/>**Text files** are the most versatile

- data: .txt, .csv, .tsv
- analysis: R scripts, python scripts
- write-up: LaTeX, Markdown, HTML



---

## Store and share

Reproducible data analysis is particularly important when working in teams, to share and communicate your work.

- [Dropbox](https://www.dropbox.com)
    - good option to work in teams, initially free
    - no versioning, branches
- [Git](https://git-scm.com)
    - free and opensource control system
    - great to work in teams and share your work publicly
    - can be more difficult at first
    - [GitHub](https://github.com) offers free public and private repositories
    - [GitLab](https://about.gitlab.com/) offers free public and private repositories



---

## Vectors

**Vectors** are ordered list of values.

- Vectors can be of any data type
    - numeric
    - character
    - logic
- All items in a vector have to be of the same type
- Vectors can be of any length


---

## Defining vectors

A vector variable can be defined using 

- an **identifier** (e.g., `a_vector`) 
- on the left of an **assignment operator** `<-`
- followed by the object to be linked to the identifier
- in this case, the result returned by the function `c`
- which creates a vector containing the provided elements

```{r, echo=TRUE}
a_vector <- c("Birmingham", "Derby", "Leicester",
  "Lincoln", "Nottingham", "Wolverhampton")
a_vector
```

<!--
Note that the second line of the answer provided by R in the console when asking to evaluate the variable starts with `[5]`, as the second line stats with the fifth element of the answer (i.e., the vector)
-->

---

## Creating vectors

- the operator `:`
- the function `seq`
- the function `rep`

```{r, echo=TRUE}
4:7
seq(1, 7, by = 0.5)
seq(1, 10, length.out = 7)
rep("Ciao", 4)
```

---

## Functions on vectors

Functions can be used on a vector variable directly

```{r, echo=TRUE}
a_numeric_vector <- 1:5
a_numeric_vector + 10
sqrt(a_numeric_vector)
a_numeric_vector >= 3
```



---

## Any and all

Overall expressions can be tested using the functions:

- **any**, TRUE if any of the elements satisfies the condition
- **all**, TRUE if all of the elements satisfy the condition

```{r, echo=TRUE}
any(a_numeric_vector >= 3)
all(a_numeric_vector >= 3)
```


---

## Selection

Each element of a vector can be retrieved specifying the related **index** between square brackets, after the identifier of the vector. The first element of the vector has index 1.

```{r, echo=TRUE}
a_vector[3]
```


A vector of indexes can be used to retrieve more than one element.

```{r, echo=TRUE}
a_vector[c(5, 3)]
```



---

## Factors

A **factor** is a data type similar to a vector. However, the values contained in a factor can only be selected from a set of **levels**.

```{r, echo=TRUE}
houses_vector <- c("Bungalow", "Flat", "Flat",
  "Detached", "Flat", "Terrace", "Terrace")
houses_vector
houses_factor <- factor(c("Bungalow", "Flat", "Flat",
  "Detached", "Flat", "Terrace", "Terrace"))
houses_factor
```

---

## table

The function **table** can be used to obtain a tabulated count for each level.

```{r, echo=TRUE}
houses_factor <- factor(c("Bungalow", "Flat", "Flat",
  "Detached", "Flat", "Terrace", "Terrace"))
houses_factor
table(houses_factor)
```

---

## Specified levels

A specific set of levels can be specified when creating a factor by providing a **levels** argument.

```{r, echo=TRUE}
houses_factor_spec <- factor(
  c("People Carrier", "Flat", "Flat", "Hatchback",
      "Flat", "Terrace", "Terrace"),
  levels = c("Bungalow", "Flat", "Detached",
       "Semi", "Terrace"))

table(houses_factor_spec)
```


---

## (Unordered) Factors

In statistics terminology, (unordered) factors are **categorical** (i.e., binary or nominal) variables. Levels are not ordered.

```{r, echo=TRUE}
income_nominal <- factor(
  c("High", "High", "Low", "Low", "Low",
      "Medium", "Low", "Medium"),
  levels = c("Low", "Medium", "High"))
```

The *greater than* operator is not meaningful on the `income_nominal` factor defined above

```{r, echo=TRUE}
income_nominal > "Low"
```

---

## Ordered Factors

In statistics terminology, ordered factors are **ordinal** variables. Levels are ordered.

```{r, echo=TRUE}
income_ordered <- ordered(
  c("High", "High", "Low", "Low", "Low",
      "Medium", "Low", "Medium"),
  levels = c("Low", "Medium", "High"))

income_ordered > "Low"

sort(income_ordered)
```


---

## Lists

Variables of the type **list** can contain elements of different types (including vectors and matrices), whereas elements of vectors are all of the same type. 

```{r, echo=TRUE}
employee <- list("Stef", 2015)
employee
employee[[1]] # Note the double square brackets for selection
```


---

## Named Lists

In **named lists** each element has a name, and elements can be selected using their name after the symbol `$`. 

```{r, echo=TRUE}
employee <- list(employee_name = "Stef", start_year = 2015)
employee
employee$employee_name
```


---

## Data Frames

A **data frame** is equivalent to a *named list* where all elements are *vectors of the same length*.

```{r, echo=TRUE}
employees <- data.frame(
  EmployeeName = c("Maria", "Pete", "Sarah"),
  Age = c(47, 34, 32),
  Role = c("Professor", "Researcher", "Researcher"))
employees
```

Data frames are the most common way to represent tabular data in R. Matrices and lists can be converted to data frames.


---

## tibble

A [tibble](https://tibble.tidyverse.org/) is a modern reimagining of the data.frame within `tidyverse`

- they do less 
    - don’t change column names or types
    - don’t do partial matching
- complain more
    - e.g. when referring to a column that does not exist
    
That forces you to confront problems earlier, typically leading to cleaner, more expressive code.




---

## dplyr

The `dplyr` (pronounced *dee-ply-er*) library is part of `tidyverse` and it offers a grammar for data manipulation

- `select`: select specific columns
- `filter`: select specific rows
- `arrange`: arrange rows in a particular order
- `summarise`: calculate aggregated values (e.g., mean, max, etc)
- `group_by`: group data based on common column values
- `mutate`: add columns
- `join`: merge tables (`tibbles` or `data.frames`)

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
```



---

## Example dataset

The library `nycflights13` contains a dataset storing data about all the flights departed from New York City in 2013
<!--
```{r, echo=TRUE, eval=FALSE}
install.packages("nycflights13")
```
-->
```{r, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
library(nycflights13)

nycflights13::flights
```
```{r, echo=FALSE}
library(nycflights13)

nycflights13::flights %>% 
  print(n= 3, width = 70)
```



---

## Selecting table columns

Columns of **data frames** and **tibbles** can be selected

- specifying the column index

```{r, echo=TRUE, eval=FALSE}
nycflights13::flights[, c(13, 14)]
```

- specifying the column name

```{r, echo=TRUE, eval=FALSE}
nycflights13::flights[, c("origin", "dest")]
```
```{r, echo=FALSE}
nycflights13::flights[, c("origin", "dest")] %>% 
  print(n = 3)
```



---

## dplyr::select

`select` can be used to specify which columns to retain

```{r, echo=TRUE, eval=FALSE}
nycflights13::flights %>%
  dplyr::select( 
    origin, dest, dep_delay, arr_delay, year:day
  )
```
```{r, echo=FALSE}
nycflights13::flights %>%
  dplyr::select( 
    origin, dest, dep_delay, arr_delay, year:day
  ) %>% 
  print(n = 5)
```


---

## dplyr::select

... or whichones to drop, using - in front of the column name

```{r, echo=TRUE, eval=FALSE}
nycflights13::flights %>%
  dplyr::select(origin, dest, dep_delay, arr_delay, year:day) %>% 
  dplyr::select(-arr_delay)
```
```{r, echo=FALSE}
nycflights13::flights %>%
  dplyr::select(
    origin, dest, dep_delay, arr_delay, year:day
  ) %>% 
  dplyr::select(
    -arr_delay
  ) %>% 
  print(n = 3)
```


---

## Logical filtering

Conditional statements can be used to filter a vector

- i.e. to retain only certain values
- where the specified value is `TRUE`

```{r, echo=TRUE}
a_numeric_vector <- -3:3
a_numeric_vector
a_numeric_vector[c(FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE)]
```



---

## Conditional filtering

As a conditional expression results in a logic vector...

```{r, echo=TRUE}
a_numeric_vector > 0
```

<br/>
... conditional expressions can be used for filtering

```{r, echo=TRUE}
a_numeric_vector[a_numeric_vector > 0]
```



---

## Filtering data frames

The same approach can be applied to **data frames** and **tibbles**

```{r, echo=TRUE, eval=FALSE}
nycflights13::flights$month
```
```{r, echo=FALSE}
capture.output(nycflights13::flights$month)[1] %>% str_trunc(52) %>% cat()
```

```{r, echo=TRUE, eval=FALSE}
nycflights13::flights$month == 11
```
```{r, echo=FALSE}
capture.output(nycflights13::flights$month == 11)[1] %>% str_trunc(52) %>% cat()
```

```{r, echo=TRUE, eval=FALSE}
nycflights13::flights[nycflights13::flights$month == 11, ]
```
```{r, echo=FALSE}
nycflights13::flights[nycflights13::flights$month == 11, ] %>% 
  print(n = 1, width = 52)
```



---

## dplyr::filter

```{r, echo=TRUE, eval=FALSE}
nycflights13::flights %>%
  # Flights in November
  dplyr::filter(month == 11)
```
```{r, echo=FALSE}
nycflights13::flights %>%
  dplyr::filter(month == 11) %>%
  print(n = 3, width = 52)
```




---

## Select and filter

```{r, echo=TRUE, eval=FALSE}
nycflights13::flights %>%
  # Select the columns you need
  dplyr::select(origin, dest, dep_delay, arr_delay, year:day) %>% 
  # Drop arr_delay... because you don't need it after all
  dplyr::select(-arr_delay) %>%
  # Filter in only November flights
  dplyr::filter(month == 11)
```
```{r, echo=FALSE}
nycflights13::flights %>%
  dplyr::select(origin, dest, dep_delay, arr_delay, year:day) %>%
  dplyr::select(-arr_delay) %>%
  dplyr::filter(month == 11) %>%
  print(n = 3, width = 52)
```



---

## Recap

Data types

- Vectors
- Factors
- Matrices, arrays
- Lists

**Next**: Control structures

- Conditional statements
- Loops


<br/>

.referencenote[
Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan). The chakra comes from [remark.js](https://remarkjs.com), [**knitr**](https://yihui.org/knitr), and [R Markdown](https://rmarkdown.rstudio.com).

```{r cleanup, include=FALSE}
rm(list = ls())
```
