---
title: "Lecture 102"
author: "Dr Stefano De Sabbata<br/><small>School of Geography, Geology, and the Env., University of Leicester<br/><a href=\"https://github.com/sdesabbata/r-for-geographic-data-science\" style=\"color: white\">github.com/sdesabbata/r-for-geographic-data-science</a><br/><a href=\"mailto:s.desabbata@le.ac.uk\" style=\"color: white\">s.desabbata&commat;le.ac.uk</a> &vert; <a href=\"https://twitter.com/maps4thought\" style=\"color: white\">&commat;maps4thought</a><br/>text licensed under <a href=\"https://creativecommons.org/licenses/by-sa/4.0/\" style=\"color: white\">CC BY-SA 4.0</a>, code licensed under <a href=\"https://www.gnu.org/licenses/gpl-3.0.html\" style=\"color: white\">GNU GPL v3.0</a></small>"
institute: ""
date: "<small>2022-10-04</small>"
output:
  xaringan::moon_reader:
    self_contained: true
    css: ["css/sdesabbata-uol.css", "css/sdesabbata-uol-fonts.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "macros/my_remark_macros.js"
---
class: inverse, center, middle

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("RGDS_HOME"))
rm(list = ls())
```



# Reproducible data science



---

## Recap

**Prev**: Introduction to R

- ...

**Today**: Reproducible data science

- ...


---

## Data science

.pull-left[

The term *"data science"* has gained wide-spread usage in the past decade

- akin **data mining**
- including or bordering
    - machine learning
    - artificial intelligence
    - statistical analysis
    - data visualisation
- commonly associated with
    - **big data**
    - the use of **programming**
        - python and R in particular
        - but not necessarily
    - **interdisciplinarity**
        - generally used for those who *use* those tools within an *application field* (e.g., geography or geology)

]
.pull-right[

```{r}
knitr::include_url(
  "https://sdesabbata.github.io/2021-census-analysis/100-exploratory-analysis/101-population-households.html",
  height = "500px")
```

.referencenote[
A simple example: [UK population and household change between 2011 and 2021 censuses](https://sdesabbata.github.io/2021-census-analysis/100-exploratory-analysis/101-population-households.html#Visualisation:_plots)
]

]




---
class: inverse, center, middle

# Complex data types




---

## Vectors

.pull-left[

**Vectors** are ordered list of values.

- Vectors can be of any data type
    - numeric
    - character
    - logic
- All items in a vector have to be of the same type
- Vectors can be of any length
- Functions can be used on a vector variable directly

]
.pull-right[

A vector variable can be defined using 

- an **identifier** (e.g., `a_vector`) 
- on the left of an **assignment operator** `<-`
- followed by the object to be linked to the identifier
- in this case, the result returned by the function `c`
- which creates a vector containing the provided elements

]

<br/>

```{r, echo=TRUE}
a_vector <- c("Birmingham", "Derby", "Leicester",
  "Lincoln", "Nottingham", "Wolverhampton")
a_vector
```




---

## Selection and filtering

Each element of a vector can be retrieved specifying the related **index** between square brackets, after the identifier of the vector. The first element of the vector has index 1.

```{r, echo=TRUE}
a_vector[3]
```


A vector of indexes can be used to retrieve more than one element.

```{r, echo=TRUE}
a_vector[c(5, 3)]
```

A condition can be provided for filtering only the elements for which the condition is true

```{r, echo=TRUE}
a_numeric_vector <- 1:5
a_numeric_vector[a_numeric_vector >= 3]
```

---

## Factors

A **factor** is a data type similar to a vector. However, the values contained in a factor can only be selected from a set of **levels**.

A specific set of levels can be specified when creating a factor by providing a **levels** argument.

```{r, echo=TRUE}
houses_factor_spec <- factor(
  c("People Carrier", "Flat", "Flat", "Hatchback",
      "Flat", "Terrace", "Terrace"),
  levels = c("Bungalow", "Flat", "Detached",
       "Semi", "Terrace"))

table(houses_factor_spec)
```

Factors are unordered by default, but ordered factors can be created


---

## Lists and data frames

Variables of the type **list** can contain elements of different types (including vectors and matrices), whereas elements of vectors are all of the same type. 

```{r, echo=TRUE}
employee <- list("Stef", 2015)
employee
employee[[1]] # Note the double square brackets for selection
```

In **named lists** each element has a name, and elements can be selected using the operator `$`. 

```{r, echo=TRUE}
employee <- list(employee_name = "Stef", start_year = 2015)
employee$employee_name
```

---

## Data frames

A **data frame** is equivalent to a *named list* where all elements are *vectors of the same length*.

```{r, echo=TRUE}
employees <- data.frame(
  employee_name = c("Maria", "Pete", "Sarah"),
  age = c(47, 34, 32),
  role = c("Professor", "Researcher", "Researcher"))

employees
```

Data frames are the most common way to represent tabular data in R.

```{r, echo=FALSE, eval=TRUE}
employees %>% 
  knitr::kable()
```

---

## Selecting and filtering

.pull-left[

**Columns** of data frames can be **selected**

- specifying the column index

```{r, echo=TRUE, eval=FALSE}
employees[, c(3, 2)]
```

- specifying the column name

```{r, echo=TRUE}
employees[, c("role", "employee_name")]
```

The same applies to **rows** (**slicing**)

```{r, echo=TRUE}
employees[3, ]
```

]
.pull-right[

As with vectors, a condition can be provided for **filtering** only the **rows** for which the condition is true

```{r, echo=TRUE}
employees$role
```

```{r, echo=TRUE}
employees$role == "Researcher"
```

```{r, echo=TRUE}
employees[employees$role == "Researcher", ]
```

]



---
class: inverse, center, middle

# Reproduciblity


---

## Reproduciblity

.pull-left[
In quantitative research, an analysis or project are considered to be **reproducible** if:

- *"the data and code used to make a finding are available and they are sufficient for an independent researcher to recreate the finding."* [Christopher Gandrud, *Reproducible Research with R and R Studio*](https://www.crcpress.com/Reproducible-Research-with-R-and-R-Studio/Gandrud/p/book/9781498715379)

That is becoming more and more important in science:

- as programming and scripting are becoming integral in most disciplines
- as the amount of data increases

]
.pull-right[

In **scientific research**:

- verificability of claims through replication
- incremental work, avoid duplication

For your **working practice**:

- better working practices
    - coding
    - project structure
    - versioning
- better teamwork
- higher impact (not just results, but code, data, etc.)

]

---

## Reproducibility in GIScience

[Singleton *et al.*](https://www.tandfonline.com/doi/abs/10.1080/13658816.2015.1137579) have discussed the issue of reproducibility in GIScience, identifying the following best practices:

1. Data should be accessible within the public domain and available to researchers.
2. Software used should have open code and be scrutable.
3. Workflows should be public and link data, software, methods of analysis and presentation with discursive narrative
4. The peer review process and academic publishing should require submission of a workflow model and ideally open archiving of those materials necessary for
replication.
5. Where full reproducibility is not possible (commercial software or sensitive data) aim to adopt aspects attainable within circumstances


---

## Reproducibility and software engineering

.pull-left[

Emergence of **"big data"**

- volume, velocity, variety, ...

Beyond the hype of the moment, as the **amount** and **complexity** of data increases

- the time required to replicate an analysis using point-and-click software becomes unsustainable
- room for error increases

Two options:

- Workflow management software (e.g., ArcGIS ModelBuilder) 
- Reproducible data analysis (based on script languages like R)


]
.pull-right[

Core aspects of **software engineering** are:

- project design
- software **readibility**
- testing
- **versioning**

As programming becomes integral to research, similar necessities arise among scientists and data analysts.

5 key aspects identified by [Gandrud](https://www.crcpress.com/Reproducible-Research-with-R-and-R-Studio/Gandrud/p/book/9781498715379)

- Document everything
- Document well
- Workflow
- Future-proof formats
- Store and share

]


---

.pull-left[

## Document everything

In order to be reproducible, every step of your project should be documented in detail

- data gathering
- data analysis
- results presentation

Well documented R scripts are an excellent way to document your project. 

]
.pull-right[

## Document well

Create code that can be **easily understood** by someone outside your project, including yourself in six-month time!

- use a style guide (e.g. [tidyverse](http://style.tidyverse.org/)) consistently
- also add a **comment** before any line that could be ambiguous or particularly difficult or important
- add a **comment** before each code block, describing what the code does
- add a **comment** at the beginning of a file, including
    - date
    - contributors
    - other files the current file depends on
    - materials, sources and other references 

]


---

## Workflow

Relationships between files in a project are not simple:

- in which order are file executed?
- when to copy files from one folder to another, and where?

A common solution is using **make files**

- commonly written in *bash* on Linux systems
- they can be written in R, using commands like
    - *source* to execute R scripts
    - *system* to interact with the operative system

Example: section of the [*R for Geographic Data Science*](https://github.com/sdesabbata/r-for-geographic-data-science) project's make file [Make.R](https://github.com/sdesabbata/r-for-geographic-data-science/blob/main/src/Make.R)

```{}
cat("\n\n>>> Rendering 102-slides-reproducible-data-science.Rmd <<<\n\n")
rmarkdown::render(
  paste0(Sys.getenv("RGDS_HOME"), "/src/slides/102-slides-reproducible-data-science.Rmd"), 
  quiet = TRUE, 
  output_dir = paste0(Sys.getenv("RGDS_HOME"), "/docs/slides")
)

```



---

.pull-left[

## Future-proof formats

Complex formats (e.g., .docx, .xlsx, .shp, ArcGIS .mxd)

- can become obsolete
- are not always portable
- usually require proprietary software

Use the simplest format to **future-proof** your analysis.<br/>**Text files** are the most versatile

- data: .txt, .csv, .tsv
- analysis: R scripts, python scripts
- write-up: LaTeX, Markdown, HTML


]
.pull-right[

## Store and share

Reproducible data analysis is particularly important when working in teams, to share and communicate your work.

- [Dropbox](https://www.dropbox.com)
    - good option to work in teams, initially free
    - no versioning, branches
- [Git](https://git-scm.com)
    - free and opensource control system
    - great to work in teams and share your work publicly
    - can be more difficult at first
    - [GitHub](https://github.com) offers free public and private repositories
    - [GitLab](https://about.gitlab.com/) offers free public and private repositories

]




---
class: inverse, center, middle

# Into the Tidyverse

---

.pull-left[

## dplyr

The `dplyr` (pronounced *dee-ply-er*) library is part of `tidyverse` and it offers a grammar for data manipulation

- `select`: select specific columns
- `filter`: select specific rows
- `arrange`: arrange rows in a particular order
- `summarise`: calculate aggregated values (e.g., mean, max, etc)
- `group_by`: group data based on common column values
- `mutate`: add columns
- `join`: merge tables (`data.frames` or `tibbles`)

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
```

]
.pull-right[

## tibble

A [tibble](https://tibble.tidyverse.org/) is a modern reimagining of the data.frame within `tidyverse`

- they do less 
    - don’t change column names or types
    - don’t do partial matching
- complain more
    - e.g. when referring to a column that does not exist
    
That forces you to confront problems earlier, typically leading to cleaner, more expressive code.

]

---

## Example dataset

The library `nycflights13` contains a dataset storing data about all the flights departed from New York City in 2013

```{r, echo=TRUE, eval=FALSE}
install.packages("nycflights13")
```

```{r, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
library(nycflights13)

flights
```
```{r, echo=FALSE}
library(nycflights13)

flights %>% 
  print(n= 5, width = 100)
```





---

## dplyr::select

.pull-left[

`select` can be used to specify which columns to retain

```{r, echo=TRUE, eval=FALSE}
flights %>%
  select( 
    origin, dest, 
    dep_delay, arr_delay, 
    year:day
  )

```


]
.pull-right[

... or whichones to drop, using - in front of the column name

```{r, echo=TRUE, eval=FALSE}
flights %>%
  select(
    origin, dest, 
    dep_delay, arr_delay, 
    year:day
  ) %>% 
  select(-arr_delay)
```

]

.pull-left[

```{r, echo=FALSE}
flights %>%
  select( 
    origin, dest, dep_delay, arr_delay, year:day
  ) %>% 
  print(n = 7)
```

]
.pull-right[

```{r, echo=FALSE}
flights %>%
  select(
    origin, dest, dep_delay, arr_delay, year:day
  ) %>% 
  select(
    -arr_delay
  ) %>% 
  print(n = 7)
```
]

---

## dplyr::filter

.pull-left[

`filter` can be used to specify a condition for **filtering** only the **rows** for which the condition is true

```{r, echo=TRUE, eval=FALSE}
flights %>%
  # Flights in November
  filter(month == 11)
```

```{r, echo=FALSE}
flights %>%
  filter(month == 11) %>%
  print(n = 3, width = 52)
```

]
.pull-right[

... and can be concatenated with `select` or any other function using `%>%` to create a more complex process

```{r, echo=TRUE, eval=FALSE}
flights %>%
  # Select the columns you need
  select(
    origin, dest, 
    dep_delay, arr_delay, 
    year:day
  ) %>% 
  # Drop arr_delay... because you don't need it after all
  select(-arr_delay) %>%
  # Filter in only November flights
  filter(month == 11)
```

```{r, echo=FALSE}
flights %>%
  select(origin, dest, dep_delay, arr_delay, year:day) %>%
  select(-arr_delay) %>%
  filter(month == 11) %>%
  print(n = 3, width = 52)
```

]


---

## Recap

**Today**: Reproducible data science

- ...

**Next**: ...


<br/>

.referencenote[
Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan). The chakra comes from [remark.js](https://remarkjs.com), [**knitr**](https://yihui.org/knitr), and [R Markdown](https://rmarkdown.rstudio.com).

```{r cleanup, include=FALSE}
rm(list = ls())
```
