---
title: "Lecture 102"
author: "granolarr by Dr Stefano De Sabbata<br/>School of Geography, Geology, and the Env.<br/>University of Leicester<br/><a href=\"https://github.com/sdesabbata/granolarr\">github.com/sdesabbata/granolarr</a><br/><a href=\"mailto:s.desabbata@le.ac.uk\">s.desabbata&commat;le.ac.uk</a> &vert; <a href=\"https://twitter.com/maps4thought\">&commat;maps4thought</a><br/>licensed under <a href=\"https://www.gnu.org/licenses/gpl-3.0.html\">GNU GPL v3.0</a>"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    template: ../utils/IOSlides/UoL_Template.html
    css: ../utils/RMarkdown/rmarkdown_classes.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("RGDS_HOME"))
rm(list = ls())
```



# Core concepts



## Recap

**Prev**: An introduction to R

- Basic types
- Basic operators

**Now**: Core concepts

- Variables
- Functions
- Libraries



## Variables

Variables **store data** and can be defined 

- using an *identifier* (e.g., `a_variable`) 
- on the left of an *assignment operator* `<-`
- followed by the object to be linked to the identifier
- such as a *value* (e.g., `1`) 

```{r, echo=TRUE}
a_variable <- 1
```

The value of the variable can be invoked by simply specifying the **identifier**.

```{r, echo=TRUE}
a_variable
```



## Algorithms and functions

*An* **algorithm** *or effective procedure is a mechanical rule, or automatic method, or programme for performing some mathematical operation* (Cutland, 1980).

A **program** is a specific set of instructions that implement an abstract algorithm.

The definition of an algorithm (and thus a program) can consist of one or more **function**s

- set of instructions that preform a task 
- possibly using an input, possibly returning an output value

Programming languages usually provide pre-defined functions that implement common algorithms (e.g., to find the square root of a number or to calculate a linear regression)



## Functions

Functions execute complex operations and can be invoked 

- specifying the *function name*
- the *arguments* (input values) between simple brackets
    - each *argument* corresponds to a *parameter*
    - sometimes the *parameter* name must be specified

```{r, echo=TRUE}
sqrt(2)
round(1.414214, digits = 2)
```



## Functions and variables

- functions can be used on the right side of `<-` 
- variables and functions can be used as *arguments*

```{r, echo=TRUE}
sqrt_of_two <- sqrt(2)
sqrt_of_two
round(sqrt_of_two, digits = 2)
round(sqrt(2), digits = 2)
```



## Naming

When creating an identifier for a variable or function

- R is a **case sensitive** language
    - UPPER and lower case are not the same
    - `a_variable` is different from `a_VARIABLE`
- names can include
    -  alphanumeric symbols
    - `.` and `_`
- names must start with
    - a letter



## Libraries

Once a number of related, reusable functions are created

- they can be collected and stored in **libraries** (a.k.a. *packages*)
  - `install.packages` is a function that can be used to install libraries (i.e., downloads it on your computer)
  - `library` is a function that *loads* a library (i.e., makes it available to a script)

Libraries can be of any size and complexity, e.g.:

- `base`: base R functions, including the `sqrt` function above
- `rgdal`: implementation of the [GDAL (Geospatial Data Abstraction Library)](https://gdal.org/) functionalities


## stringr

R provides some basic functions to manipulate strings, but the `stringr` library provides a more consistent and well-defined set

```{r, echo=TRUE}
library(stringr)

str_length("Leicester")
str_detect("Leicester", "e")
str_replace_all("Leicester", "e", "x")
```




## Summary

Core concepts

- Variables
- Functions
- Libraries

**Next**: Tidyverse

- Tidyverse libraries
- *pipe* operator

```{r cleanup, include=FALSE}
rm(list = ls())
```
