---
title: "Lecture 111"
author: "granolarr by Dr Stefano De Sabbata<br/>School of Geography, Geology, and the Env.<br/>University of Leicester<br/><a href=\"https://github.com/sdesabbata/granolarr\">github.com/sdesabbata/granolarr</a><br/><a href=\"mailto:s.desabbata@le.ac.uk\">s.desabbata&commat;le.ac.uk</a> &vert; <a href=\"https://twitter.com/maps4thought\">&commat;maps4thought</a><br/>licensed under <a href=\"https://www.gnu.org/licenses/gpl-3.0.html\">GNU GPL v3.0</a>"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    template: ../utils/IOSlides/UoL_Template.html
    logo: ../../src/utils/IOSlides/uol_logo.png
    css: ../utils/RMarkdown/rmarkdown_classes.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("RGDS_HOME"))
rm(list = ls())
```



# Data types



## Recap

**Prev**: Introduction

- 101 Lecture: Introduction to R
- 102 Lecture: Core concepts
- 103 Lecture: Tidyverse
- 104 Practical session

**Now**: Data types

- vectors
- factors
- matrices, arrays
- lists



## Vectors
**Vectors** are ordered list of values.

- Vectors can be of any data type
    - numeric
    - character
    - logic
- All items in a vector have to be of the same type
- Vectors can be of any length


## Defining vectors

A vector variable can be defined using 

- an **identifier** (e.g., `a_vector`) 
- on the left of an **assignment operator** `<-`
- followed by the object to be linked to the identifier
- in this case, the result returned by the function `c`
- which creates a vector containing the provided elements

```{r, echo=TRUE}
a_vector <- c("Birmingham", "Derby", "Leicester",
  "Lincoln", "Nottingham", "Wolverhampton")
a_vector
```

<!--
Note that the second line of the answer provided by R in the console when asking to evaluate the variable starts with `[5]`, as the second line stats with the fifth element of the answer (i.e., the vector)
-->

## Creating vectors

- the operator `:`
- the function `seq`
- the function `rep`

```{r, echo=TRUE}
4:7
seq(1, 7, by = 0.5)
seq(1, 10, length.out = 7)
rep("Ciao", 4)
```


## Selection

Each element of a vector can be retrieved specifying the related **index** between square brackets, after the identifier of the vector. The first element of the vector has index 1.

```{r, echo=TRUE}
a_vector[3]
```


A vector of indexes can be used to retrieve more than one element.

```{r, echo=TRUE}
a_vector[c(5, 3)]
```


<!--
## Examples

```{r, echo=TRUE}
east_midlands_cities <- c("Derby", "Leicester", "Lincoln", "Nottingham")
my_indexes <- 2:4
east_midlands_cities[my_indexes]
selected_cities <- c(east_midlands_cities[1], east_midlands_cities[3:4])
```
-->


## Functions on vectors

Functions can be used on a vector variable directly

```{r, echo=TRUE}
a_numeric_vector <- 1:5
a_numeric_vector + 10
sqrt(a_numeric_vector)
a_numeric_vector >= 3
```



## Any and all

Overall expressions can be tested using the functions:

- **any**, TRUE if any of the elements satisfies the condition
- **all**, TRUE if all of the elements satisfy the condition

```{r, echo=TRUE}
any(a_numeric_vector >= 3)
all(a_numeric_vector >= 3)
```



## Factors

A **factor** is a data type similar to a vector. However, the values contained in a factor can only be selected from a set of **levels**.

```{r, echo=TRUE}
houses_vector <- c("Bungalow", "Flat", "Flat",
  "Detached", "Flat", "Terrace", "Terrace")
houses_vector
houses_factor <- factor(c("Bungalow", "Flat", "Flat",
  "Detached", "Flat", "Terrace", "Terrace"))
houses_factor
```

## table

The function **table** can be used to obtain a tabulated count for each level.

```{r, echo=TRUE}
houses_factor <- factor(c("Bungalow", "Flat", "Flat",
  "Detached", "Flat", "Terrace", "Terrace"))
houses_factor
table(houses_factor)
```

## Specified levels

A specific set of levels can be specified when creating a factor by providing a **levels** argument.

```{r, echo=TRUE}
houses_factor_spec <- factor(
  c("People Carrier", "Flat", "Flat", "Hatchback",
      "Flat", "Terrace", "Terrace"),
  levels = c("Bungalow", "Flat", "Detached",
       "Semi", "Terrace"))

table(houses_factor_spec)
```


## (Unordered) Factors

In statistics terminology, (unordered) factors are **categorical** (i.e., binary or nominal) variables. Levels are not ordered.

```{r, echo=TRUE}
income_nominal <- factor(
  c("High", "High", "Low", "Low", "Low",
      "Medium", "Low", "Medium"),
  levels = c("Low", "Medium", "High"))
```

The *greater than* operator is not meaningful on the `income_nominal` factor defined above

```{r, echo=TRUE}
income_nominal > "Low"
```

## Ordered Factors

In statistics terminology, ordered factors are **ordinal** variables. Levels are ordered.

```{r, echo=TRUE}
income_ordered <- ordered(
  c("High", "High", "Low", "Low", "Low",
      "Medium", "Low", "Medium"),
  levels = c("Low", "Medium", "High"))

income_ordered > "Low"

sort(income_ordered)
```



## Matrices

**Matrices** are collections of numerics arranged in a two-dimensional rectangular layout

- the first argument is a vector of values
- the second specifies number of rows and columns
- R offers operators and functions for matrix algebra

```{r, echo=TRUE}
a_matrix <- matrix(c(3, 5, 7, 4, 3, 1), c(3, 2))
a_matrix
```

<!--
## Matrices

Variables of the type **matrix** are collections of data elements arranged in a two-dimensional rectangular layout. The first argument is vector a containing the values, the second is a vector specifying the number of rows and columns.

```{r, echo=TRUE, eval=FALSE}
x <- matrix(c(3, 5, 7, 4, 3, 1), c(3, 2))
x
y <- matrix(c(3, 5, 7, 4, 3, 1), c(2, 3))
y
z <- matrix(c(3, 5, 7, 4, 3, 1), c(2, 3), byrow = TRUE)
z
```


## Matrix algebra

Operations as in mathematical matrix algebra.

```{r, echo=TRUE, eval=FALSE}
x
y
z
```

```{r, echo=TRUE, eval=FALSE}
y + z
y %*% x
t(x)
```



## Matrix algebra

Operations as in mathematical matrix algebra.

```{r, echo=TRUE, eval=FALSE}
A <- matrix(
      c(3, 1, 2, 1),
      nrow = 2, ncol = 2)
A
b <- matrix(
      c(8, 2),
      nrow = 2, ncol = 1)
b
```

```{r, echo=TRUE, eval=FALSE}
# inverse
solve(A)
# returns x, where b = Ax
solve(A, b)
# That is
# 8 = 3*4 + 2*-2
# 2 = 1*4 + 1*-2
```
-->

## Arrays

:::::: {.cols data-latex=""}

::: {.col data-latex="{0.5\textwidth}"}

Variables of the type **array**<br/>are higher-dimensional<br/>matrices.

- the first argument is a<br/>vector containing the<br/>values
- the second argument is<br/>avector specifying the<br/>depth of each dimension

```{r, echo=TRUE}
a3dim_array <- array(1:24, dim=c(4, 3, 2))
```

```{r, echo=TRUE, eval=FALSE}
a3dim_array
```
:::

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=FALSE}
a3dim_array
```

:::
::::::



<!--
## Matrices and arrays: selection

The subset can be specified as a matrix of indexes.

```{r, echo=TRUE, eval=FALSE}
x2 
choose_x2 <- matrix(
              c(3, 3, 3, 1, 2, 1, 2, 3),
              c(4, 2), byrow = TRUE)
choose_x2
x2[choose_x2]
```
-->


## Selection

Subsets of matrices (and arrays) can be selected as seen for vectors.

```{r, echo=TRUE}
a_matrix[2, c(1, 2)]
a3dim_array[c(1, 2), 2, 2]
```



<!--
## apply

`apply` applies another function to each level of a set dimension of an array

```{r, echo=TRUE}
apply(a3dim_array, 3, min) # apply on third dimension
apply(a3dim_array, 1, min) # apply on first dimension
apply(a3dim_array, 2, min) # apply on second dimension
```
-->



## Lists

Variables of the type **list** can contain elements of different types (including vectors and matrices), whereas elements of vectors are all of the same type. 

```{r, echo=TRUE}
employee <- list("Stef", 2015)
employee
employee[[1]] # Note the double square brackets for selection
```


## Named Lists

In **named lists** each element has a name, and elements can be selected using their name after the symbol `$`. 

```{r, echo=TRUE}
employee <- list(employee_name = "Stef", start_year = 2015)
employee
employee$employee_name
```


<!--
## lapply

With `lapply` take care that the function makes sense for *any* element in the list

```{r, echo=TRUE}
various <- list(
  "Some text",
  matrix(c(6, 3, 1, 2), c(2, 2))
)
lapply(various, is.numeric)
```



## Lists: append

The function **append** can be used to append a new element at the end of a list.

```{r, echo=TRUE, eval=FALSE}
various <- list("Stefano", c(2015, 2016), "Lecturer",
                matrix(c(6, 3, 1, 2), c(2, 2)))
various <- append(various, list(c(7, 6, 9, 1))) 
various
```
-->



## Recap

Data types

- Vectors
- Factors
- Matrices, arrays
- Lists

**Next**: Control structures

- Conditional statements
- Loops

```{r cleanup, include=FALSE}
rm(list = ls())
```
