---
title: "Lecture 112"
author: "granolarr by Dr Stefano De Sabbata<br/>School of Geography, Geology, and the Env.<br/>University of Leicester<br/><a href=\"https://github.com/sdesabbata/granolarr\">github.com/sdesabbata/granolarr</a><br/><a href=\"mailto:s.desabbata@le.ac.uk\">s.desabbata&commat;le.ac.uk</a> &vert; <a href=\"https://twitter.com/maps4thought\">&commat;maps4thought</a><br/>licensed under <a href=\"https://www.gnu.org/licenses/gpl-3.0.html\">GNU GPL v3.0</a>"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    template: ../utils/IOSlides/UoL_Template.html
    css: ../utils/RMarkdown/rmarkdown_classes.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("RGDS_HOME"))
rm(list = ls())
```



# Control structures



## Recap

**Prev**: Data types

- Vectors
- Factors
- Matrices and arrays
- Lists

**Now**: Control structures

- Conditional statements
- Loops



## If

Format: **if** (*condition*) *statement*

- *condition*: expression returning a logic value (`TRUE` or `FALSE`)
- *statement*: any valid R statement
- *statement* only executed if *condition* is `TRUE`


```{r, echo=TRUE}
a_value <- -7
if (a_value < 0) cat("Negative")
a_value <- 8
if (a_value < 0) cat("Negative")
```


## Else
Format: **if** (*condition*) *statement1* **else** *statement2*

- *condition*: expression returning a logic value (`TRUE` or `FALSE`)
- *statement1* and *statement2*: any valid R statements
- *statement1*  executed if *condition* is `TRUE`
- *statement2*  executed if *condition* is `FALSE`


```{r, echo=TRUE}
a_value <- -7
if (a_value < 0) cat("Negative") else cat("Positive")
a_value <- 8
if (a_value < 0) cat("Negative") else cat("Positive")
```

<!--
## Example

```{r, echo=TRUE}
x <- 10
if (is.numeric(x) & (!(x < 0) & (x != 0)) ) cat("greater than zero")
```
-->


## Code blocks

**Code blocks** allow to encapsulate **several** statements in a single group

- `{` and `}` contain code blocks
- the statements are execute together

```{r, echo=TRUE}
first_value <- 8
second_value <- 5
if (first_value > second_value) {
  cat("First is greater than second\n") 
  difference <- first_value - second_value
  cat("Their difference is ", difference)
}
```



## Loops
Loops are a fundamental component of (procedural) programming.


There are two main types of loops:

- **conditional** loops are executed as long as a defined condition holds true
    - construct `while`
    - construct `repeat`
- **deterministic** loops are executed a pre-determined number of times
    - construct `for`


## While

The *while* construct can be defined using the `while` reserved word, followed by the conditional statement between simple brackets, and a code block. The instructions in the code block are re-executed as long as the result of the evaluation of the conditional statement is `TRUE`.

```{r, echo=TRUE}
current_value <- 0

while (current_value < 3) {
  cat("Current value is", current_value, "\n")
  current_value <- current_value + 1
}
```

<!--
## Repeat

The *repeat* construct can be defined using the `repeat` reserved word, followed by a code block. The instructions in the code block are re-executed until the command `break` is given. The latter is currently given through an `if` construct, which tests the condition that would stop the loop.

```{r, echo=TRUE}
current_value <- 0

repeat {
  cat("Current value is", current_value, "\n")
  current_value <- current_value + 1
  if (current_value >= 3) break
}
```


## While vs Repeat
The difference between `while` and `repeat` is mostly syntactical.

- Sometimes one or the other might fit better with the algorithm you have in mind
- Use the one that comes easier to you in the given situation
-->

## For

The *for* construct can be defined using the `for` reserved word, followed by the definition of an **iterator**. The iterator is a variable which is temporarily assigned with the current element of a vector, as the construct iterates through all elements of the vector. This definition is followed by a code block, whose instructions are re-executed once for each element of the vector.

```{r, echo=TRUE}
cities <- c("Derby", "Leicester", "Lincoln", "Nottingham")
for (city in cities) {
  cat("Do you live in", city, "?\n")
}
```


## For

It is common practice to create a vector of integers on the spot in order to execute a certain sequence of steps a pre-defined number of times.

```{r, echo=TRUE}
for (i in 1:3) {
  cat("This is exectuion number", i, ":\n")
  cat("    See you later!\n")
}
```


## Loops with conditional statements

```{r, echo=TRUE}
3:0
#Example: countdown!
for (i in 3:0) {
  if (i == 0) {
    cat("Go!\n")
  } else {
    cat(i, "\n")
  }
}
```



## Summary

Control structures

- Conditional statements
- Loops

**Next**: Functions

- Defining functions
- Scope of a variable

```{r cleanup, include=FALSE}
rm(list = ls())
```
