---
title: "Lecture 201"
author: "granolarr by Dr Stefano De Sabbata<br/>School of Geography, Geology, and the Env.<br/>University of Leicester<br/><a href=\"https://github.com/sdesabbata/granolarr\">github.com/sdesabbata/granolarr</a><br/><a href=\"mailto:s.desabbata@le.ac.uk\">s.desabbata&commat;le.ac.uk</a> &vert; <a href=\"https://twitter.com/maps4thought\">&commat;maps4thought</a><br/>licensed under <a href=\"https://www.gnu.org/licenses/gpl-3.0.html\">GNU GPL v3.0</a>"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    template: ../utils/IOSlides/UoL_Template.html
    logo: ../../src/utils/IOSlides/uol_logo.png
    css: ../utils/RMarkdown/rmarkdown_classes.css
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("RGDS_HOME"))
rm(list = ls())
```



# Data Frames



## Recap

**Prev**: R programming

- 111 Lecture: Data types
- 112 Lecture: Control structures
- 113 Lecture: Functions
- 114 Practical session

**Now**: Data Frames

- Data Frames
- Tibbles


## Lists and named lists

**List**

- can contain elements of different types
  - whereas elements of vectors are all of the same type
- in **named lists**, each element has a name
  - elements can be selected using the operator `$`

```{r, echo=TRUE}
employee <- list(employee_name = "Stef", start_year = 2015)
employee[[1]]
employee$employee_name
```



## Data Frames

A **data frame** is equivalent to a *named list* where all elements are *vectors of the same length*.

```{r, echo=TRUE}
employees <- data.frame(
  EmployeeName = c("Maria", "Pete", "Sarah"),
  Age = c(47, 34, 32),
  Role = c("Professor", "Researcher", "Researcher"))
employees
```

Data frames are the most common way to represent tabular data in R. Matrices and lists can be converted to data frames.

<!--
**Note:** As is common in R, the elements of an eventual too short column are recycled to match column lengths.
-->

## Selection

Selection is similar to vectors and lists.

```{r, echo=TRUE}
employees[1, 1] # value selection
employees[1, ] # row selection
employees[, 1] # column selection
```


## Selection

Selection is similar to vectors and lists.

```{r, echo=TRUE}
employees$EmployeeName # column selection, as for named lists
employees$EmployeeName[1]
```



## Table manipulation

- Values can be assigned to cells
  - using any selection method 
  - and the assignment operator `<-`
- New columns can be defined 
  - assigning a vector to a new name

```{r, echo=TRUE}
employees$Age[3] <- 33 
employees$Place <- c("Leicester", "Leicester","Leicester") 
employees
```



## Column processing

Operations can be performed on columns as they where vectors

```{r, echo=TRUE}
10 - c(1, 2, 3)
```

```{r, echo=TRUE}
# Use Sys.Date to retrieve the current year
current_year <- as.integer(format(Sys.Date(), "%Y"))

# Calculate employee year of birth
employees$Year_of_birth <- current_year - employees$Age
employees
```



## tibble

A [tibble](https://tibble.tidyverse.org/) is a modern reimagining of the data.frame within `tidyverse`

- they do less 
    - don’t change column names or types
    - don’t do partial matching
- complain more
    - e.g. when referring to a column that does not exist
    
That forces you to confront problems earlier, typically leading to cleaner, more expressive code.



## Summary

Data Frames

- Data Frames
- Tibbles

**Next**: Data selection and filtering

- dplyr
- dplyr::select
- dplyr::filter

```{r cleanup, include=FALSE}
rm(list = ls())
```
