---
title: "Lecture 202"
author: "granolarr by Dr Stefano De Sabbata<br/>School of Geography, Geology, and the Env.<br/>University of Leicester<br/><a href=\"https://github.com/sdesabbata/granolarr\">github.com/sdesabbata/granolarr</a><br/><a href=\"mailto:s.desabbata@le.ac.uk\">s.desabbata&commat;le.ac.uk</a> &vert; <a href=\"https://twitter.com/maps4thought\">&commat;maps4thought</a><br/>licensed under <a href=\"https://www.gnu.org/licenses/gpl-3.0.html\">GNU GPL v3.0</a>"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    template: ../utils/IOSlides/UoL_Template.html
    css: ../utils/RMarkdown/rmarkdown_classes.css
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("RGDS_HOME"))
rm(list = ls())
```



# Selection and filtering



## Recap

**Prev**: Data Frames

- Data Frames
- Tibbles

**Now**: Data selection and filtering

- dplyr
- dplyr::select
- dplyr::filter



## dplyr

The `dplyr` (pronounced *dee-ply-er*) library is part of `tidyverse` and it offers a grammar for data manipulation

- `select`: select specific columns
- `filter`: select specific rows
- `arrange`: arrange rows in a particular order
- `summarise`: calculate aggregated values (e.g., mean, max, etc)
- `group_by`: group data based on common column values
- `mutate`: add columns
- `join`: merge tables (`tibbles` or `data.frames`)

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
```



## Example dataset

The library `nycflights13` contains a dataset storing data about all the flights departed from New York City in 2013
<!--
```{r, echo=TRUE, eval=FALSE}
install.packages("nycflights13")
```
-->
```{r, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
library(nycflights13)

nycflights13::flights
```
```{r, echo=FALSE}
library(nycflights13)

nycflights13::flights %>% 
  print(n= 3, width = 70)
```



## Selecting table columns

Columns of **data frames** and **tibbles** can be selected

- specifying the column index

```{r, echo=TRUE, eval=FALSE}
nycflights13::flights[, c(13, 14)]
```

- specifying the column name

```{r, echo=TRUE, eval=FALSE}
nycflights13::flights[, c("origin", "dest")]
```
```{r, echo=FALSE}
nycflights13::flights[, c("origin", "dest")] %>% 
  print(n = 3)
```



## dplyr::select

`select` can be used to specify which columns to retain

```{r, echo=TRUE, eval=FALSE}
nycflights13::flights %>%
  dplyr::select( 
    origin, dest, dep_delay, arr_delay, year:day
  )
```
```{r, echo=FALSE}
nycflights13::flights %>%
  dplyr::select( 
    origin, dest, dep_delay, arr_delay, year:day
  ) %>% 
  print(n = 5)
```


## dplyr::select

... or whichones to drop, using - in front of the column name

```{r, echo=TRUE, eval=FALSE}
nycflights13::flights %>%
  dplyr::select(origin, dest, dep_delay, arr_delay, year:day) %>% 
  dplyr::select(-arr_delay)
```
```{r, echo=FALSE}
nycflights13::flights %>%
  dplyr::select(
    origin, dest, dep_delay, arr_delay, year:day
  ) %>% 
  dplyr::select(
    -arr_delay
  ) %>% 
  print(n = 3)
```


## Logical filtering

Conditional statements can be used to filter a vector

- i.e. to retain only certain values
- where the specified value is `TRUE`

```{r, echo=TRUE}
a_numeric_vector <- -3:3
a_numeric_vector
a_numeric_vector[c(FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE)]
```



## Conditional filtering

As a conditional expression results in a logic vector...

```{r, echo=TRUE}
a_numeric_vector > 0
```

<br/>
... conditional expressions can be used for filtering

```{r, echo=TRUE}
a_numeric_vector[a_numeric_vector > 0]
```



## Filtering data frames

The same approach can be applied to **data frames** and **tibbles**

```{r, echo=TRUE, eval=FALSE}
nycflights13::flights$month
```
```{r, echo=FALSE}
capture.output(nycflights13::flights$month)[1] %>% str_trunc(52) %>% cat()
```

```{r, echo=TRUE, eval=FALSE}
nycflights13::flights$month == 11
```
```{r, echo=FALSE}
capture.output(nycflights13::flights$month == 11)[1] %>% str_trunc(52) %>% cat()
```

```{r, echo=TRUE, eval=FALSE}
nycflights13::flights[nycflights13::flights$month == 11, ]
```
```{r, echo=FALSE}
nycflights13::flights[nycflights13::flights$month == 11, ] %>% 
  print(n = 1, width = 52)
```



## dplyr::filter

```{r, echo=TRUE, eval=FALSE}
nycflights13::flights %>%
  # Flights in November
  dplyr::filter(month == 11)
```
```{r, echo=FALSE}
nycflights13::flights %>%
  dplyr::filter(month == 11) %>%
  print(n = 3, width = 52)
```




## Select and filter

```{r, echo=TRUE, eval=FALSE}
nycflights13::flights %>%
  # Select the columns you need
  dplyr::select(origin, dest, dep_delay, arr_delay, year:day) %>% 
  # Drop arr_delay... because you don't need it after all
  dplyr::select(-arr_delay) %>%
  # Filter in only November flights
  dplyr::filter(month == 11)
```
```{r, echo=FALSE}
nycflights13::flights %>%
  dplyr::select(origin, dest, dep_delay, arr_delay, year:day) %>%
  dplyr::select(-arr_delay) %>%
  dplyr::filter(month == 11) %>%
  print(n = 3, width = 52)
```



## Summary

Data selection and filtering

- dplyr
- dplyr::select
- dplyr::filter

**Next**: Data manipulation

- dplyr::arrange
- dplyr::summarise
- dplyr::group_by
- dplyr::mutate



```{r cleanup, include=FALSE}
rm(list = ls())
```
