---
title: "Lecture 203"
author: "Dr Stefano De Sabbata<br/><small>School of Geography, Geology, and the Env., University of Leicester<br/><a href=\"https://github.com/sdesabbata/r-for-geographic-data-science\" style=\"color: white\">github.com/sdesabbata/r-for-geographic-data-science</a><br/><a href=\"mailto:s.desabbata@le.ac.uk\" style=\"color: white\">s.desabbata&commat;le.ac.uk</a> &vert; <a href=\"https://twitter.com/maps4thought\" style=\"color: white\">&commat;maps4thought</a><br/>text licensed under <a href=\"https://creativecommons.org/licenses/by-sa/4.0/\" style=\"color: white\">CC BY-SA 4.0</a>, code licensed under <a href=\"https://www.gnu.org/licenses/gpl-3.0.html\" style=\"color: white\">GNU GPL v3.0</a></small>"
institute: ""
date: "<small>2022-11-23</small>"
output:
  xaringan::moon_reader:
    self_contained: true
    css: ["css/sdesabbata-uol.css", "css/sdesabbata-uol-fonts.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "macros/my_remark_macros.js"
---
class: inverse, center, middle

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("RGDS_HOME"))
rm(list = ls())
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(magrittr)  

library(knitr)
library(kableExtra)

library(nycflights13)
library(palmerpenguins)

library(pastecs)

leicester_2011OAC <- 
  read_csv(paste0(Sys.getenv("RGDS_HOME"), "/data/", "2011_OAC_Raw_uVariables_Leicester.csv"))
```

# Comparing data


---
## Recap

.pull-left[

<br/>

**Prev**: Exploratory statistics

- Descriptive statistics
- Exploring assumptions
    - Normality
    - Skewness and kurtosis
    - Homogeneity of variance

**Today**: Comparing data

- Comparing distributions through mean
- Correlation analysis
- Variable transformation

<br/>

]
.pull-right[

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7, fig.height = 7}
library(GGally)

penguins %>%
  select(
    bill_length_mm,
    bill_depth_mm,
    flipper_length_mm, 
    body_mass_g,
    species
  ) %>%
  ggpairs(
    mapping = aes(color = species),
    upper = 
      list(continuous = 
        wrap(ggally_cor, method = "kendall")),
    lower = 
      list(continuous = 
        wrap("points", alpha = 0.3, size=0.1))
  ) +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
  theme_bw()
```


]




---
## Comparing means

.pull-left[


When are (two) group means  different?

Independent T-test:

- null hypothesis
  - there is **no difference** between the groups
- if *p-value* (significance) below threshold (e.g., `0.05` or `0.01`)
  - **group means are different**
- assumptions
    - normally distributed values in groups
    - homogeneity of variance of values in groups
      - if groups have different sizes
    - independence of groups
      - e.g. different conditions of an experiment

]
.pull-right[


```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 6, fig.height = 6}
penguins %>%
  filter(species %in% c("Adelie", "Gentoo")) %>%
  ggplot(
    aes(
      x = species, 
      y = bill_depth_mm,
      fill = species
    )
  ) +
  geom_boxplot() +
  scale_fill_manual(values = c("darkorange", "cyan4")) +
  theme_bw()
```

]



---
## General linear model

<br>

**General linear model**

- observation $i$ can be predicted by a $model$ (predictors)
- accounting for some amount of error

$$outcome_i = (model) + error_i $$

<br>

**Independent T-test** as a general linear model

- groups is the predictor (categorical variable)
- single observation value as group mean plus error

$$outcome_i = (group\ mean) + error_i $$


---
## Example: bill depth

.pull-left[

<br/>

Is the bill depth of *Adelie* and *Gentoo* penguins different?

1. Check assumptions
    1. Indipendent groups: ok
    2. normal distribution: check using Shapiro-Wilk test
    3. homogeneity of variance: not necessary
2. Run T-test
    1. `stats::t.test`

]
.pull-right[

Values are normally distributed for both groups

```{r, echo=TRUE, message=FALSE, warning=FALSE}
penguins %>% filter(species == "Adelie") %>% 
  pull(bill_depth_mm) %>% shapiro.test()

penguins %>% filter(species == "Gentoo") %>% 
  pull(bill_depth_mm) %>% shapiro.test()
```

]


---
## stats::t.test

.pull-left[

The test is significant, the group means are different

```{r, echo=TRUE, message=FALSE, warning=FALSE}
penguins %>%
  filter(species %in% c("Adelie", "Gentoo")) %$%
  t.test(bill_depth_mm ~ species)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
penguins_t_test <- penguins %>%
  filter(species %in% c("Adelie", "Gentoo")) %$%
  t.test(bill_depth_mm ~ species)
```

How to report: 

- *t*(`r penguins_t_test %$% parameter[["df"]] %>% round(digits = 2)`) = `r penguins_t_test %$% statistic[["t"]] %>% round(digits = 2)`, *p* < .01

]
.pull-right[


```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 6, fig.height = 6}
penguins %>%
  filter(species %in% c("Adelie", "Gentoo")) %>%
  ggplot(
    aes(
      x = species, 
      y = bill_depth_mm,
      fill = species
    )
  ) +
  geom_boxplot() +
  scale_fill_manual(values = c("darkorange", "cyan4")) +
  theme_bw()
```

]



---
## Example: bill depth (2)

.pull-left[

Is the bill depth of *Adelie* and *Chinstrap* penguins different?

We already checked normality for *Adelie* and *Gentoo*. Are values for *Chinstrap* normally distributed?

```{r, echo=TRUE, message=FALSE, warning=FALSE}
penguins %>% filter(species == "Chinstrap") %>% 
  pull(bill_depth_mm) %>% shapiro.test()
```

Values are normally distributed for all three groups


]
.pull-right[


```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 6, fig.height = 6}
penguins %>%
  filter(species %in% c("Adelie", "Chinstrap")) %>%
  ggplot(
    aes(
      x = species, 
      y = bill_depth_mm,
      fill = species
    )
  ) +
  geom_boxplot() +
  scale_fill_manual(values = c("darkorange", "darkorchid")) +
  theme_bw()
```

]


---
## stats::t.test

.pull-left[

The test is not significant, the group means are not different

```{r, echo=TRUE, message=FALSE, warning=FALSE}
penguins %>%
  filter(species %in% c("Adelie", "Chinstrap")) %$%
  t.test(bill_depth_mm ~ species)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
penguins_t_test <- penguins %>%
  filter(species %in% c("Adelie", "Chinstrap")) %$%
  t.test(bill_depth_mm ~ species)
```

How to report: 

- *t*(`r penguins_t_test %$% parameter[["df"]] %>% round(digits = 2)`) = `r penguins_t_test %$% statistic[["t"]] %>% round(digits = 2)`, *p* = `r penguins_t_test %$% p.value %>% round(digits = 2)`

]
.pull-right[


```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 6, fig.height = 6}
penguins %>%
  filter(species %in% c("Adelie", "Chinstrap")) %>%
  ggplot(
    aes(
      x = species, 
      y = bill_depth_mm,
      fill = species
    )
  ) +
  geom_boxplot() +
  scale_fill_manual(values = c("darkorange", "darkorchid")) +
  theme_bw()
```

]



---
## ANalysis Of VAriance (ANOVA)

.pull-left[

ANOVA is similar to the T-tests, but more than two groups

- null hypothesis
  - **no difference** between the groups
- if *p-value* (significance) below threshold (e.g., `0.05` or `0.01`)
  - **group means are different**
- assumptions
    - normally distributed values in groups
        - especially if groups have different sizes
    - homogeneity of variance of values in groups
      - if groups have different sizes
    - independence of groups
      - e.g. different conditions of an experiment
      
]
.pull-right[

**General linear model**

- observation $i$ can be predicted by a $model$ (predictors)
- accounting for some amount of error

$$outcome_i = (model) + error_i $$

<br/>

**ANOVA** as a general linear model

- groups is the predictor (categorical variable)
- single observation value as group mean plus error

$$outcome_i = (group\ mean) + error_i $$

]

---
## Example: bill depth (3)

.pull-left[

<br/>

Is the bill depth different between all three species?

1. Check assumptions
    1. Indipendent groups: ok
    2. normal distribution: check using Shapiro-Wilk test
    3. homogeneity of variance: not necessary
2. Run ANOVA
    1. `stats::aov`

]
.pull-right[

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 6, fig.height = 6}
penguins %>%
  ggplot(
    aes(
      x = species, 
      y = bill_depth_mm,
      fill = species, 
    )
  ) +
  geom_boxplot() +
  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
  theme_bw()
```

]

---
## stats::aov

<br/>

.pull-left[

The test is significant, the group means are different

```{r, echo=TRUE, message=FALSE, warning=FALSE}
penguins %$%
  aov(bill_depth_mm ~ species) %>%
  summary()
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
penguins_anova_summary <- penguins %$%
  aov(bill_depth_mm ~ species) %>%
  summary()
```

How to report:

- *F*(`r penguins_anova_summary[[1]] %$% Df[1] %>% round(digits = 2)`, `r penguins_anova_summary[[1]] %$% Df[2] %>% round(digits = 2)`) = `r penguins_anova_summary[[1]][1, 4] %>% round(digits = 2)`, *p* < .01

]
.pull-right[

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 6, fig.height = 6}
penguins %>%
  ggplot(
    aes(
      x = species, 
      y = bill_depth_mm,
      fill = species, 
    )
  ) +
  geom_boxplot() +
  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
  theme_bw()
```

]


---
class: inverse, center, middle

# Correlation


---
## Correlation

<br/>

**Correlation** is a standardised measure of covariance

.pull-left[

Two continuous variables can be

- not related at all
- related
    - positively: 
      - entities with *high values* in one 
      - tend to have *high values* in the other
    - negatively: 
      - entities with *high values* in one 
      - tend to have *low values* in the other

]
.pull-right[

Three different approaches

- **Pearson's r**
  - if two variables are **normally distributed**
- **Spearman’s rho**
  - if two variables are **not normally distributed**
- **Kendall’s tau**
  - if **not normally distributed** 
  - and there are a **large number of ties**

]

---
## Example

Are flipper length and body mass related in Chinstrap penguins?

.center[

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 10, fig.height = 6}
penguins %>%
  filter(species == "Chinstrap") %>%
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species,
                 shape = species),
             size = 2) +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  theme_bw()
```

]



---
## Pearson’s r

.pull-left[

<br/>

If  two variables are **normally distributed**, use **Pearson's r**

- null hypothesis
  - there is no relationship between the variables
- assumptions
  - variables are normally distributed

The square of the correlation value indicates the percentage of shared variance

]
.pull-right[

Flipper length and body mass are normally distributed in Chinstrap penguins

```{r, echo=TRUE, message=FALSE, warning=FALSE}
penguins %>%
  filter(species == "Chinstrap") %>%
  pull(flipper_length_mm) %>% shapiro.test()

penguins %>%
  filter(species == "Chinstrap") %>%
  pull(body_mass_g) %>% shapiro.test()
```

]

---
## stats::cor.test

.pull-left[

<br/>

If  two variables are **normally distributed**, use **Pearson's r**

- null hypothesis
  - there is no relationship between the variables
- assumptions
  - variables are normally distributed

The square of the correlation value indicates the percentage of shared variance

]
.pull-right[

```{r, echo=TRUE, message=FALSE, warning=FALSE}
penguins %>%
  filter(species == "Chinstrap") %$%
  cor.test(flipper_length_mm, body_mass_g)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
penguins %>%
  filter(species == "Chinstrap") %$%
  cor.test(flipper_length_mm, body_mass_g) ->
  chinstrap_flip_mass_cor
```

- Flipper length and body mass are related
  - `p-value < 0.01`
- sharing `r ((chinstrap_flip_mass_cor$estimate ^ 2) * 100) %>% round(digits = 1)`% of variance
  - `r chinstrap_flip_mass_cor$estimate %>% round(digits = 3)` `^ 2 =` `r chinstrap_flip_mass_cor$estimate ^ 2 %>% round(digits = 3)`

]

---
## Example: flipper length and body mass

.pull-left[

But, are flipper length and body mass related in penguins (without considering species as separated groups)?

.center[

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 6, fig.height = 4}
penguins %>%
  ggplot(aes(
    x = flipper_length_mm, 
    y = body_mass_g
  )) +
  geom_point(
    aes(
      color = species,
      shape = species),
    size = 2
  ) +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  theme_bw()
```

]

]
.pull-right[

Flipper length and body mass are not normally distributed when all penguins are taken into account as a single group

```{r, echo=TRUE, message=FALSE, warning=FALSE}
penguins %>%
  pull(flipper_length_mm) %>% shapiro.test()

penguins %>%
  pull(body_mass_g) %>% shapiro.test()
```

]

---
## Spearman’s rho

.pull-left[

If  two variables are **not normally distrib.**, use **Spearman’s rho**

- null hypothesis
  - there is no relationship between the variables
- non-parametric
  - uses full dataset to calculate the statistics
  - rather than estimate key parameters of distributions from data
- based on rank difference
- assumptions
  - ties are uncommon

The square of the correlation value indicates the percentage of shared variance

]
.pull-right[

```{r, echo=TRUE, message=FALSE, warning=FALSE}
penguins %$%
  cor.test(
    flipper_length_mm, body_mass_g, 
    method = "spearman"
  )
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
penguins %$%
  cor.test(flipper_length_mm, body_mass_g, method = "spearman") ->
  allpenguins_flip_mass_cor
```

- Flipper length and body mass are related
  - `p-value < 0.01`
- sharing `r ((allpenguins_flip_mass_cor$estimate ^ 2) * 100) %>% round(digits = 1)`% of variance
  - `r allpenguins_flip_mass_cor$estimate %>% round(digits = 3)` `^ 2 =` `r allpenguins_flip_mass_cor$estimate ^ 2 %>% round(digits = 3)`

]

---
## Correlation with ties

Spearman's rho

- `Cannot compute exact p-value with ties`

```{r, echo=TRUE, eval=FALSE}
penguins %$%
  cor.test(flipper_length_mm, body_mass_g, method = "spearman")
```

```{r, echo=TRUE, eval=FALSE}
## Warning in cor.test.default(flipper_length_mm, body_mass_g, method = "spearman"):
##  Cannot compute exact p-value with ties
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
penguins %$%
  cor.test(flipper_length_mm, body_mass_g, method = "spearman")
```


---
## Kendall’s tau

.pull-left[

If  two variables are **not normally distributed** and there are **many ties**, use **Kendall’s tau**

- null hypothesis
  - there is no relationship between the variables
- non-parametric
- based on rank difference
- no assumptions
- less *powerful*
  - even if there is a relationship, significance might be high

The square of the correlation value indicates the percentage of shared variance

]
.pull-right[

```{r, echo=TRUE, message=FALSE, warning=FALSE}
penguins %$%
  cor.test(
    flipper_length_mm, body_mass_g, 
    method = "kendall"
  )
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
penguins %$%
  cor.test(flipper_length_mm, body_mass_g, method = "kendall")->
  allpenguins_flip_mass_cor_kendall
```

- Flipper length and body mass are related
  - `p-value < 0.01`
- sharing `r ((allpenguins_flip_mass_cor_kendall$estimate ^ 2) * 100) %>% round(digits = 1)`% of variance
  - `r allpenguins_flip_mass_cor_kendall$estimate %>% round(digits = 3)` `^ 2 =` `r allpenguins_flip_mass_cor_kendall$estimate ^ 2 %>% round(digits = 3)`


]


---
## psych::pairs.panels

.pull-left[

Combining: 

- histograms
- scatter plots 
- correlations

```{r, eval=FALSE, echo=TRUE, message=FALSE, warning=FALSE}
library(psych)

penguins %>%
  select(
    bill_length_mm,
    bill_depth_mm,
    flipper_length_mm, 
    body_mass_g
  ) %>%
  pairs.panels(
    method = "kendall",
    stars = TRUE
  )

## Signif.: 0 '***' 0.001 '**' 0.01
## 0.01 '*' 0.05 0.05 '.' 0.1 ' ' 1
```

]
.pull-right[

.center[

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7, fig.height = 7}
library(psych)

penguins %>%
  select(
    bill_length_mm,
    bill_depth_mm,
    flipper_length_mm, 
    body_mass_g
  ) %>%
  pairs.panels(
    method = "kendall",
    stars = TRUE
  )
```

]

]

---
## GGally::ggpairs

More flexible and sophisticated pairs plots

.pull-left[

```{r, eval=FALSE, echo=TRUE, message=FALSE, warning=FALSE}
library(GGally)

penguins %>%
  select(
    bill_length_mm, bill_depth_mm,
    flipper_length_mm, body_mass_g, species
  ) %>%
  ggpairs(
    mapping = aes(color = species),
    upper = list(continuous = 
        wrap(ggally_cor, method = "kendall")),
    lower = list(continuous = 
        wrap("points", alpha = 0.3, size=0.1))
  ) +
  scale_color_manual(values = 
    c("darkorange","darkorchid","cyan4")) +
  scale_fill_manual(values = 
    c("darkorange","darkorchid","cyan4")) +
  theme_bw()

## Signif.: 0 '***' 0.001 '**' 0.01
## 0.01 '*' 0.05 0.05 '.' 0.1 ' ' 1
```

]
.pull-right[

.center[

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7, fig.height = 7}
library(GGally)

penguins %>%
  select(
    bill_length_mm, bill_depth_mm,
    flipper_length_mm, body_mass_g,
    species
  ) %>%
  ggpairs(
    mapping = aes(color = species),
    upper = 
      list(continuous = 
        wrap(ggally_cor, method = "kendall")),
    lower = 
      list(continuous = 
        wrap("points", alpha = 0.3, size=0.1))
  ) +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
  theme_bw()
```

]

]



---
## Chi-squre

.pull-left[

How to test a relationship between two **categorical** variables?

Chi-square test:

- null hypothesis
  - no relationship between variables
- non-parametric
- based on cross-tabulated expected counts
- no assumptions

```{r, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
library(gmodels)
penguins %$%
  CrossTable(
    island, species, chisq = TRUE, 
    expected = TRUE, prop.c = FALSE, 
    prop.t = FALSE, prop.chisq = FALSE, 
    sresid = TRUE, format = "SPSS"
  )
```

]
.pull-right[

.small[

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(gmodels)

penguins %$%
  CrossTable(island, species, chisq = TRUE, expected = TRUE, prop.c = FALSE, prop.t = FALSE, prop.chisq = FALSE, sresid = TRUE, format = "SPSS")
```

]

]

???

Different islands do have different amounts of penguins from different species



---
class: inverse, center, middle

# Data transformation


---
## Z-scores

.pull-left[

Transform the values as relative to

- distribution's mean $\mu$
- standard deviation $\sigma$ 

$$ z_i = \frac{x_i - \mu}{\sigma} $$


]
.pull-right[

```{r, echo=TRUE, eval=FALSE,message=FALSE, warning=FALSE, fig.width = 10, fig.height = 6}
penguins %>%
  mutate(
    flipper_length_zscore = 
      scale(flipper_length_mm)
  ) %>%
  ggplot(aes(x = flipper_length_zscore)) +
  geom_histogram() + 
  theme_bw()
```

]

.pull-left[

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 10, fig.height = 6}
penguins %>%
  ggplot(aes(x = flipper_length_mm)) +
  geom_histogram() + 
  theme_bw()
```

]
.pull-right[

```{r, echo=FALSE, eval=TRUE,message=FALSE, warning=FALSE, fig.width = 10, fig.height = 6}
penguins %>%
  mutate(
    flipper_length_zscore = 
      scale(flipper_length_mm)
  ) %>%
  ggplot(aes(x = flipper_length_zscore)) +
  geom_histogram() +
  theme_bw()
```

]

???

Same distribution (bar some aggregation in the histogram)

Commonly used to render two variables easier to compare


---
## Common transformations

.pull-left[

**Logarithmic** transformations are useful to *"un-skew"* variables

Common approaches include:

- natural logarithm (`log`)
- binary logarithm (`log2`)
- logarithm base 10 (`log10`)

Only possible on values `> 0`

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 2, fig.width = 6, fig.align = 'center'}
ggplot() + 
  stat_function(
    fun = log10
  ) +
  xlim(1, 1000) +
  xlab("Original values") +
  ylab("Logarithm base 10") +
  theme_bw()
```

]
.pull-right[

**Inverse hyperbolic sine** transformations are useful to *"un-skew"* variables

- similar to logarithmic transformations
- defined on all values
- in R: `asinh`

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 2, fig.width = 6, fig.align = 'center'}
ggplot() + 
  stat_function(
    fun = asinh
  ) +
  xlim(-500, 500) +
  xlab("Original values") +
  ylab("Inverse hyperbolic sine") +
  theme_bw()
```

]

.center[
Not always sufficient to *"un-skew"* variables
]

---
## Example: residents aged 20 to 24

.pull-left[

The number of residents aged 20 to 24 (`u011`) in the areas of Leicester described as *"Cosmopolitans"* by the [2011 Output Area Classification](https://www.gov.uk/government/statistics/2011-area-classification-for-super-output-areas) is skewed

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 4, fig.width = 6}
leicester_2011OAC %>% 
  filter(supgrpname == "Cosmopolitans") %>%
  ggplot(aes(x = u011)) +
  geom_histogram() + theme_bw()
```

]
.pull-right[

```{r, echo=TRUE, message=FALSE, warning=FALSE}
leicester_2011OAC %>% 
  filter(supgrpname == "Cosmopolitans") %>% 
  select(u011) %>% 
  stat.desc(
    basic = FALSE, desc = FALSE, 
    norm = TRUE
  ) %>%
  kable("html", digits = 3)
```

]



---
## Example: log10

.pull-left[

However, it's logarithm base 10 is normally distributed, thus it can be used with tests requiring normally distributed values

```{r, echo=TRUE, eval=FALSE}
mutate(log10_u011 = log10(u011))
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 4, fig.width = 6}
leicester_2011OAC %>% 
  filter(supgrpname == "Cosmopolitans") %>%
  mutate(log10_u011 = log10(u011)) %>%
  ggplot(aes(x = log10_u011)) +
  geom_histogram() + theme_bw()
```

]
.pull-right[

```{r, echo=TRUE, message=FALSE, warning=FALSE}
leicester_2011OAC %>% 
  filter(supgrpname == "Cosmopolitans") %>%
  mutate(log10_u011 = log10(u011)) %>% 
  select(log10_u011) %>% 
  stat.desc(
    basic = FALSE, desc = FALSE, 
    norm = TRUE
  ) %>%
  kable("html", digits = 3)
```

]



---
## Example: asinh

.pull-left[

The Inverse hyperbolic sine is also normally distributed

```{r, echo=TRUE, eval=FALSE}
mutate(ihs_u011 = asinh(u011))
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 4, fig.width = 6}
leicester_2011OAC %>% 
  filter(supgrpname == "Cosmopolitans") %>%
  mutate(ihs_u011 = asinh(u011)) %>%
  ggplot(aes(x = ihs_u011)) +
  geom_histogram() + theme_bw()
```


]
.pull-right[

```{r, echo=TRUE, message=FALSE, warning=FALSE}
leicester_2011OAC %>% 
  filter(supgrpname == "Cosmopolitans") %>%
  mutate(ihs_u011 = asinh(u011)) %>%
  select(ihs_u011) %>% 
  stat.desc(
    basic = FALSE, desc = FALSE, 
    norm = TRUE
  ) %>%
  kable("html", digits = 3)
```

]


---
## Summary

.pull-left[

<br/>

**Today**: Comparing data

- Comparing distributions through mean
- Correlation analysis
- Variable transformation

**Next time**: Regression analysis

- Simple regression
- Multiple regression
- Comparing models

<br/>

.referencenote[
Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan). The chakra comes from [remark.js](https://remarkjs.com), [**knitr**](https://yihui.org/knitr), and [R Markdown](https://rmarkdown.rstudio.com).
]

]
.pull-right[

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 7, fig.height = 7}
library(GGally)

penguins %>%
  select(
    bill_length_mm,
    bill_depth_mm,
    flipper_length_mm, 
    body_mass_g,
    species
  ) %>%
  ggpairs(
    mapping = aes(color = species),
    upper = 
      list(continuous = 
        wrap(ggally_cor, method = "kendall")),
    lower = 
      list(continuous = 
        wrap("points", alpha = 0.3, size=0.1))
  ) +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  scale_fill_manual(values = c("darkorange","darkorchid","cyan4")) +
  theme_bw()
```


]


```{r cleanup, include=FALSE}
rm(list = ls())
```
