---
title: "Lecture 203"
author: "granolarr by Dr Stefano De Sabbata<br/>School of Geography, Geology, and the Env.<br/>University of Leicester<br/><a href=\"https://github.com/sdesabbata/granolarr\">github.com/sdesabbata/granolarr</a><br/><a href=\"mailto:s.desabbata@le.ac.uk\">s.desabbata&commat;le.ac.uk</a> &vert; <a href=\"https://twitter.com/maps4thought\">&commat;maps4thought</a><br/>licensed under <a href=\"https://www.gnu.org/licenses/gpl-3.0.html\">GNU GPL v3.0</a>"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    template: ../utils/IOSlides/UoL_Template.html
    logo: ../../src/utils/IOSlides/uol_logo.png
    css: ../utils/RMarkdown/rmarkdown_classes.css
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("RGDS_HOME"))
rm(list = ls())
```



# Data manipulation



## Recap

**Prev**: Data selection and filtering

- dplyr
- dplyr::select
- dplyr::filter

**Now**: Data manipulation

- dplyr::arrange
- dplyr::summarise
- dplyr::group_by
- dplyr::mutate


## Example

```{r, echo=TRUE, eval=FALSE}
library(tidyverse)
library(nycflights13)

nov_dep_delays <- 
  nycflights13::flights %>%
  dplyr::select(origin, dest, dep_delay, year:day) %>%
  dplyr::filter(month == 11)

nov_dep_delays
```
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(nycflights13)

nov_dep_delays <- 
  nycflights13::flights %>%
  dplyr::select(origin, dest, dep_delay, year:day) %>%
  dplyr::filter(month == 11)

nov_dep_delays %>% print(n = 3)
```



## dplyr::arrange

Arranges rows in a particular order

- descending orders specified by using `-` (minus symbol)

```{r, echo=TRUE, eval=FALSE}
nov_dep_delays %>%
  dplyr::arrange(
    # Ascending destination name
    dest,
    # Descending delay
    -dep_delay
  )
```
```{r, echo=FALSE, message=FALSE, warning=FALSE}
nov_dep_delays %>%
  dplyr::arrange( 
    dest, # Ascending destination name
    -dep_delay # Descending delay
  ) %>% 
  print(n = 2)
```



## dplyr::summarise

Calculates aggregated values

- e.g., using functions such as mean, max, etc.

```{r, echo=TRUE}
nov_dep_delays %>%
  # Need to filter out rows where delay is NA
  dplyr::filter(!is.na(dep_delay)) %>%
  # Create two aggregated columns
  dplyr::summarise(
    avg_dep_delay = mean(dep_delay), 
    tot_dep_delay = sum(dep_delay)
  )
```



## dplyr::group_by

Groups rows based on common values for specified column(s)

- combined with `summarise`, aggregated values per group

```{r, echo=TRUE, eval=FALSE}
nov_dep_delays %>%
  # First group by same destination
  dplyr::group_by(dest) %>%
  # Then calculate aggregated value
  dplyr::filter(!is.na(dep_delay)) %>%
  dplyr::summarise(tot_dep_delay = sum(dep_delay))
```
```{r, echo=FALSE, message=FALSE, warning=FALSE}
nov_dep_delays %>%
  # First group by same destination
  dplyr::group_by(dest) %>%
  # Need to filter out rows where delay is NA
  dplyr::filter(!is.na(dep_delay)) %>%
  # Then calculate aggregated value
  dplyr::summarise(
    tot_dep_delay = sum(dep_delay)
  ) %>% 
  print(n = 2)
```



## dplyr::tally and dplyr::count

- `dplyr::tally` short-hand for `summarise` with `n`
  - number of rows
- `dplyr::count`short-hand for `group_by` and `tally`
  - number of rows per group


```{r, echo=TRUE, eval=FALSE}
nov_dep_delays %>%
  # Count flights by same destination
  dplyr::count(dest)
```
```{r, echo=FALSE, message=FALSE, warning=FALSE}
nov_dep_delays %>%
  # Count flights by same destination
  dplyr::count(dest) %>% 
  print(n = 3)
```


## dplyr::mutate

Calculate values for new columns based on current columns

```{r, echo=TRUE, eval=FALSE}
nov_dep_delays %>%
  dplyr::mutate(
    # Combine origin and destination into one column
    orig_dest = str_c(origin, dest, sep = "->"),
    # Departure delay in days (rather than minutes)
    delay_days = ((dep_delay / 60) /24)
  )
```
```{r, echo=FALSE, message=FALSE, warning=FALSE}
nov_dep_delays %>%
  dplyr::mutate(
    orig_dest = str_c(origin, dest, sep = "->"),
    delay_days = ((dep_delay / 60) /24)
  ) %>% 
  print(n = 3)
```


## Full pipe example

```{r, echo=TRUE, eval=FALSE}
nycflights13::flights %>%
  dplyr::select(
    origin, dest, dep_delay, arr_delay, 
    year:day
  ) %>%
  dplyr::select(-arr_delay) %>%
  dplyr::filter(month == 11) %>%
  dplyr::filter(!is.na(dep_delay)) %>%
  dplyr::arrange(dest, -dep_delay) %>%
  dplyr::group_by(dest) %>%
  dplyr::summarise(
    tot_dep_delay = sum(dep_delay)
  ) %>%
  dplyr::mutate(
    tot_dep_delay_days = ((tot_dep_delay / 60) /24)
  )
```



## Full pipe example

```{r, echo=FALSE, message=FALSE, warning=FALSE}
nycflights13::flights %>%
  dplyr::select(
    origin, dest, dep_delay, arr_delay, 
    year:day
  ) %>%
  dplyr::select(-arr_delay) %>%
  dplyr::filter(month == 11) %>%
  dplyr::filter(!is.na(dep_delay)) %>%
  dplyr::arrange(dest, -dep_delay) %>%
  dplyr::group_by(dest) %>%
  dplyr::summarise(
    tot_dep_delay = sum(dep_delay)
  ) %>%
  dplyr::mutate(
    tot_dep_delay_days = ((tot_dep_delay / 60) /24)
  )
```



## Summary

Data manipulation

- dplyr::arrange
- dplyr::summarise
- dplyr::group_by
- dplyr::mutate

**Next**: Practical session

- Creating R projects
- Creating R scripts
- Data wrangling script

```{r cleanup, include=FALSE}
rm(list = ls())
```
