---
title: "Lecture 303"
author: "granolarr by Dr Stefano De Sabbata<br/>School of Geography, Geology, and the Env.<br/>University of Leicester<br/><a href=\"https://github.com/sdesabbata/granolarr\">github.com/sdesabbata/granolarr</a><br/><a href=\"mailto:s.desabbata@le.ac.uk\">s.desabbata&commat;le.ac.uk</a> &vert; <a href=\"https://twitter.com/maps4thought\">&commat;maps4thought</a><br/>licensed under <a href=\"https://www.gnu.org/licenses/gpl-3.0.html\">GNU GPL v3.0</a>"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    template: ../utils/IOSlides/UoL_Template.html
    logo: ../../src/utils/IOSlides/uol_logo.png
    css: ../utils/RMarkdown/rmarkdown_classes.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("RGDS_HOME"))
rm(list = ls())
```



# Exploring assumptions


## Recap

**Prev**: Descriptive statistics

- stat.desc
- dplyr::across

**Next**: Exploring assumptions

- Normality
- Skewness and kurtosis
- Homogeneity of variance

```{r, echo=FALSE, message=FALSE, warning=FALSE,}
library(tidyverse)
library(magrittr)
library(knitr)
library(pastecs)
library(palmerpenguins)
```


## Normal distribution

:::::: {.cols data-latex=""}

::: {.col style="width: 50%;" data-latex="{0.5\textwidth}"}

- characterized by the bell-shaped curve 
- majority of values lie around the centre of the distribution
- the further the values are from the centre, the lower their frequency
- about 95% of values within 2 standard deviations from the mean

:::

::: {.col style="width: 60%; text-align: right;" data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, fig.height=6, fig.width=6, out.width='100%'}
x <- seq(-4, 4, length=100)
hx <- dnorm(x)

plot(x, hx, type="l", xlab="Values",
  ylab="Density", main="Normal distribution")
```

:::
::::::


## Density histogram

:::::: {.cols data-latex=""}

::: {.col style="width: 50%" data-latex="{0.5\textwidth}"}

```{r, echo=TRUE, eval=FALSE}
palmerpenguins::penguins %>% 
  ggplot2::ggplot(
    aes(x = flipper_length_mm)
  ) +
  ggplot2::geom_histogram(
    aes(
      y =..density..
    )
  ) + 
  ggplot2::stat_function(
    fun = dnorm, 
    args = list(
      # mean and stddev
      # calculations
      # omitted here
      mean = ...,
      sd = ... ),
    colour = "black", size = 1)
```


:::

::: {.col style="width: 50%" data-latex="{0.5\textwidth}"}


```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 5.5, fig.width = 4, out.width='100%'}
palmerpenguins::penguins %>% 
  ggplot2::ggplot(
    aes(
      x = flipper_length_mm
    )
  ) +
  ggplot2::geom_histogram(
    aes(
      y =..density..
    ),
  ) + 
  ggplot2::stat_function(
    fun = dnorm, 
    args = list(
      mean = palmerpenguins::penguins %>% dplyr::filter(!is.na(flipper_length_mm)) %>% dplyr::pull(flipper_length_mm) %>% mean(),
      sd = palmerpenguins::penguins %>% dplyr::filter(!is.na(flipper_length_mm)) %>% dplyr::pull(flipper_length_mm) %>% sd()),
    colour = "red", size = 1)
```

:::
::::::


## Q-Q plot

Values against the cumulative probability of a particular distribution (in this case, *normal* distribution)

:::::: {.cols data-latex=""}

::: {.col style="width: 50%" data-latex="{0.5\textwidth}"}


```{r, echo=TRUE, eval=FALSE}
palmerpenguins::penguins %>% 
  ggplot2::ggplot(
    aes(
      sample = 
        flipper_length_mm
    )
  ) +
  ggplot2::stat_qq() +
  ggplot2::stat_qq_line()
```

:::

::: {.col style="width: 50%" data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 4, fig.width = 4}
palmerpenguins::penguins %>% 
  ggplot2::ggplot(
    aes(sample = flipper_length_mm)
  ) +
  ggplot2::stat_qq() +
  ggplot2::stat_qq_line()
```


:::
::::::


## Normality

**Shapiro–Wilk test** compares the distribution of a variable with a normal distribution having same mean and standard deviation

- If significant, the distribution is not normal
- `shapiro.test` function in `stats`
- or `normtest` values in `pastecs::stat.desc`

```{r, echo=TRUE}
palmerpenguins::penguins %>% 
  dplyr::pull(flipper_length_mm) %>%
  stats::shapiro.test()
```



## Significance

Most statistical tests are based on the idea of hypothesis testing

- a **null hypothesis** is set
- the data are fit into a statistical model
- the model is assessed with a **test statistic**
- the **significance** is the probability of obtaining that test statistic value by chance

The threshold to accept or reject an hypotheis is arbitrary and based on conventions (e.g., *p < .01* or *p < .05*)

**Example:** The null hypotheis of the Shapiro–Wilk test is that the sample is normally distributed and *p < .01* indicates that the probability of that being true is very low. So, the *flipper length* of  penguins in the Palmer Station dataset **is not** normally distributed.


## Example

The *flipper length* of **Adelie** penguins **is normally distributed**

:::::: {.cols data-latex=""}

::: {.col style="width: 50%" data-latex="{0.5\textwidth}"}

```{r, echo=TRUE}
palmerpenguins::penguins %>% 
  filter(
    species == "Adelie"
  ) %>%
  dplyr::pull(
    flipper_length_mm
  ) %>%
  stats::shapiro.test()
```

:::

::: {.col style="width: 50%" data-latex="{0.5\textwidth}"}


```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 5.5, fig.width = 4, out.width='100%'}
palmerpenguins::penguins %>% 
  filter(species == "Adelie") %>%
  ggplot2::ggplot(
    aes(
      x = flipper_length_mm
    )
  ) +
  ggplot2::geom_histogram(
    aes(
      y =..density..
    ),
  ) + 
  ggplot2::stat_function(
    fun = dnorm, 
    args = list(
      mean = palmerpenguins::penguins %>% dplyr::filter(!is.na(flipper_length_mm) & species == "Adelie") %>% dplyr::pull(flipper_length_mm) %>% mean(),
      sd = palmerpenguins::penguins %>% dplyr::filter(!is.na(flipper_length_mm) & species == "Adelie") %>% dplyr::pull(flipper_length_mm) %>% sd()),
    colour = "red", size = 1)
```

:::
::::::

## Example

The *flipper length* of **Adelie** penguins **is normally distributed**

:::::: {.cols data-latex=""}

::: {.col style="width: 50%" data-latex="{0.5\textwidth}"}

```{r, echo=TRUE}
palmerpenguins::penguins %>% 
  filter(
    species == "Adelie"
  ) %>%
  dplyr::pull(
    flipper_length_mm
  ) %>%
  stats::shapiro.test()
```

:::

::: {.col style="width: 50%" data-latex="{0.5\textwidth}"}


```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 5.5, fig.width = 4, out.width='100%'}
palmerpenguins::penguins %>% 
  filter(species == "Adelie") %>%
  ggplot2::ggplot(
    aes(sample = flipper_length_mm)
  ) +
  ggplot2::stat_qq() +
  ggplot2::stat_qq_line()
```

:::
::::::



## Skewness and kurtosis

In a normal distribution, *skewness* and *kurtosis* should be **zero**

- `skewness`: **skewness** value indicates
  - positive: the distribution is skewed towards the left
  - negative: the distribution is skewed towards the right
- `kurtosis`: **kurtosis** value indicates
  - positive: heavy-tailed distribution
  - negative: flat distribution
- `skew.2SE` and `kurt.2SE`: skewness and kurtosis divided by 2 standard errors. Therefore
  - if `> 1` (or `< -1`) then the stat significant *(p < .05)*
  - if `> 1.29` (or `< -1.29`) then stat significant *(p < .01)*



## Example

*Flipper length* is not normally distributed

- skewed left (skewness positive, `skew.2SE > 1.29`)
- flat distribution (kurtosis negative, `kurt.2SE < -1.29`)

```{r, echo=TRUE, eval=FALSE}
palmerpenguins::penguins %>% 
  dplyr::select(bill_length_mm, bill_depth_mm, flipper_length_mm) %>%
  pastecs::stat.desc(basic = FALSE, desc = FALSE, norm = TRUE)
```
```{r, echo=FALSE}
palmerpenguins::penguins %>% 
  dplyr::select(bill_length_mm, bill_depth_mm, flipper_length_mm) %>%
  pastecs::stat.desc(basic = FALSE, desc = FALSE, norm = TRUE) %>%
  knitr::kable()
```


## Example

Values are instead not significant for **Adelie** penguins

- both `skew.2SE` and `kurt.2SE` between `-1` and `1`

```{r, echo=TRUE, eval=FALSE}
palmerpenguins::penguins %>% 
  filter(species == "Adelie") %>%
  dplyr::select(bill_length_mm, bill_depth_mm, flipper_length_mm) %>%
  pastecs::stat.desc(basic = FALSE, desc = FALSE, norm = TRUE)
```
```{r, echo=FALSE}
palmerpenguins::penguins %>% 
  filter(species == "Adelie") %>%
  dplyr::select(bill_length_mm, bill_depth_mm, flipper_length_mm) %>%
  pastecs::stat.desc(basic = FALSE, desc = FALSE, norm = TRUE) %>%
  knitr::kable()
```



## Homogeneity of variance

**Levene’s test** for equality of variance in different levels

- If significant, the variance is different in different levels

:::::: {.cols data-latex=""}

::: {.col style="width: 60%" data-latex="{0.5\textwidth}"}

```{r, echo=TRUE, eval=FALSE}
library(car)
palmerpenguins::penguins %>% 
  car::leveneTest(
    body_mass_g ~ species, data = .
  )
```

:::

::: {.col style="width: 50%" data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE}
palmerpenguins::penguins %>% 
  ggplot2::ggplot(
    aes(
      x = species, 
      y = body_mass_g
    )
  ) +
  ggplot2::geom_boxplot()
```

:::
::::::

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(car)

palmerpenguins::penguins %>% 
 car::leveneTest(body_mass_g ~ species, data = .)
```


## Summary

Exploring assumptions

- Normality
- Skewness and kurtosis
- Homogeneity of variance

**Next**: Practical session

- Data visualisation
- Descriptive statistics
- Exploring assumptions

```{r cleanup, include=FALSE}
rm(list = ls())
```
