---
title: "Lecture 311"
author: "granolarr by Dr Stefano De Sabbata<br/>School of Geography, Geology, and the Env.<br/>University of Leicester<br/><a href=\"https://github.com/sdesabbata/granolarr\">github.com/sdesabbata/granolarr</a><br/><a href=\"mailto:s.desabbata@le.ac.uk\">s.desabbata&commat;le.ac.uk</a> &vert; <a href=\"https://twitter.com/maps4thought\">&commat;maps4thought</a><br/>licensed under <a href=\"https://www.gnu.org/licenses/gpl-3.0.html\">GNU GPL v3.0</a>"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    template: ../utils/IOSlides/UoL_Template.html
    logo: ../../src/utils/IOSlides/uol_logo.png
    css: ../utils/RMarkdown/rmarkdown_classes.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("RGDS_HOME"))
rm(list = ls())
```


<style type="text/css">
.small_r_all pre{
  font-size: 16px;
  line-height: 18px;
}
.small_r_output pre:not(.prettyprint){
  font-size: 16px;
  line-height: 18px;
}
.verysmall_r_output pre:not(.prettyprint){
  font-size: 12px;
  line-height: 14px;
}
</style>



# Comparing groups


## Recap

**Prev**: Exploratory data analysis

- 301 Lecture Data visualisation
- 302 Lecture Descriptive statistics
- 303 Lecture Exploring assumptions
- 304 Practical session

**Now**: Comparing groups

- T-test
- ANOVA

```{r, echo=FALSE, message=FALSE, warning=FALSE,}
library(tidyverse)
library(magrittr)  
library(nycflights13)
```



## Iris

:::::: {.cols data-latex=""}

::: {.col style="width: 80%;" data-latex="{0.5\textwidth}"}

A classic R dataset

- 3 species of iris
- 50 flowers per species
- 4 measurements
  - sepal length and width
  - petal length and width

<br/>
Fisher, R.A., 1936. The use of multiple measurements in taxonomic problems. Annals of eugenics, 7(2), pp.179-188.

:::

::: {.col style="width: 70%; text-align: right;" data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 3.5, fig.height = 5}
iris %>%
  ggplot2::ggplot(
    aes(
      x = Species, 
      y = Petal.Length
    )
  ) +
  ggplot2::geom_boxplot()
```

:::
::::::



## Independent T-test

Are two group means  different?

- null hypothesis
  - there is **no difference** between the groups
- if *p-value* (significance) below threshold (e.g., `0.05` or `0.01`)
  - **group means are different**
- assumptions
    - normally distributed values in groups
    - homogeneity of variance of values in groups
      - if groups have different sizes
    - independence of groups
      - e.g. different conditions of an experiment



## Independent T-test

Independent T-test as a general linear model

**General linear model**

- observation $i$ can be predicted by a $model$ (predictors)
- accounting for some amount of error

$$outcome_i = (model) + error_i $$

**Independent T-test**

- groups is the predictor (categorical variable)
- single observation value as group mean plus error

$$outcome_i = (group\ mean) + error_i $$


## Example: Petal lengths

:::::: {.cols data-latex=""}

::: {.col style="width: 80%;" data-latex="{0.5\textwidth}"}

Are the petal lengths of *versicolor* and *virginica* different?

1. Check assumptions
    1. Indipendent groups: ok
    2. normal distribution: check using Shapiro-Wilk test
    3. homogeneity of variance: not necessary
2. Run T-test
    1. `stats::t.test`

:::

::: {.col style="width: 70%; text-align: right;" data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 3.5, fig.height = 5}
iris %>%
  dplyr::filter(Species %in% c("versicolor", "virginica")) %>%
  ggplot2::ggplot(
    aes(
      x = Species, 
      y = Petal.Length
    )
  ) +
  ggplot2::geom_boxplot()
```

:::
::::::



## Assumptions: normality

Values are normally distributed for both groups

<div class="small_r_all">

```{r, echo=TRUE, message=FALSE, warning=FALSE}
iris %>% dplyr::filter(Species == "versicolor") %>% 
  dplyr::pull(Petal.Length) %>% stats::shapiro.test()

iris %>% dplyr::filter(Species == "virginica") %>% 
  dplyr::pull(Petal.Length) %>% stats::shapiro.test()
```

</div>



## stats::t.test

The test is significant, the group means are different

<div class="small_r_output">

```{r, echo=TRUE, message=FALSE, warning=FALSE}
iris %>%
  dplyr::filter(Species %in% c("versicolor", "virginica")) %$%
  stats::t.test(Petal.Length ~ Species)
```

</div>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
iris_t_test <- iris %>%
  dplyr::filter(Species %in% c("versicolor", "virginica")) %$%
  stats::t.test(Petal.Length ~ Species)
```

How to report: 

- *t*(`r iris_t_test %$% parameter[["df"]] %>% round(digits = 2)`) = `r iris_t_test %$% statistic[["t"]] %>% round(digits = 2)`, *p* < .01



## ANalysis Of VAriance

ANOVA is similar to the T-tests, but more than two groups

- null hypothesis
  - there is **no difference** between the groups
- if *p-value* (significance) below threshold (e.g., `0.05` or `0.01`)
  - **group means are different**
- assumptions
    - normally distributed values in groups
        - especially if groups have different sizes
    - homogeneity of variance of values in groups
      - if groups have different sizes
    - independence of groups
      - e.g. different conditions of an experiment
      
      

## ANalysis Of VAriance

ANOVA as a general linear model

**General linear model**

- observation $i$ can be predicted by a $model$ (predictors)
- accounting for some amount of error

$$outcome_i = (model) + error_i $$

**ANOVA**

- groups is the predictor (categorical variable)
- single observation value as group mean plus error

$$outcome_i = (group\ mean) + error_i $$



## Example: Petal lengths

:::::: {.cols data-latex=""}

::: {.col style="width: 80%;" data-latex="{0.5\textwidth}"}

Are the petal lengths different between all three species?

1. Check assumptions
    1. Indipendent groups: ok
    2. normal distribution: check using Shapiro-Wilk test
    3. homogeneity of variance: not necessary
2. Run ANOVA
    1. `stats::aov`

:::

::: {.col style="width: 70%; text-align: right;" data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 3.5, fig.height = 5}
iris %>%
  ggplot2::ggplot(
    aes(
      x = Species, 
      y = Petal.Length
    )
  ) +
  ggplot2::geom_boxplot()
```

:::
::::::



## Assumptions: normality

We already checked normality for *versicolor* and *virginica*.

Are values for *setosa* normally distributed?

<div class="small_r_all">

```{r, echo=TRUE, message=FALSE, warning=FALSE}
iris %>% dplyr::filter(Species == "setosa") %>% 
  dplyr::pull(Petal.Length) %>% stats::shapiro.test()
```

</div>

Values are normally distributed for all three groups

- although significance for *setosa* is borderline


## stats::aov

The test is significant, the group means are different

<div class="small_r_output">

```{r, echo=TRUE, message=FALSE, warning=FALSE}
iris %$%
  stats::aov(Petal.Length ~ Species) %>%
  summary()
```

</div>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
iris_anova_summary <- iris %$%
  stats::aov(Petal.Length ~ Species) %>%
  summary()
```

How to report:

- *F*(`r iris_anova_summary[[1]] %$% Df[1] %>% round(digits = 2)`, `r iris_anova_summary[[1]] %$% Df[2] %>% round(digits = 2)`) = `r iris_anova_summary[[1]][1, 4] %>% round(digits = 2)`, *p* < .01

<!--

## Homogeneity of variance

Just double-checking...

<div class="small_r_all">

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(car)

iris %$%
  leveneTest(Petal.Length, Species)
```

</div>

The test significant

- the variance in petal length is different in different species
-->


## Summary

Comparing groups

- T-test
- ANOVA

**Next**: Correlation

- Pearson’s r
- Spearman’s rho
- Kendall’s tau
- Pairs panel
- Chi-square

```{r cleanup, include=FALSE}
rm(list = ls())
```
