---
title: "Lecture 312"
author: "granolarr by Dr Stefano De Sabbata<br/>School of Geography, Geology, and the Env.<br/>University of Leicester<br/><a href=\"https://github.com/r-for-geographic-data-science/granolarr\">github.com/sdesabbata/r-for-geographic-data-science</a><br/><a href=\"mailto:s.desabbata@le.ac.uk\">s.desabbata&commat;le.ac.uk</a> &vert; <a href=\"https://twitter.com/maps4thought\">&commat;maps4thought</a><br/>text and code licensed under <a href=\"https://www.gnu.org/licenses/gpl-3.0.html\">GNU GPL v3.0</a>"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    template: ../utils/IOSlides/RfGDS_Template.html
    css: ../utils/RMarkdown/rmarkdown_classes.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("RGDS_HOME"))
rm(list = ls())
```



<style type="text/css">
.small_r_all pre{
  font-size: 16px;
  line-height: 18px;
}
.small_r_all pre:not(.prettyprint){
  font-size: 16px;
  line-height: 18px;
}
.verysmall_r_all pre:not(.prettyprint){
  font-size: 12px;
  line-height: 14px;
}
.veryverysmall_r_all pre:not(.prettyprint){
  font-size: 10px;
  line-height: 10px;
}
</style>



# Correlation


## Recap

**Prev**: Comparing groups

- T-test
- ANOVA

**Now**: Correlation

- Pearson’s r
- Spearman’s rho
- Kendall’s tau
- Pairs panel
- Chi-square

```{r, echo=FALSE, message=FALSE, warning=FALSE,}
library(tidyverse)
library(magrittr)  
library(nycflights13)
```



## Correlation

Two continuous variables can be

- not related at all
- related
    - positively: 
      - entities with *high values* in one 
      - tend to have *high values* in the other
    - negatively: 
      - entities with *high values* in one 
      - tend to have *low values* in the other

**Correlation** is a standardised measure of covariance


## Correlation

Three different approaches

- **Pearson's r**
  - if two variables are **normally distributed**
- **Spearman’s rho**
  - if two variables are **not normally distributed**
- **Kendall’s tau**
  - if **not normally distributed** 
  - and there are a **large number of ties**



## Example

Are flipper length and body mass related in Chinstrap penguins?

<center>
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 6, fig.height = 4}
palmerpenguins::penguins %>%
  dplyr::filter(species == "Chinstrap") %>%
  ggplot2::ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  ggplot2::geom_point(aes(color = species,
                 shape = species),
             size = 2) +
  ggplot2::scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  ggplot2::theme_bw()
```
</center>



## Pearson’s r

If  two variables are **normally distributed**, use **Pearson's r**

- null hypothesis
  - there is no relationship between the variables
- assumptions
  - variables are normally distributed

The square of the correlation value indicates the percentage of shared variance


## Assumptions: normality

Flipper length and body mass are normally distributed in Chinstrap penguins

<div class="small_r_all">

```{r, echo=TRUE, message=FALSE, warning=FALSE}
palmerpenguins::penguins %>%
  dplyr::filter(species == "Chinstrap") %>%
  dplyr::pull(flipper_length_mm) %>% stats::shapiro.test()

palmerpenguins::penguins %>%
  dplyr::filter(species == "Chinstrap") %>%
  dplyr::pull(body_mass_g) %>% stats::shapiro.test()
```

</div>


## stats::cor.test

```{r, echo=FALSE, message=FALSE, warning=FALSE}
palmerpenguins::penguins %>%
  dplyr::filter(species == "Chinstrap") %$%
  stats::cor.test(flipper_length_mm, body_mass_g) ->
  chinstrap_flip_mass_cor
```

- Flipper length and body mass are related
  - `p-value < 0.01`
- sharing `r ((chinstrap_flip_mass_cor$estimate ^ 2) * 100) %>% round(digits = 1)`% of variance
  - `r chinstrap_flip_mass_cor$estimate %>% round(digits = 3)` `^ 2 =` `r chinstrap_flip_mass_cor$estimate ^ 2 %>% round(digits = 3)`

<div class="small_r_all">

```{r, echo=TRUE, message=FALSE, warning=FALSE}
palmerpenguins::penguins %>%
  dplyr::filter(species == "Chinstrap") %$%
  stats::cor.test(flipper_length_mm, body_mass_g)
```

</div>


## Example

But, are flipper length and body mass related in penguins (without considering species as separated groups)?

<center>
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 6, fig.height = 4}
palmerpenguins::penguins %>%
  ggplot2::ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  ggplot2::geom_point(aes(color = species,
                 shape = species),
             size = 2) +
  ggplot2::scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  ggplot2::theme_bw()
```
</center>



## Assumptions: normality

Flipper length and body mass are not normally distributed when all penguins are taken into account as a single group

<div class="small_r_all">

```{r, echo=TRUE, message=FALSE, warning=FALSE}
palmerpenguins::penguins %>%
  dplyr::pull(flipper_length_mm) %>% stats::shapiro.test()

palmerpenguins::penguins %>%
  dplyr::pull(body_mass_g) %>% stats::shapiro.test()
```

</div>


## Spearman’s rho

If  two variables are **not normally distrib.**, use **Spearman’s rho**

- null hypothesis
  - there is no relationship between the variables
- non-parametric
  - uses full dataset to calculate the statistics
  - rather than estimate key parameters of distributions from data
- based on rank difference
- assumptions
  - ties are uncommon

The square of the correlation value indicates the percentage of shared variance


## stats::cor.test (method = "spearman")

```{r, echo=FALSE, message=FALSE, warning=FALSE}
palmerpenguins::penguins %$%
  stats::cor.test(flipper_length_mm, body_mass_g, method = "spearman") ->
  allpenguins_flip_mass_cor
```

- Flipper length and body mass are related
  - `p-value < 0.01`
- sharing `r ((allpenguins_flip_mass_cor$estimate ^ 2) * 100) %>% round(digits = 1)`% of variance
  - `r allpenguins_flip_mass_cor$estimate %>% round(digits = 3)` `^ 2 =` `r allpenguins_flip_mass_cor$estimate ^ 2 %>% round(digits = 3)`

<div class="small_r_all">

```{r, echo=TRUE, message=FALSE, warning=FALSE}
palmerpenguins::penguins %$%
  stats::cor.test(flipper_length_mm, body_mass_g, method = "spearman")
```

</div>


## Correlation with ties

Spearman's rho

- `Cannot compute exact p-value with ties`

<div class="small_r_all">

```{r, echo=TRUE, eval=FALSE}
palmerpenguins::penguins %$%
  stats::cor.test(flipper_length_mm, body_mass_g, method = "spearman")
```

```{r, echo=TRUE, eval=FALSE}
## Warning in cor.test.default(flipper_length_mm, body_mass_g, method = "spearman"):
##  Cannot compute exact p-value with ties
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
palmerpenguins::penguins %$%
  stats::cor.test(flipper_length_mm, body_mass_g, method = "spearman")
```

</div>


## Kendall’s tau

If  two variables are **not normally distributed** and there are **many ties**, use **Kendall’s tau**

- null hypothesis
  - there is no relationship between the variables
- non-parametric
- based on rank difference
- no assumptions
- less *powerful*
  - even if there is a relationship, significance might be high

The square of the correlation value indicates the percentage of shared variance


## stats::cor.test (method = "kendall")

```{r, echo=FALSE, message=FALSE, warning=FALSE}
palmerpenguins::penguins %$%
  stats::cor.test(flipper_length_mm, body_mass_g, method = "kendall")->
  allpenguins_flip_mass_cor_kendall
```

- Flipper length and body mass are related
  - `p-value < 0.01`
- sharing `r ((allpenguins_flip_mass_cor_kendall$estimate ^ 2) * 100) %>% round(digits = 1)`% of variance
  - `r allpenguins_flip_mass_cor_kendall$estimate %>% round(digits = 3)` `^ 2 =` `r allpenguins_flip_mass_cor_kendall$estimate ^ 2 %>% round(digits = 3)`

<div class="small_r_all">

```{r, echo=TRUE, message=FALSE, warning=FALSE}
palmerpenguins::penguins %$%
  stats::cor.test(flipper_length_mm, body_mass_g, method = "kendall")
```

</div>



## psych::pairs.panels

:::::: {.cols data-latex=""}

::: {.col style="width: 60%;" data-latex="{0.5\textwidth}"}

Combining: 

- histograms
- scatter plots 
- correlations

<div class="small_r_all">

```{r, eval=FALSE, echo=TRUE, message=FALSE, warning=FALSE}
library(psych)

palmerpenguins::penguins %>%
  dplyr::select(
    bill_length_mm,
    bill_depth_mm,
    flipper_length_mm, 
    body_mass_g
  ) %>%
  psych::pairs.panels(
    method = "kendall",
    stars = TRUE
  )

## Signif.: 0 '***' 0.001 '**' 0.01
## 0.01 '*' 0.05 0.05 '.' 0.1 ' ' 1
```

</div>

:::

::: {.col style="width: 80%; text-align: right;" data-latex="{0.5\textwidth}"}

<center>
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 6, fig.height = 6}
library(psych)

palmerpenguins::penguins %>%
  dplyr::select(
    bill_length_mm,
    bill_depth_mm,
    flipper_length_mm, 
    body_mass_g
  ) %>%
  psych::pairs.panels(
    method = "kendall",
    stars = TRUE
  )
```

</center>

:::
::::::

</div>



## Chi-squre

How to test the correlation between two **categorical** variables?

Chi-square test:

- null hypothesis
  - there is no relationship between the variables
- non-parametric
- based on cross-tabulated expected counts
- no assumptions

```{r, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
library(gmodels)
palmerpenguins::penguins %$%
  gmodels::CrossTable(
    island, species, chisq = TRUE, expected = TRUE, prop.c = FALSE, 
    prop.t = FALSE, prop.chisq = FALSE, sresid = TRUE, format = "SPSS")
```


## gmodels::CrossTable

There is a relationship (`p-value < 0.01`), different islands have different amounts of penguins from different species

<div class="veryverysmall_r_all">

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(gmodels)

palmerpenguins::penguins %$%
  gmodels::CrossTable(island, species, chisq = TRUE, expected = TRUE, prop.c = FALSE, prop.t = FALSE, prop.chisq = FALSE, sresid = TRUE, format = "SPSS")
```

</div>



## Summary

Correlation

- Pearson’s r
- Spearman’s rho
- Kendall’s tau
- Pairs plot
- Chi-square

**Next**: Data transformations

- Z-scores
- Logarithmic transformations
- Inverse hyperbolic sine transformations

```{r cleanup, include=FALSE}
rm(list = ls())
```
