---
title: "Lecture 313"
author: "granolarr by Dr Stefano De Sabbata<br/>School of Geography, Geology, and the Env.<br/>University of Leicester<br/><a href=\"https://github.com/sdesabbata/granolarr\">github.com/sdesabbata/granolarr</a><br/><a href=\"mailto:s.desabbata@le.ac.uk\">s.desabbata&commat;le.ac.uk</a> &vert; <a href=\"https://twitter.com/maps4thought\">&commat;maps4thought</a><br/>licensed under <a href=\"https://www.gnu.org/licenses/gpl-3.0.html\">GNU GPL v3.0</a>"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    template: ../utils/IOSlides/UoL_Template.html
    logo: ../../src/utils/IOSlides/uol_logo.png
    css: ../utils/RMarkdown/rmarkdown_classes.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("RGDS_HOME"))
rm(list = ls())
```



<style type="text/css">
.small_r_all pre{
  font-size: 16px;
  line-height: 18px;
}
.small_r_output pre:not(.prettyprint){
  font-size: 16px;
  line-height: 18px;
}
.verysmall_r_output pre:not(.prettyprint){
  font-size: 12px;
  line-height: 14px;
}
</style>



# Data transformations


## Recap

**Prev**: Correlation

- Pearson’s r
- Spearman’s rho
- Kendall’s tau
- Pairs panel
- Chi-square

**Now**: Data transformations

- Z-scores
- Logarithmic transformations
- Inverse hyperbolic sine transformations


```{r, echo=FALSE, message=FALSE, warning=FALSE,}
library(tidyverse)
library(magrittr)  
library(nycflights13)
library(palmerpenguins)
library(pastecs)

leicester_2011OAC <- readr::read_csv(paste0(Sys.getenv("RGDS_HOME"), "/data/", "2011_OAC_Raw_uVariables_Leicester.csv"))
```




## Z-scores

Transform the values as relative to

- the distribution's mean 
- and standard deviation

- the z-score of a value i-th $x_i$ is calculated as below, where
  - $\mu$ is the distribution's mean
  - $\sigma$ is the distribution's standard deviation

$$
z_i = \frac{x_i - \mu}{\sigma} 
$$

Commonly used to render two variables easier to compare



## Example

Distribution of flipper lengths in Palmer's penguins

```{r, echo=TRUE, message=FALSE, warning=FALSE, fig.width = 6, fig.height = 3}
palmerpenguins::penguins %>%
  ggplot2::ggplot(aes(x = flipper_length_mm)) +
  ggplot2::geom_histogram() + ggplot2::theme_bw()
```


## base::scale

Distribution of **zscores** derived from flipper lengths

```{r, echo=TRUE, message=FALSE, warning=FALSE, fig.width = 6, fig.height = 3}
palmerpenguins::penguins %>%
  dplyr::mutate(flipper_length_zscore = scale(flipper_length_mm)) %>%
  ggplot2::ggplot(aes(x = flipper_length_zscore)) +
  ggplot2::geom_histogram() + ggplot2::theme_bw()
```



## base::scale

Distribution of **zscores** derived from body mass

```{r, echo=TRUE, message=FALSE, warning=FALSE, fig.width = 6, fig.height = 3}
palmerpenguins::penguins %>%
  dplyr::mutate(body_mass_g_zscore = scale(body_mass_g)) %>%
  ggplot2::ggplot(aes(x = body_mass_g_zscore)) +
  ggplot2::geom_histogram() + ggplot2::theme_bw()
```



## Comparison

But, are flipper length and body mass related in penguins (without considering species as separated groups)?

<center>
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 6, fig.height = 4}
palmerpenguins::penguins %>%
  dplyr::mutate(flipper_length_zscore = scale(flipper_length_mm)) %>%
  dplyr::mutate(body_mass_g_zscore = scale(body_mass_g)) %>%
  ggplot2::ggplot(aes(x = flipper_length_zscore, y = body_mass_g_zscore)) +
  ggplot2::geom_point(aes(color = species,
                 shape = species),
             size = 2) +
  ggplot2::scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  ggplot2::theme_bw()
```
</center>



## Log transformation

**Logarithmic** transformations are useful to *"un-skew"* variables

Common approaches include:

- natural logarithm (`log`)
- binary logarithm (`log2`)
- logarithm base 10 (`log10`)

Only possible on values `> 0`

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 2, fig.width = 6, fig.align = 'center'}
ggplot2::ggplot() + 
  ggplot2::stat_function(
    fun = log10
  ) +
  ggplot2::xlim(1, 1000) +
  ggplot2::xlab("Original values") +
  ggplot2::ylab("Logarithm base 10") +
  ggplot2::theme_bw()
```


## Example

:::::: {.cols data-latex=""}

::: {.col style="width: 80%;" data-latex="{0.5\textwidth}"}

The number of residents aged 20 to 24 (`u011`) in the areas of Leicester described as *"Cosmopolitans"* by the [2011 Output Area Classification](https://www.gov.uk/government/statistics/2011-area-classification-for-super-output-areas) is skewed

:::

::: {.col style="width: 80%; text-align: right;" data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 4, fig.width = 6}
leicester_2011OAC %>% 
  dplyr::filter(supgrpname == "Cosmopolitans") %>%
  ggplot2::ggplot(aes(x = u011)) +
  ggplot2::geom_histogram() + ggplot2::theme_bw()
```

:::

::::::

```{r, echo=FALSE, message=FALSE, warning=FALSE}
leicester_2011OAC %>% 
  dplyr::filter(supgrpname == "Cosmopolitans") %>% 
  dplyr::select(u011) %>% 
  pastecs::stat.desc(basic = FALSE, desc = FALSE, norm = TRUE) %>%
  knitr::kable(digits = 3)
```


## Example

:::::: {.cols data-latex=""}

::: {.col style="width: 80%;" data-latex="{0.5\textwidth}"}

However, it's logarithm base 10 is normally distributed, thus it can be used with tests requiring normally distributed values

```{r, echo=TRUE, eval=FALSE}
mutate(log10_u011 = log10(u011))
```

:::

::: {.col style="width: 80%; text-align: right;" data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 4, fig.width = 6}
leicester_2011OAC %>% 
  dplyr::filter(supgrpname == "Cosmopolitans") %>%
  dplyr::mutate(log10_u011 = log10(u011)) %>%
  ggplot2::ggplot(aes(x = log10_u011)) +
  ggplot2::geom_histogram() + theme_bw()
```

:::

::::::

```{r, echo=FALSE, message=FALSE, warning=FALSE}
leicester_2011OAC %>% 
  dplyr::filter(supgrpname == "Cosmopolitans") %>%
  dplyr::mutate(log10_u011 = log10(u011)) %>% 
  dplyr::select(log10_u011) %>% 
  pastecs::stat.desc(basic = FALSE, desc = FALSE, norm = TRUE) %>%
  knitr::kable(digits = 3)
```



## Inverse hyperbolic sine

**Inverse hyperbolic sine** transformations are useful to *"un-skew"* variables

- similar to logarithmic transformations
- defined on all values
- in R: `asinh`

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 2, fig.width = 6, fig.align = 'center'}
ggplot2::ggplot() + 
  ggplot2::stat_function(
    fun = asinh
  ) +
  ggplot2::xlim(-500, 500) +
  ggplot2::xlab("Original values") +
  ggplot2::ylab("Inverse hyperbolic sine") +
  ggplot2::theme_bw()
```


## Example

:::::: {.cols data-latex=""}

::: {.col style="width: 80%;" data-latex="{0.5\textwidth}"}

The Inverse hyperbolic sine is also normally distributed

```{r, echo=TRUE, eval=FALSE}
mutate(ihs_u011 = asinh(u011))
```

:::

::: {.col style="width: 80%; text-align: right;" data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 4, fig.width = 6}
leicester_2011OAC %>% 
  dplyr::filter(supgrpname == "Cosmopolitans") %>%
  dplyr::mutate(ihs_u011 = asinh(u011)) %>%
  ggplot2::ggplot(aes(x = ihs_u011)) +
  ggplot2::geom_histogram() + theme_bw()
```

:::

::::::

```{r, echo=FALSE, message=FALSE, warning=FALSE}
leicester_2011OAC %>% 
  dplyr::filter(supgrpname == "Cosmopolitans") %>%
  dplyr::mutate(log10_u011 = log10(u011)) %>% 
  dplyr::select(log10_u011) %>% 
  pastecs::stat.desc(basic = FALSE, desc = FALSE, norm = TRUE) %>%
  knitr::kable(digits = 3)
```



## Example

Logarithmic transformation can't be applied to arrival delays in the New York City 2013 flights dataset

- skewed towards left
- but there are values lower or equal to zero


```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 6, fig.height = 3}
nycflights13::flights %>%
  dplyr::filter(!is.na(dep_delay), !is.na(arr_delay), month == 11, day ==20) %>%
  ggplot2::ggplot(aes(x = dep_delay)) +
  ggplot2::geom_histogram()
```



## Example

:::::: {.cols data-latex=""}

::: {.col style="width: 80%;" data-latex="{0.5\textwidth}"}

*Inverse hyperbolic sine* can still be applied. Here it partially unskews the distribution

```{r, echo=TRUE, eval=FALSE}
mutate(
  dep_delay_ihs = 
    asinh(dep_delay)
)
```

:::

::: {.col style="width: 80%; text-align: right;" data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 4, fig.width = 6}
nycflights13::flights %>%
  dplyr::filter(!is.na(dep_delay), !is.na(arr_delay), month == 11, day ==20) %>%
  dplyr::mutate(
    dep_delay_ihs = 
      asinh(dep_delay)
  ) %>%
  ggplot2::ggplot(
    aes(x = dep_delay_ihs)) + 
  ggplot2::geom_histogram()
```

:::

::::::

```{r, echo=FALSE, message=FALSE, warning=FALSE}
nycflights13::flights %>%
  dplyr::filter(!is.na(dep_delay), !is.na(arr_delay), month == 11, day ==20) %>%
  dplyr::mutate(
    dep_delay_ihs = 
      asinh(dep_delay)
  ) %>%
  dplyr::select(dep_delay_ihs) %>%
  pastecs::stat.desc(basic = FALSE, desc = FALSE, norm = TRUE) %>%
  knitr::kable(digits = 3)
```




## Summary

Data transformations

- Z-scores
- Logarithmic transformations
- Inverse hyperbolic sine transformations

**Next**: Practical session

- Comparing means
- Correlation

```{r cleanup, include=FALSE}
rm(list = ls())
```
