---
title: "Lecture 401"
author: "Dr Stefano De Sabbata<br/><small>School of Geography, Geology, and the Env., University of Leicester<br/><a href=\"https://github.com/sdesabbata/r-for-geographic-data-science\" style=\"color: white\">github.com/sdesabbata/r-for-geographic-data-science</a><br/><a href=\"mailto:s.desabbata@le.ac.uk\" style=\"color: white\">s.desabbata&commat;le.ac.uk</a> &vert; <a href=\"https://twitter.com/maps4thought\" style=\"color: white\">&commat;maps4thought</a><br/>text licensed under <a href=\"https://creativecommons.org/licenses/by-sa/4.0/\" style=\"color: white\">CC BY-SA 4.0</a>, code licensed under <a href=\"https://www.gnu.org/licenses/gpl-3.0.html\" style=\"color: white\">GNU GPL v3.0</a></small>"
institute: ""
date: "<small>2022-12-06</small>"
output:
  xaringan::moon_reader:
    self_contained: true
    css: ["css/sdesabbata-uol.css", "css/sdesabbata-uol-fonts.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "macros/my_remark_macros.js"
---
class: inverse, center, middle

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("RGDS_HOME"))
rm(list = ls())
```

# Control structures


---
## Control structures

Building blocks of programming

.pull-left[

- So far, *simple scripting*
   - series of single steps to achieve an output
   - independent from input or current environment

- Control flow
  - actual (imperative) programming
  - including
    - choice statements
    - loops

]
.pull-right[


]

---
## If

<br/>

Format: **if** (*condition*) *statement*

- *condition*: expression returning a logic value (`TRUE` or `FALSE`)
- *statement*: any valid R statement
- *statement* only executed if *condition* is `TRUE`


```{r, echo=TRUE}
a_value <- -7
if (a_value < 0) cat("Negative")
a_value <- 8
if (a_value < 0) cat("Negative")
```


---
## Else

<br/>

Format: **if** (*condition*) *statement1* **else** *statement2*

- *condition*: expression returning a logic value (`TRUE` or `FALSE`)
- *statement1* and *statement2*: any valid R statements
- *statement1*  executed if *condition* is `TRUE`
- *statement2*  executed if *condition* is `FALSE`


```{r, echo=TRUE}
a_value <- -7
if (a_value < 0) cat("Negative") else cat("Positive")
a_value <- 8
if (a_value < 0) cat("Negative") else cat("Positive")
```


---
## Code blocks

**Code blocks** allow to encapsulate **several** statements in a single group

- `{` and `}` contain code blocks
- the statements are execute together

.pull-left[

```{r, echo=TRUE}
first_value <- 8
second_value <- 5

if (first_value > second_value) {
  cat("The first value is higher\n")
  difference <- first_value - second_value
  cat("The difference is ", difference, "\n")
}
```

]
.pull-right[

```{r, echo=TRUE}
first_value <- 2
second_value <- 5

if (first_value > second_value) {
  cat("The first value is higher\n")
  difference <- first_value - second_value
  cat("The difference is ", difference, "\n")
} else {
  cat("The first value is lower or equal\n")
  difference <- first_value - second_value
  cat("The difference is ", difference, "\n")
}
```

]



---
## dplyr::if_else

Allows to use conditional statements in `mutate`

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)

leicester_2011OAC <- 
  read_csv(paste0(Sys.getenv("RGDS_HOME"), "/data/", "2011_OAC_Raw_uVariables_Leicester.csv"))
```

.pull-left[

```{r, echo=TRUE, eval=FALSE}
leicester_2011OAC %>% 
  select(OA11CD, u006) %>% 
  slice_head(n = 5) %>% 
  kable()
```

]
.pull-right[

```{r, echo=TRUE, eval=FALSE}
leicester_2011OAC %>% 
  select(OA11CD, u006) %>% 
  mutate(
    density_cat = if_else(
      u006 < 100,
      "low density", "high density"
    )
  ) %>% 
  slice_head(n = 5) %>% 
  kable()
```

]

.pull-left[

```{r, echo=FALSE, eval=TRUE}
leicester_2011OAC %>% 
  select(OA11CD, u006) %>% 
  slice_head(n = 5) %>% 
  kable()
```

]
.pull-right[

```{r, echo=FALSE, eval=TRUE}
leicester_2011OAC %>% 
  select(OA11CD, u006) %>% 
  mutate(
    density_cat = if_else(
      u006 < 100,
      "low density", 
      "high density"
    )
  ) %>% 
  slice_head(n = 5) %>% 
  kable()
```

]


---
## Loops
Loops are a fundamental component of (procedural) programming.


There are two main types of loops:

- **conditional** loops 
  - executed as long as a defined condition holds true
  - construct `while`
  - construct `repeat`
- **deterministic** loops 
  - are executed a pre-determined number of times
  - construct `for`


---
## While

The *while* construct

- Defined using the `while` reserved word
  - followed by the conditional statement between simple brackets
  - and a code block
- The instructions in the code block are re-executed 
  - as long as the result of the evaluation of the conditional statement is `TRUE`

```{r, echo=TRUE}
current_value <- 0

while (current_value < 3) {
  cat("Current value is", current_value, "\n")
  current_value <- current_value + 1
}
```

<!--
## Repeat

The *repeat* construct can be defined using the `repeat` reserved word, followed by a code block. The instructions in the code block are re-executed until the command `break` is given. The latter is currently given through an `if` construct, which tests the condition that would stop the loop.

```{r, echo=TRUE}
current_value <- 0

repeat {
  cat("Current value is", current_value, "\n")
  current_value <- current_value + 1
  if (current_value >= 3) break
}
```


## While vs Repeat
The difference between `while` and `repeat` is mostly syntactical.

- Sometimes one or the other might fit better with the algorithm you have in mind
- Use the one that comes easier to you in the given situation
-->

---
## For

The *for* construct 

- Defined using the `for` reserved word
  - followed by the definition of an **iterator**
    - a variable which is assigned temporarily
    - with values from a vector
  - followed by a code block
- The construct iterates through all elements of the vector
  - the code block instructions are re-executed 
  - once for each element of the vector.

```{r, echo=TRUE}
cities <- c("Derby", "Leicester", "Lincoln", "Nottingham")
for (city in cities) {
  cat("Do you live in", city, "?\n")
}
```


---
## Example: histogram for all supergrups

.pull-left[

For instance

- we can loop through all the 2011OAC supergrups
- plot a histogra of population for each one of them

```{r, echo=TRUE, eval=FALSE}
for (
  current_supgrp in
    leicester_2011OAC %>% 
      pull(supgrpname) %>% 
      unique()
  ) {
  current_hist <-
    leicester_2011OAC %>% 
    filter(supgrpname == current_supgrp) %>% 
    ggplot(aes(
      x = Total_Population
    )) +
    geom_histogram() +
    ggtitle(current_supgrp) +
    theme_bw() 
  print(current_hist)
}
```

]
.pull-right[

```{r, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE, fig.height=1.8, fig.width=3}
for (
  current_supgrp in
    leicester_2011OAC %>% 
      pull(supgrpname) %>% 
      unique()
  ) {
  current_hist <-
    leicester_2011OAC %>% 
    filter(supgrpname == current_supgrp) %>% 
    ggplot(aes(
      x = Total_Population
    )) +
    geom_histogram() +
    ggtitle(current_supgrp) +
    theme_bw() 
  print(current_hist)
}
```

]


---
## Example: testing normality for all supergrups

.pull-left[

For instance

- we can loop through all the 2011OAC supergrups
- test the normality of each distribution

```{r, echo=TRUE, eval=FALSE}
for (
  current_supgrp in
    leicester_2011OAC %>% 
      pull(supgrpname) %>% 
      unique()
  ) {
  
  print(current_supgrp)
  leicester_2011OAC %>% 
    filter(supgrpname == current_supgrp) %>% 
    pull(Total_Population) %>%
    shapiro.test() %>% 
    print()
  
}
```

]
.pull-right[

```{r, echo=FALSE, eval=TRUE}
for (
  current_supgrp in
    leicester_2011OAC %>% 
      pull(supgrpname) %>% 
      unique()
  ) {
  
  print(current_supgrp)
  leicester_2011OAC %>% 
    filter(supgrpname == current_supgrp) %>% 
    pull(Total_Population) %>%
    shapiro.test() %>% 
    print()
  
}
```

]


---
## Loops with conditional statements

.pull-left[

```{r, echo=TRUE, eval=FALSE}
for (
  current_supgrp in
    leicester_2011OAC %>% 
      pull(supgrpname) %>% 
      unique()
  ) {
  
  current_p_value <-
    leicester_2011OAC %>% 
    filter(supgrpname == current_supgrp) %>% 
    pull(Total_Population) %>%
    shapiro.test() %$% 
    p.value
  
  if (current_p_value > 0.01){
    current_hist <-
      leicester_2011OAC %>% 
      filter(supgrpname == current_supgrp) %>% 
      ggplot(aes(
        x = Total_Population
      )) +
      geom_histogram() +
      ggtitle(current_supgrp) +
      theme_bw() 
    print(current_hist)
  } else {
    print(
      paste(
        current_supgrp,
        "is not normally distributed"
      )
    )
  }
  
}
```

]
.pull-right[

```{r, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE, fig.height=1.8, fig.width=3}
library(magrittr)

for (
  current_supgrp in
    leicester_2011OAC %>% 
      pull(supgrpname) %>% 
      unique()
  ) {
  
  current_p_value <-
    leicester_2011OAC %>% 
    filter(supgrpname == current_supgrp) %>% 
    pull(Total_Population) %>%
    shapiro.test() %$% 
    p.value
  
  if (current_p_value > 0.01){
    current_hist <-
      leicester_2011OAC %>% 
      filter(supgrpname == current_supgrp) %>% 
      ggplot(aes(
        x = Total_Population
      )) +
      geom_histogram() +
      ggtitle(current_supgrp) +
      theme_bw() 
    print(current_hist)
  } else {
    print(
      paste(
        current_supgrp,
        "is not normally distributed"
      )
    )
  }
  
}
```

]

---
## Summary

Control structures

- Conditional statements
- Loops

**Next**: Functions

- Defining functions
- Scope of a variable

```{r cleanup, include=FALSE}
rm(list = ls())
```
