---
title: "Lecture 413"
author: "granolarr by Dr Stefano De Sabbata<br/>School of Geography, Geology, and the Env.<br/>University of Leicester<br/><a href=\"https://github.com/r-for-geographic-data-science/granolarr\">github.com/sdesabbata/r-for-geographic-data-science</a><br/><a href=\"mailto:s.desabbata@le.ac.uk\">s.desabbata&commat;le.ac.uk</a> &vert; <a href=\"https://twitter.com/maps4thought\">&commat;maps4thought</a><br/>text and code licensed under <a href=\"https://www.gnu.org/licenses/gpl-3.0.html\">GNU GPL v3.0</a>"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    template: ../utils/IOSlides/RfGDS_Template.html
    css: ../utils/RMarkdown/rmarkdown_classes.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("RGDS_HOME"))
rm(list = ls())

data_to_cluster <- data.frame(
  x_values = c(rnorm(40, 5, 1), rnorm(60, 10, 1), rnorm(20, 12, 3)),
  y_values = c(rnorm(40, 5, 1), rnorm(60, 5, 3), rnorm(20, 15, 1)),
  original_group = c(rep("A", 40), rep("B", 60), rep("C", 20)) )
```



# Hierarchical and density-based clustering



## Recap

**Prev**: Centroid-based clustering

- K-means
- Fuzzy c-means
- Geodemographic classification

**Now**: Hierarchical and density-based clustering

- Hierarchical
- Mixed 
- Density-based

```{r, echo=FALSE}
library(tidyverse)
library(magrittr)
library(e1071)
library(dbscan)
```


## Example

Can we automatically identify the two groups visible in the scatterplot, without any previous knowledge of the groups?


:::::: {.cols data-latex=""}

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=TRUE, message=FALSE, warning=FALSE}
penguins_norm <-
  palmerpenguins::penguins %>%
  dplyr::filter(
    species %in%
      c("Adelie", "Gentoo")
  ) %>%
  dplyr::filter(
    !is.na(body_mass_g) | 
      !is.na(bill_depth_mm)
  ) %>%
  dplyr::mutate(
    body_mass_norm = 
      scale(body_mass_g),
    bill_depth_norm = 
      scale(bill_depth_mm)
  )
```

:::

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 5, fig.height = 5}
penguins_norm %>%
  ggplot2::ggplot(aes(x = body_mass_norm, y = bill_depth_norm)) +
  ggplot2::geom_point(aes(color = species,
                 shape = species),
             size = 2) +
  ggplot2::scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

:::
::::::


## Hierarchical clustering

Bottom-up approach

- rather than splitting objects into clusters
- aggregate from single objects upwards

**Algorithm**: 

- each object is initialised as it's own cluster
- then repeat
  - join the two most similar clusters 
    - based on a distance-based metric
    - e.g., Ward's (1963) approach is based on variance
  - until only one single cluster is achieved

**Limitation**: computationally expensive


## stats::hclust

```{r, echo=TRUE, message=FALSE, warning=FALSE}
penguins_hclust_result <- 
  penguins_norm %>%
  dplyr::select(
    body_mass_norm, 
    bill_depth_norm
  ) %>%
  # Calculate distance matrix
  stats::dist(method="euclidean") %>%
  # Cluster data
  stats::hclust(method="ward.D2")

penguins_bm_bd_hclust <- penguins_norm %>%
  tibble::add_column(
    bm_bd_hclust = stats::cutree(
      penguins_hclust_result, 
      k = 2
    )
  )
```



## clustering tree

Generates a clustering tree (dendrogram), which can then be *"cut"* at the desired height

```{r, echo=FALSE, message=FALSE, warning=FALSE}
penguins_hclust_result %>%
  plot() + 
  abline(h = 20, col = "red")
```



## Hierarchical clustering result

:::::: {.cols data-latex=""}

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 5}
penguins_bm_bd_hclust %>%
  ggplot2::ggplot(aes(x = body_mass_norm, y = bill_depth_norm)) +
  ggplot2::geom_point(aes(color = species,
                 shape = species),
             size = 2) +
  ggplot2::scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

:::

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 5}
penguins_bm_bd_hclust %>%
  ggplot2::ggplot(aes(
    x = body_mass_norm, y = bill_depth_norm, 
    colour = factor(bm_bd_hclust))) +
  ggplot2::geom_point() +
  ggplot2::scale_color_manual(values = c("#d95f02", "#1b9e77")) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

:::
::::::


## Bagged clustering

Bootstrap aggregating (*b-agg-ed*) clustering approach

- first, k-means
  - randomly select a sample
  - calculate K-means
  - repeat on *many* samples
- then, hierarchical 
  - execute hierarchical clustering on the centroids of the clusters generated in the previous step
- finally
  - select required number of clusters
  - assign object to closest centroid
  
<font size="4"> 
Leisch, F., 1999. Bagged clustering.
</font>

## e1071::bclust

```{r, echo=TRUE, message=FALSE, warning=FALSE, results="hide"}
penguins_bclust_result <- 
  penguins_norm %>%
  dplyr::select(body_mass_norm, bill_depth_norm) %>%
  e1071::bclust(
    hclust.method = "ward.D2", 
    resample = TRUE
  )

penguins_bm_bd_bclust <- 
  penguins_norm %>%
  tibble::add_column(
    bm_bd_bclust = e1071::clusters.bclust(
      penguins_bclust_result, 
      2
    )
  )
```

## Bagged clustering result

:::::: {.cols data-latex=""}

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 5}
penguins_bm_bd_bclust %>%
  ggplot2::ggplot(aes(x = body_mass_norm, y = bill_depth_norm)) +
  ggplot2::geom_point(aes(color = species,
                 shape = species),
             size = 2) +
  ggplot2::scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

:::

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 5}
penguins_bm_bd_bclust %>%
  ggplot2::ggplot(aes(
    x = body_mass_norm, y = bill_depth_norm, 
    colour = factor(bm_bd_bclust))) +
  ggplot2::geom_point() +
  ggplot2::scale_color_manual(values = c("#d95f02", "#1b9e77")) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

:::
::::::



## Density based clustering

Density-based spatial clustering of applications with noise (DBSCAN)

- start from a random unclustered point
  - proceed by aggregating its neighbours to the same cluster
    - as long as they are within a certain distance `eps`
- once no more objects can be added
  - select another random point
  - and start aggregating again to a new cluster

**Limitation**: selection of `eps`

<font size="4"> 
Ester, M., Kriegel, H.P., Sander, J. and Xu, X., 1996, August. Density-based spatial clustering of applications with noise. In Int. Conf. Knowledge Discovery and Data Mining (Vol. 240, p. 6).
</font>
  
## dbscan::dbscan

```{r, echo=TRUE, message=FALSE, warning=FALSE}
penguins_dbscan_result <- 
  penguins_norm %>%
  dplyr::select(body_mass_norm, bill_depth_norm) %>%
  dbscan::dbscan(
    eps = 1, 
    minPts = 5
  )

penguins_bm_bd_dbscan <- 
  penguins_norm %>%
  tibble::add_column(
    bm_bd_dbscan = 
      penguins_dbscan_result %$% 
      cluster
  )
```



## DBSCAN result

Using: `dbscan(eps = 1, minPts = 5)`

:::::: {.cols data-latex=""}

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 5}
penguins_bm_bd_dbscan %>%
  ggplot2::ggplot(aes(x = body_mass_norm, y = bill_depth_norm)) +
  ggplot2::geom_point(aes(color = species,
                 shape = species),
             size = 2) +
  ggplot2::scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

:::

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 5}
penguins_bm_bd_dbscan %>%
  ggplot2::ggplot(aes(
    x = body_mass_norm, y = bill_depth_norm, 
    colour = factor(bm_bd_dbscan))) +
  ggplot2::geom_point() +
  ggplot2::scale_color_manual(values = c("#d95f02", "#1b9e77")) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

:::
::::::


## DBSCAN result

Using: `dbscan(eps = 0.5, minPts = 5)`


```{r, echo=FALSE, message=FALSE, warning=FALSE}
penguins_dbscan_result2 <- penguins_norm %>%
  dplyr::select(body_mass_norm, bill_depth_norm) %>%
  dbscan::dbscan(eps = 0.5, minPts = 5)

penguins_bm_bd_dbscan2 <- penguins_norm %>%
  tibble::add_column(bm_bd_dbscan = penguins_dbscan_result2 %$% cluster)
```

:::::: {.cols data-latex=""}

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 5}
penguins_bm_bd_dbscan2 %>%
  ggplot2::ggplot(aes(x = body_mass_norm, y = bill_depth_norm)) +
  ggplot2::geom_point(aes(color = species,
                 shape = species),
             size = 2) +
  ggplot2::scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

:::

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 5}
penguins_bm_bd_dbscan2 %>%
  ggplot(aes(
    x = body_mass_norm, y = bill_depth_norm, 
    colour = factor(bm_bd_dbscan))) +
  ggplot2::geom_point() +
  ggplot2::scale_color_manual(values = c("#d95f02", "#1b9e77", "#7570b3")) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

:::
::::::


## DBSCAN result

Using: `dbscan(eps = 0.1, minPts = 5)`


```{r, echo=FALSE, message=FALSE, warning=FALSE}
penguins_dbscan_result3 <- penguins_norm %>%
  dplyr::select(body_mass_norm, bill_depth_norm) %>%
  dbscan::dbscan(eps = 0.1, minPts = 5)

penguins_bm_bd_dbscan3 <- penguins_norm %>%
  tibble::add_column(bm_bd_dbscan = penguins_dbscan_result3 %$% cluster)
```

:::::: {.cols data-latex=""}

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 5}
penguins_bm_bd_dbscan3 %>%
  ggplot2::ggplot(aes(x = body_mass_norm, y = bill_depth_norm)) +
  ggplot2::geom_point(aes(color = species,
                 shape = species),
             size = 2) +
  ggplot2::scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

:::

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 5}
penguins_bm_bd_dbscan3 %>%
  ggplot2::ggplot(aes(
    x = body_mass_norm, y = bill_depth_norm, 
    colour = factor(bm_bd_dbscan))) +
  ggplot2::geom_point() +
  ggplot2::scale_color_brewer(palette = "Dark2") +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

:::
::::::


## Not alwasy that easy...

```{r, include=FALSE, message=FALSE, warning=FALSE}
penguins_norm_bills <-
  palmerpenguins::penguins %>%
  dplyr::filter(
    !is.na(bill_length_mm) | 
      !is.na(bill_depth_mm)
  ) %>%
  dplyr::mutate(
    bill_length_norm = 
      scale(bill_length_mm),
    bill_depth_norm = 
      scale(bill_depth_mm)
  )

# K-means
penguins_norm_bills_kmeans <- 
 penguins_norm_bills %>%
  dplyr::select(bill_length_norm, bill_depth_norm) %>%
  stats::kmeans(centers = 3,  iter.max = 50)

penguins_norm_bills <- 
  penguins_norm_bills %>%
  tibble::add_column(bills_kmeans = penguins_norm_bills_kmeans %$% cluster)

# Hierarchical
penguins_norm_bills_hclust <- 
  penguins_norm_bills %>%
  dplyr::select(bill_length_norm, bill_depth_norm) %>%
  stats::dist(method="euclidean") %>%
  stats::hclust(method="ward.D2")

penguins_norm_bills <- 
  penguins_norm_bills %>%
  tibble::add_column(bills_hclust = stats::cutree(penguins_norm_bills_hclust, k = 3))


# Bagged
penguins_norm_bills_bclust <- 
  penguins_norm_bills %>%
  dplyr::select(bill_length_norm, bill_depth_norm) %>%
  e1071::bclust(hclust.method="ward.D2", resample = TRUE)

penguins_norm_bills <- 
  penguins_norm_bills %>%
  tibble::add_column(bills_bclust = e1071::clusters.bclust(penguins_norm_bills_bclust, 3))


# DBSCAN
penguins_norm_bills_dbscan <- 
  penguins_norm_bills %>%
  dplyr::select(bill_length_norm, bill_depth_norm) %>%
  dbscan::dbscan(eps = 0.2, minPts = 5)

penguins_norm_bills <- 
  penguins_norm_bills %>%
  tibble::add_column(bills_dbscan = penguins_norm_bills_dbscan %$% cluster)
```

:::::: {.cols data-latex=""}

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 2.75}
penguins_norm_bills %>%
  ggplot2::ggplot(aes(
    x = bill_length_norm, y = bill_depth_norm, 
    colour = factor(bills_kmeans))) +
  ggplot2::geom_point() +
  ggplot2::scale_color_brewer(palette = "Dark2") +
  ggplot2::ggtitle("K-means") +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "none")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 2.75}
penguins_norm_bills %>%
  ggplot2::ggplot(aes(
    x = bill_length_norm, y = bill_depth_norm, 
    colour = factor(bills_hclust))) +
  ggplot2::geom_point() +
  ggplot2::scale_color_brewer(palette = "Dark2") +
  ggtitle("Hierarchical") +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "none")
```

:::

::: {.col data-latex="{0.5\textwidth}"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 2.75}
penguins_norm_bills %>%
  ggplot2::ggplot(aes(
    x = bill_length_norm, y = bill_depth_norm, 
    colour = factor(bills_bclust))) +
  ggplot2::geom_point() +
  ggplot2::scale_color_brewer(palette = "Dark2") +
  ggplot2::ggtitle("Bagged") +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "none")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 4, fig.height = 2.75}
penguins_norm_bills %>%
  ggplot2::ggplot(aes(
    x = bill_length_norm, y = bill_depth_norm, 
    colour = factor(bills_dbscan))) +
  ggplot2::geom_point() +
  ggplot2::scale_color_brewer(palette = "Dark2") +
  ggplot2::ggtitle("DBSCAN") +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "none")
```

:::
::::::



## Summary

Hierarchical and density-based clustering

- Hierarchical
- Mixed 
- Density-based

**Next**: Practical session

- Geodemographic classification

```{r cleanup, include=FALSE}
rm(list = ls())
```
