---
title: "Lecture 213"
author: "granolarr by Dr Stefano De Sabbata<br/>School of Geography, Geology, and the Env.<br/>University of Leicester<br/><a href=\"https://github.com/r-for-geographic-data-science/granolarr\">github.com/sdesabbata/r-for-geographic-data-science</a><br/><a href=\"mailto:s.desabbata@le.ac.uk\">s.desabbata&commat;le.ac.uk</a> &vert; <a href=\"https://twitter.com/maps4thought\">&commat;maps4thought</a><br/>text licensed under <a href=\"https://creativecommons.org/licenses/by-sa/4.0/\">CC BY-SA 4.0</a>, code licensed under <a href=\"https://www.gnu.org/licenses/gpl-3.0.html\">GNU GPL v3.0</a>"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    template: ../utils/IOSlides/RfGDS_Template.html
    css: ../utils/RMarkdown/rmarkdown_classes.css
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("RGDS_HOME"))
rm(list = ls())
```



# Read and write data



## Summary

Tidy-up your data

- Wide and long data
- Re-shape data
- Handle missing values

**Next**: Read and write data

- file formats
- read
- write

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
```



## Text file formats

A series of formats based on plain-text files

For instance

- comma-separated values files `.csv`
- semi-colon-separated values files `.csv`
- tab-separated values files `.tsv`
- other formats using custom delimiters
- fix-width files `.fwf`



## Comma Separated Values

The file `2011_OAC_supgrp_Leicester.csv` contains

- one row for each [Output Area (OA)](https://www.ons.gov.uk/methodology/geography/ukgeographies/censusgeography) in Leicester
- [Lower-Super Output Area (LSOA)](https://www.ons.gov.uk/methodology/geography/ukgeographies/censusgeography) containing the OA
- code and name of the supergroup assigned to the OA by the [2011 Output Area Classification](http://geogale.github.io/2011OAC/)
- total population of the OA

Extract showing only the first few rows

```{}
OA11CD,LSOA11CD,supgrpcode,supgrpname,Total_Population
E00069517,E01013785,6,Suburbanites,313
E00069514,E01013784,2,Cosmopolitans,323
E00169516,E01013713,4,Multicultural Metropolitans,341
E00169048,E01032862,4,Multicultural Metropolitans,345
```



## readr

The [`readr`](https://readr.tidyverse.org/) (pronounced *read-er*) library is part of [`tidyverse`](https://www.tidyverse.org/)

Provides functions to read and write text files

- `readr::read_csv`: comma-separated files `.csv`
- `readr::read_csv2`: semi-colon-separated files `.csv`
- `readr::read_tsv`: tab-separated files `.tsv`
- `readr::read_fwf`: fix-width files `.fwf`
- `readr::read_delim`: files using a custom delimiter

and their *write* counterpart, such as

- `readr::write_csv`: comma-separated files `.csv`


## readr::read_csv

The `readr::read_csv` function of the [`readr`](https://readr.tidyverse.org/index.html) library reads a *csv* file from the path provided as the first argument

```{r, echo=TRUE, eval=FALSE}
leicester_2011OAC <- 
  readr::read_csv("2011_OAC_supgrp_Leicester.csv")

leicester_2011OAC
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
leicester_2011OAC <- readr::read_csv(paste0(Sys.getenv("RGDS_HOME"), "/data/", "2011_OAC_supgrp_Leicester.csv"))
leicester_2011OAC %>% 
  print(n= 3, width = 60)
```



## Read options

Read functions provide options about how to interpret a file contents

- For instance, `readr::read_csv`
  - `col_names`: 
    - `TRUE` or `FALSE` whether top row is column names
    - or a vector of column names
  - `col_types`:
    - a `cols()` specification or a string
  - `skip`: lines to skip before reading data
  - `n_max`: max number of record to read



## Column specifications

- `col_logical()` or `l` as logic values
- `col_integer()` or `i` as integer
- `col_double()` or `d` as numeric (double)
- `col_character()` or `c` as character
- `col_factor(levels, ordered)` or `f` as factor
- `col_date(format = "")` or `D` as data type
- `col_time(format = "")` or `t` as time type
- `col_datetime(format = "")` or `T` as datetime
- `col_number()` or `n` as numeric (dropping marks)
- `col_skip()` or `_` or `-` don’t import
- `col_guess()` or `?` use best type based on the input



## readr::read_csv

Using `readr::read_csv` as in the previous example with no further options will generate the following warning

```{r, echo=TRUE, eval=FALSE}
leicester_2011OAC <- 
  readr::read_csv("2011_OAC_supgrp_Leicester.csv")

leicester_2011OAC
```

```{}
Parsed with column specification:
cols(
  OA11CD = col_character(),
  LSOA11CD = col_character(),
  supgrpcode = col_double(),
  supgrpname = col_character(),
  Total_Population = col_double()
)
```



## readr::read_csv

```{r, echo=TRUE, eval=FALSE}
leicester_2011OAC <- readr::read_csv(
    "2011_OAC_supgrp_Leicester.csv",
    col_types = cols(
      OA11CD  = col_character(),
      LSOA11CD  = col_character(),
      supgrpcode  = col_character(),
      supgrpname  = col_character(),
      Total_Population  = col_integer()
    )
  )
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
leicester_2011OAC <- readr::read_csv(
    paste0(Sys.getenv("RGDS_HOME"), "/data/", "2011_OAC_supgrp_Leicester.csv"),
    col_types = cols(
      OA11CD  = col_character(),
      LSOA11CD  = col_character(),
      supgrpcode  = col_character(),
      supgrpname  = col_character(),
      Total_Population  = col_integer()
    )
  ) 

leicester_2011OAC%>% 
  print(n= 3, width = 60)
```


## readr::read_csv

```{r, echo=TRUE, eval=FALSE}
leicester_2011OAC <- readr::read_csv(
    "2011_OAC_supgrp_Leicester.csv",
    col_types = "cccci"
  )
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
leicester_2011OAC <- readr::read_csv(
    paste0(Sys.getenv("RGDS_HOME"), "/data/", "2011_OAC_supgrp_Leicester.csv"),
    col_types = "cccci"
  ) 

leicester_2011OAC%>% 
  print(n= 7, width = 60)
```


## readr::write_csv

The function `write_csv` can be used to save a dataset to `csv`

Example:

1. **read** the 2011 OAC dataset
2. **select** a few columns
3. **filter** only those OA in the supergroup *Suburbanites* (code `6`)
4. **write** the results to a file named *2011_OAC_supgrp_Leicester_supgrp6.csv*

```{r, echo=TRUE, eval=FALSE}
readr::read_csv("2011_OAC_supgrp_Leicester.csv") %>%
  dplyr::select(OA11CD, supgrpcode, Total_Population) %>%
  dplyr::filter(supgrpcode == "6") %>%
  readr::write_csv("2011_OAC_supgrp_Leicester_supgrp6.csv")
```
```{r, echo=FALSE, message=FALSE, warning=FALSE}
readr::read_csv(
    paste0(Sys.getenv("RGDS_HOME"), "/data/", "2011_OAC_supgrp_Leicester.csv")
  ) %>%
  dplyr::select(OA11CD, supgrpcode, Total_Population) %>%
  dplyr::filter(supgrpcode == "6") %>%
  readr::write_csv(
    paste0(Sys.getenv("RGDS_HOME"), "/data/", "2011_OAC_supgrp_Leicester_supgrp6.csv")
  )
```



## readr::write_tsv

```{r, echo=TRUE, eval=FALSE}
readr::read_csv("2011_OAC_supgrp_Leicester.csv") %>%
  dplyr::select(OA11CD, supgrpcode, Total_Population) %>%
  dplyr::filter(supgrpcode == "6") %>%
  readr::write_tsv("2011_OAC_supgrp_Leicester_supgrp6.tsv")
```
```{r, echo=FALSE, message=FALSE, warning=FALSE}
readr::read_csv(
    paste0(Sys.getenv("RGDS_HOME"), "/data/", "2011_OAC_supgrp_Leicester.csv")
  ) %>%
  dplyr::select(OA11CD, supgrpcode, Total_Population) %>%
  dplyr::filter(supgrpcode == "6") %>%
  readr::write_tsv(
    paste0(Sys.getenv("RGDS_HOME"), "/data/", "2011_OAC_supgrp_Leicester_supgrp6.tsv")
  )
```

```{}
OA11CD      supgrpcode  Total_Population
E00069517 6         313
E00069468 6         251
E00069528 6         270
E00069538 6         307
E00069174 6         321
E00069170 6         353
E00069171 6         351
E00068713 6         265
E00069005 6         391
E00069014 6         316
E00068989 6         354
```




## Other data imports

[Tidyverse](https://www.tidyverse.org/) also imports [other packages for reading data](https://www.tidyverse.org/packages/#import)

- Tabular formats
  - [`readxl`](https://readxl.tidyverse.org/) for Excel (`.xls` and `.xlsx`)
  - [`haven`](https://haven.tidyverse.org/) for SPSS, Stata, and SAS data.
- Databases
  - [`DBI`](https://cran.r-project.org/web/packages/DBI/) for relational databases
- NoSQL
  - [`jsonlite`](https://cran.r-project.org/web/packages/jsonlite/) for JSON
  - [`xml2`](https://cran.r-project.org/web/packages/xml2/) for XML
- Web
  - [`httr`](https://cran.r-project.org/web/packages/httr/) for web APIs

<!--
  - Specific backends require different packages, e.g. [`RPostgreSQL`](https://cran.r-project.org/web/packages/RPostgreSQL/)
  - [`rvest`](https://cran.r-project.org/web/packages/rvest/) for web scraping.
-->



## Summary

Read and write data

- file formats
- read
- write

**Next**: Practical session

- Read and write data
- Tidy data
- Join operations

```{r cleanup, include=FALSE}
rm(list = ls())
```
