---
title: "Lecture 411"
author: "granolarr by Dr Stefano De Sabbata<br/>School of Geography, Geology, and the Env.<br/>University of Leicester<br/><a href=\"https://github.com/r-for-geographic-data-science/granolarr\">github.com/sdesabbata/r-for-geographic-data-science</a><br/><a href=\"mailto:s.desabbata@le.ac.uk\">s.desabbata&commat;le.ac.uk</a> &vert; <a href=\"https://twitter.com/maps4thought\">&commat;maps4thought</a><br/>text licensed under <a href=\"https://creativecommons.org/licenses/by-sa/4.0/\">CC BY-SA 4.0</a>, code licensed under <a href=\"https://www.gnu.org/licenses/gpl-3.0.html\">GNU GPL v3.0</a>"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    template: ../utils/IOSlides/RfGDS_Template.html
    css: ../utils/RMarkdown/rmarkdown_classes.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = Sys.getenv("RGDS_HOME"))
rm(list = ls())
```

<style type="text/css">
.small_r_all pre{
  font-size: 16px;
  line-height: 18px;
}
.small_r_output pre:not(.prettyprint){
  font-size: 16px;
  line-height: 18px;
}
.verysmall_r_output pre:not(.prettyprint){
  font-size: 12px;
  line-height: 14px;
}
</style>



# Principal Component Analysis

## Recap

**Prev**: Comparing data

- 401 Lecture Introduction to Machine Learning
- 402 Lecture Artificial Neural Networks
- 403 Lecture Support vector machines
- 404 Practical session

**Now**: Principal Component Analysis

- Principal components
- `stats::prcomp`
- Dimensionality reduction



```{r, echo=FALSE, message=FALSE, warning=FALSE,}
library(tidyverse)
library(magrittr)  
library(palmerpenguins)
library(patchwork)
```


## Principal components

**Principal component** are 

- a set of directions orthogonal to each other
- that best fit a set of data

Can be interpreted as a *"re-projection"* of the data

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 8, fig.height = 3}
penguins_original_plot <-
 palmerpenguins::penguins %>%
  ggplot2::ggplot(
    aes(
      x = flipper_length_mm, 
      y = body_mass_g
    )
  ) +
  ggplot2::geom_point(
    aes(
      color = species,
      hape = species
    ),
    size = 2
  ) +
  ggplot2::scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "none")

penguins_fb_pca <-
  palmerpenguins::penguins %>%
  dplyr::select(flipper_length_mm, body_mass_g) %>%
  dplyr::filter(!is.na(flipper_length_mm)) %>%
  stats::prcomp(center = TRUE,scale. = TRUE) %$%
  x %>%
  as.data.frame()

penguins_rotated_plot <-
  palmerpenguins::penguins %>%
  dplyr::filter(!is.na(flipper_length_mm)) %>%
  dplyr::bind_cols(
    penguins_fb_pca
  ) %>%
  dplyr::rename(
    `Direction 1` = PC1,
    `Direction 2` = PC2
  ) %>%
  ggplot2::ggplot(
    aes(
      x = `Direction 1`, 
      y = `Direction 2`
    )
  ) +
  ggplot2::geom_point(
    aes(
      color = species,
      hape = species
    ),
    size = 2
  ) +
  ggplot2::scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  #ggplot2::coord_fixed(ratio = 1) +
  ylim(-2, 2) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "none")

(penguins_original_plot | penguins_rotated_plot)
```


## Dimensionality reduction

Alternatively, principal components can be interpreted as 

- **lower-dimensional** representation of the data

Especially useful when working numerous variables

- a limited number of principal components can be retained
  - most variance maintained
  - distance in data space approximated
  - high-dimensional data can be more easily plotted 
- commonly used as dimensionality reduction step
  - supervised learning models
    - linear regression
  - clustering
  

## stats::prcomp

Principal component analysis on body mass, flipper length, and bill length and depth

<div class="small_r_all">

```{r, echo=TRUE, message=FALSE, warning=FALSE}
penguins_pca <-
  palmerpenguins::penguins %>%
  dplyr::select(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) %>%
  # remove missing data
  dplyr::filter(
    !is.na(bill_length_mm) | !is.na(bill_depth_mm) |
    !is.na(flipper_length_mm) | !is.na(body_mass_g)
  ) %>%
  stats::prcomp(center = TRUE, scale. = TRUE) 

summary(penguins_pca)
```

</div>

The first component alone explains `r (summary(penguins_pca)$importance[3, 1] * 100) %>% round(digits = 2)`% of variance, and the first two together explain `r (summary(penguins_pca)$importance[3, 2] * 100) %>% round(digits = 2)`% of variance



## PCA results

<div class="small_r_all">

```{r, echo=TRUE, message=FALSE, warning=FALSE}
penguins_with_pca <- palmerpenguins::penguins %>%
  dplyr::filter(!is.na(bill_length_mm) | !is.na(bill_depth_mm) | 
                !is.na(flipper_length_mm) | !is.na(body_mass_g)) %>%
  dplyr::bind_cols(
      penguins_pca %$% x %>% as.data.frame()
  )
```

</div>

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 5, fig.height = 3.5}
penguins_with_pca %>%
  ggplot2::ggplot(
    aes(
      x = PC1, 
      y = PC2
    )
  ) +
  ggplot2::geom_point(
    aes(
      color = species,
      hape = species
    ),
    size = 2
  ) +
  ggplot2::scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```


## Plotting PCA

```{r, echo=TRUE, message=FALSE, warning=FALSE, fig.width = 8, fig.height = 4}
library(factoextra)

penguins_pca %>% fviz_pca_biplot(label = "var")
```


## Summary

Principal Component Analysis

- Principal components
- `stats::prcomp`
- Interpretation

**Next**: Centroid-based clustering

- K-means
- Fuzzy c-means
- Geodemographic classification


```{r cleanup, include=FALSE}
rm(list = ls())
```
