# Exploratory statistics
<br/><small><a href="javascript:if(window.print)window.print()">Print this chapter</a></small>

This chapter showcases an exploratory analysis of the distribution of people aged 20 to 24 in Leicester, using the `u011` variable from the [2011 Output Area Classification (2011OAC)](https://github.com/geogale/2011OAC) already used in previous chapters. Before continuing, open the project created for the previous chapter. Create new RMarkdown document within that project to replicate the analysis in this document. Once the document is set up, start by adding an R code snipped including the code below, which is loads the 2011OAC dataset and the libraries used for this chapter.

```{r, echo=TRUE, eval=FALSE}
library(tidyverse)
library(knitr)

leicester_2011OAC <- read_csv("2011_OAC_Raw_uVariables_Leicester.csv")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
leicester_2011OAC <- read_csv(paste0(Sys.getenv("RGDS_HOME"), "/data/", "2011_OAC_Raw_uVariables_Leicester.csv"))
```

The graphics above provide preliminary evidence that the distribution of people aged 20 to 24 might, in fact, be different in different 2011 supergroups. In the remainder of the chapter, we are going to explore that hypothesis further. First, load the necessary statistical libraries.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(pastecs)
library(car)
```

The code below calculates the percentage of people aged 20 to 24 (i.e., `u011`) over total population per OA, but it also recodes (see [recode](https://dplyr.tidyverse.org/reference/recode.html)) the names of the 2011OAC supergroups to a shorter 2-letter version, which is useful for the tables presented further below. Note that the code below uses the library name as a prefix `dplyr::` in front of the function name `recode` (thus, `dplyr::recode`) to make sure the function `recode` of the package `car` (thus, `car::recode`) is not used instead.

Only the OA code, the recoded 2011OAC supergroup name, and the newly created `perc_age_20_to_24` are retained in the new table `leic_2011OAC_20to24`. Such a step is sometimes useful as stepping stone for further analysis and can make the code easier to read further down the line. Sometimes it is also a necessary step when interacting with certain libraries, which are not fully compatible with Tidyverse libraries, such as `leveneTest` (more on that function below).

```{r, echo=TRUE, message=FALSE, warning=FALSE}
leic_2011OAC_20to24 <- leicester_2011OAC %>%
  mutate(
    perc_age_20_to_24 = (u011 / Total_Population) * 100,
    supgrpname = 
      dplyr::recode(
        supgrpname, 
          `Suburbanites` = "SU",
          `Cosmopolitans` = "CP",
          `Multicultural Metropolitans` = "MM",
          `Ethnicity Central` = "EC",
          `Constrained City Dwellers` = "CD",
          `Hard-Pressed Living` = "HP",
          `Urbanites` = "UR"
    )
  ) %>%
  select(OA11CD, supgrpname, perc_age_20_to_24)

leic_2011OAC_20to24 %>%
  slice_head(n = 5) %>%
  kable()
```

## Descriptive statistics

The first step of any statistical analysis or modelling should be to explore the *"shape"* of the data involved, by looking at the descriptive statistics of all variables involved. The function `stat.desc` of the `pastecs` library provides three series of descriptive statistics.

- `base`:
    - `nbr.val`: overall number of values in the dataset;
    - `nbr.null`: number of `NULL` values -- NULL is often returned by expressions and functions whose values are undefined;
    - `nbr.na`: number of `NA`s -- missing value indicator;
- `desc`:
    - `min` (see also `min` function): **minimum** value in the dataset;
    - `max` (see also `max` function): **minimum** value in the dataset;
    - `range`: difference between `min` and `max` (different from `range()`);
    - `sum` (see also `sum` function): sum of the values in the dataset;
    - `median` (see also `median` function): **median**, that is the value separating the higher half from the lower half the values
    - `mean` (see also `mean` function): **arithmetic mean**, that is `sum` over the number of values not `NA`;
    - `SE.mean`: **standard error of the mean** -- estimation of the variability of the mean calculated on different samples of the data (see also *central limit theorem*);
    - `CI.mean.0.95`: **95% confidence interval of the mean** -- indicates that there is a 95% probability that the actual mean is within that distance from the sample mean;
    - `var`: **variance** ($\sigma^2$), it quantifies the amount of variation as the average of squared distances from the mean;
    - `std.dev`: **standard deviation** ($\sigma$), it quantifies the amount of variation as the square root of the variance;
    - `coef.var`: **variation coefficient** it quantifies the amount of variation as the standard deviation divided by the mean;
- `norm` (default is `FALSE`, use `norm = TRUE` to include it in the output):
    - `skewness`: **skewness** value indicates
        - positive: the distribution is skewed towards the left;
        - negative: the distribution is skewed towards the right;
    - `kurtosis`: **kurtosis** value indicates:
        - positive: heavy-tailed distribution;
        - negative: flat distribution;
    - `skew.2SE` and `kurt.2SE`: skewness and kurtosis divided by 2 standard errors. If greater than 1, the respective statistics is significant (*p < .05*);
    - `normtest.W`: test statistics for the **Shapiro–Wilk test** for normality;
    - `normtest.p`: significance for the **Shapiro–Wilk test** for normality.

The Shapiro–Wilk test compares the distribution of a variable with a normal distribution having the same mean and standard deviation. The null hypothesis of the Shapiro–Wilk test is that the sample is normally distributed, thus if `normtest.p` is lower than `0.01` (i.e., *p < .01*), the test indicates that the distribution is most probably not normal. The threshold to accept or reject a hypothesis is arbitrary and based on conventions, where *p < .01* is the most commonly accepted threshold, or *p < .05* for relatively small data sample (e.g., 30 cases).

The next step is thus to apply the `stat.desc` to the variable we are currently exploring (i.e., `perc_age_20_to_24`), including the `norm` section.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
leic_2011OAC_20to24_stat_desc <- leic_2011OAC_20to24 %>%
  select(perc_age_20_to_24) %>%
  stat.desc(norm = TRUE)
  
leic_2011OAC_20to24_stat_desc %>%
  kable(digits = 3)
```

The table above tells us that all `r leic_2011OAC_20to24_stat_desc["nbr.val", "perc_age_20_to_24"] %>% round(digits = 0)` OA in Leicester have a valid value for the variable `perc_age_20_to_24`, as no `NULL` nor `NA` value have been found. The values vary from about `r leic_2011OAC_20to24_stat_desc["min", "perc_age_20_to_24"] %>% round(digits = 0)`% to almost `r leic_2011OAC_20to24_stat_desc["max", "perc_age_20_to_24"] %>% round(digits = 0)`%, with an average value of `r leic_2011OAC_20to24_stat_desc["mean", "perc_age_20_to_24"] %>% round(digits = 0)`% of the population in an OA aged between 20 and 24. 

The short paragraph above is reporting on the values on the table, taking advantage of two features of RMarkdown. First, the output of the `stat.desc` function in the snippet further above is stored in the variable `leic_2011OAC_20to24_stat_desc`, which is then a valid variable for the rest of the document. Second, RMarkdown allows for in-line R snippets, that can also refer to variables defined in any snippet above the text. As such, the source of the paragraph above reads as below, with the in-line R snipped opened by a single grave accent (i.e., `` `r "\u0060"` ``) followed by a lowercase `` `r "r"` `` and closed by another single grave accent.

Having included all the code above into an RMarkdown document, copy the text below verbatim into the same RMarkdown document and make sure that you understand how the code in the in-line R snippets works.

```{r, eval=FALSE}
The table above tells us that all `r leic_2011OAC_20to24_stat_desc["nbr.val",
"perc_age_20_to_24"] %>% round(digits = 0)` OA in Leicester have a valid 
value for the variable `perc_age_20_to_24`, as no `r NULL` nor `r 
NA` value have been found.The values vary from about `r 
leic_2011OAC_20to24_stat_desc["min", "perc_age_20_to_24"] %>% round(digits = 
0)`% to almost `r leic_2011OAC_20to24_stat_desc["max", 
"perc_age_20_to_24"] %>% round(digits = 0)`%, with an average value of 
`r leic_2011OAC_20to24_stat_desc["mean", "perc_age_20_to_24"] %>% 
round(digits = 0)`% of the population in an OA aged between 20 and 24. 
```

If the data described by statistics presented in the table above was a random sample of a population, the 95% confidence interval `CI.mean.0.95` would indicate that we can be 95% confident that the actual mean of the distribution is somewhere between `r leic_2011OAC_20to24_stat_desc["mean", "perc_age_20_to_24"] %>% round(digits = 3)` `-` `r leic_2011OAC_20to24_stat_desc["CI.mean.0.95", "perc_age_20_to_24"] %>% round(digits = 3)` `=` `r (leic_2011OAC_20to24_stat_desc["mean", "perc_age_20_to_24"] - leic_2011OAC_20to24_stat_desc["CI.mean.0.95", "perc_age_20_to_24"]) %>% round(digits = 3)`% and `r leic_2011OAC_20to24_stat_desc["mean", "perc_age_20_to_24"] %>% round(digits = 3)` `+` `r leic_2011OAC_20to24_stat_desc["CI.mean.0.95", "perc_age_20_to_24"] %>% round(digits = 3)` `=` `r (leic_2011OAC_20to24_stat_desc["mean", "perc_age_20_to_24"] + leic_2011OAC_20to24_stat_desc["CI.mean.0.95", "perc_age_20_to_24"]) %>% round(digits = 3)`%. 

However, this is not a sample. Thus the statistical interpretation is not valid, in the same way that the `sum` values doesn't make sense, as it is the sum of a series of percentages.

Both `skew.2SE` and `kurt.2SE` are greater than `1`, which indicate that the `skewness` and `kurtosis` values are significant (*p < .05*). The `skewness` is positive, which indicates that the distribution is skewed towards the left (low values). The `kurtosis` is positive, which indicates that the distribution is heavy-tailed.

The [function `skim`](https://docs.ropensci.org/skimr/reference/skim.html) of the [library `skimr`](https://docs.ropensci.org/skimr/) can also be used to generate a quick summary of the content of a dataset.

```{r}
library(skimr)

leic_2011OAC_20to24 %>% 
  skim()
```


## Significance

Most statistical tests are based on the idea of hypothesis testing:

1. a **null hypothesis** is set;
2. the data are fit into a statistical model;
3. the model is assessed with a **test statistic**;
4. the **significance** is the probability of obtaining that test statistic value by chance.

The threshold to accept or reject a hypothesis is arbitrary and based on conventions. A 0.05 threshold (*p < .05*) is quite common with relatively small samples (e.g., dozens of cases), while a more strict 0.01 threshold (*p < .01*) is commonly advised for large samples (e.g., hundreds of cases). However, @10.2307/24700283 advise that such thresholds might not be suitable when working with big data (e.g., tens of thousands of cases).


## Shapiro–Wilk test

As `perc_age_20_to_24` is a heavy-tailed distribution, skewed towards low values, it is not surprising that the `normtest.p` value indicates that the Shapiro–Wilk test is significant, which indicates that the distribution is not normal.

The code below present the output of the [`shapiro.test` function](https://www.rdocumentation.org/packages/stats/versions/3.6.1/topics/shapiro.test), which only present the outcome of a Shapiro–Wilk test on the values provided as input. The output values are the same as the values reported by the `norm` section of `stat.desc`. Note that the `shapiro.test` function require the argument to be a numeric vector. Thus the [`pull` function](https://dplyr.tidyverse.org/reference/pull.html) must be used to extract the `perc_age_20_to_24` column from `leic_2011OAC_20to24` as a vector, whereas using `select` with a single column name as the argument would produce as output a table with a single column.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
leic_2011OAC_20to24 %>%
  pull(perc_age_20_to_24) %>%
  shapiro.test()
```

As the null hypothesis of the Shapiro–Wilk test is that the sample is normally distributed, a *p* value lower than a `0.01` threshold (*p < .01*) indicates that the probability of that being true is very low. So, the *flipper length* of  penguins in the Palmer Station dataset **is not** normally distributed.

The two code snippets below can be used to visualise a density-based histogram including the shape of a normal distribution having the same mean and standard deviation, and a Q-Q plot, to visually confirm the fact that `perc_age_20_to_24` is not normally distributed. 

```{r echo=TRUE, fig.height=3, fig.width=3, message=FALSE, warning=FALSE}
leic_2011OAC_20to24 %>%
  ggplot(
    aes(
      x = perc_age_20_to_24
    )
  ) +
  geom_histogram(
    aes(
      y =..density..
    ),
    binwidth = 5
  ) + 
  stat_function(
    fun = dnorm, 
    args = list(
      mean = leic_2011OAC_20to24 %>% pull(perc_age_20_to_24) %>% mean(),
      sd = leic_2011OAC_20to24 %>% pull(perc_age_20_to_24) %>% sd()
    ),
    colour = "red", size = 1
  )
```

A Q-Q plot in R can be created using a variety of functions. In the example below, the plot is created using the [`stat_qq` and `stat_qq_line` functions](https://ggplot2.tidyverse.org/reference/geom_qq.html) of the `ggplot2` library. Note that the `perc_age_20_to_24` variable is mapped to a particular option of `aes` that is `sample`. 

If `perc_age_20_to_24` had been normally distributed, the dots in the Q-Q plot would be distributed straight on the line included in the plot. 

```{r echo=TRUE, fig.height=3, fig.width=3, message=FALSE, warning=FALSE}
leic_2011OAC_20to24 %>%
  ggplot(
    aes(
      sample = perc_age_20_to_24
    )
  ) +
  stat_qq() +
  stat_qq_line()
```

## Exercise 202.1

Create a new RMarkdown document, and add the code necessary to recreate the table `leic_2011OAC_20to24` used in the example above. Use the code below to re-shape the table `leic_2011OAC_20to24` by pivoting the `perc_age_20_to_24` column wider into multiple columns using `supgrpname` as new column names. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
leic_2011OAC_20to24_supgrp <- leic_2011OAC_20to24 %>%
  pivot_wider(
    names_from = supgrpname,
    values_from = perc_age_20_to_24
  )
```

That manipulation creates one column per supergroup, containing the `perc_age_20_to_24` if the OA is part of that supergroup, or an `NA` value if the OA is not part of the supergroup. The transformation is illustrated in the two tables below. The first shows an extract from the original `leic_2011OAC_20to24` dataset, followed by the wide version `leic_2011OAC_20to24_supgrp`.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
leic_2011OAC_20to24 %>%
  slice_min(OA11CD, n = 10) %>%
  kable(digits = 3)

leic_2011OAC_20to24_supgrp %>%
  slice_min(OA11CD, n = 10) %>%
  kable(digits = 3)
```


**Question 202.1.1:** The code below uses the newly created `leic_2011OAC_20to24_supgrp` table to calculate the descriptive statistics calculated for the variable `leic_2011OAC_20to24` for each supergroup. Is `leic_2011OAC_20to24` normally distributed in any of the subgroups? If yes, which supergroups and based on which values do you justify that claim? (Write up to 200 words)

```{r, echo=TRUE, message=FALSE, warning=FALSE}
leic_2011OAC_20to24_supgrp %>%
  select(-OA11CD) %>%
  stat.desc(norm = TRUE) %>%
  kable(digits = 3)
```

**Question 202.1.2:** Write the code necessary to test again the normality of `leic_2011OAC_20to24` for the supergroups where the analysis conducted for Question 202.1.1 indicated they are normal, using the function `shapiro.test`, and draw the respective Q-Q plot.

**Question 202.1.3:** Observe the output of the Levene’s test executed below. What does the result tell you about the variance of `perc_age_20_to_24` in supergroups?

Note that the `leveneTest` was not designed to work with a Tidyverse approach. As such, the code below uses [the `.` argument placeholder](https://magrittr.tidyverse.org/#the-argument-placeholder) to specify that the input table `leic_2011OAC_20to24` which is coming down from the pipe should be used as argument for the `data` parameter.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
leic_2011OAC_20to24 %>% 
  leveneTest(
    perc_age_20_to_24 ~ supgrpname, 
    data = .
  )
```

---

<small>by [Stefano De Sabbata](https://sdesabbata.github.io/) -- text licensed under the [CC BY-SA 4.0](https://creativecommons.org/licenses/by-sa/4.0/), contains public sector information licensed under the [Open Government Licence v3.0](http://www.nationalarchives.gov.uk/doc/open-government-licence), code licensed under the [GNU GPL v3.0](https://www.gnu.org/licenses/gpl-3.0.html).</small>