# Appendix 1 {-}

## Basic types {-}

A value in `R` is an instance of one of three basic types, each encoding a fundamentally different type of information: `numeric` encoding numbers; `logical` encoding truth values (also known as Boolean values); and character encoding text. Each type has its characteristics and related operations, as discussed below. 

### Numeric {-}

The *numeric* type represents numbers (both integers and reals).

```{r, echo=TRUE}
a_number <- 1.41
is.numeric(a_number)
is.integer(a_number)
is.double(a_number) # i.e., is real
```

Base numeric operators.

|Operator|Meaning         |Example|Output   |
|--------|----------------|-------|---------|
|+       |Plus            |`5+2`  |`r 5+2`  |
|-       |Minus           |`5-2`  |`r 5-2`  |
|`*`     |Product         |`5*2`  |`r 5*2`  |
|/       |Division        |`5/2`  |`r 5/2`  |
|%/%     |Integer division|`5%/%2`|`r 5%/%2`|
|%%      |Module          |`5%%2` |`r 5%%2` |
|^       |Power           |`5^2`  |`r 5^2`  |


Some pre-defined functions in `R`:

```{r, echo=TRUE}
abs(-2) # Absolute value
ceiling(3.475) # Upper round
floor(3.475) # Lower round
trunc(5.99) # Truncate
log10(100) # Logarithm 10
log(exp(2)) # Natural logarithm and e
```

Use simple brackets to specify the order of execution. If not specified the default order is: rise to power first, then multiplication and division, sum and subtraction last.

```{r, echo=TRUE}
a_number <- 1
(a_number + 2) * 3
a_number + (2 * 3)
a_number + 2 * 3
```

The object `NaN` (*Not a Number*) is returned by `R` when the result of an operation is not a number.

```{r, echo=TRUE}
0 / 0
is.nan(0 / 0)
```

That is not to be confused with the object `NA` (*Not Available*), which is returned for missing data.

### Logical {-}

The *logical* type encodes two truth values: True and False.

```{r, echo=TRUE}
logical_var <- TRUE
is.logical(logical_var)
isTRUE(logical_var)
as.logical(0) # TRUE if not zero
```

`R` provides a series of basic logic operators you can use to evaluate *conditions*. For instance, you can use the logic operator `==` to evaluate the condition `5==2`, which tests whether the value `5` is equal to the value `2`. Conditions can be tested on values as well as on variables.

```{r, echo=TRUE}
5==2
first_value <- 5
second_value <- 2
first_value == 5
first_value == 2
second_value == 5
second_value == 2
first_value == second_value
```

|Operator|Meaning          |Example         |Output            |
|--------|-----------------|----------------|------------------|
|==      |Equal            |`5==2`          |`r 5==2`          |
|!=      |Not equal        |`5!=2`          |`r 5!=2`          |
|>       |Greater than     |`5>2`           |`r 5>2`           |
|<       |Less than        |`5<2`           |`r 5<2`           |
|>=      |Greater or equal |`5>=2`          |`r 5>=2`          |
|<=      |Less or equal    |`5<=2`          |`r 5<=2`          |
|!       |Not              |`!TRUE`         |`r !TRUE`         |
|&       |And              |`TRUE & FALSE`  |`r TRUE & FALSE`  |
|\|      |Or               |`TRUE | FALSE`  |`r TRUE | FALSE`  |



### Character {-}

The *character* type represents text objects, including single characters and character strings (that is text objects longer than one character, commonly referred to simply as *strings* in computer science).

```{r, echo=TRUE}
a_string <- "Hello world!"
is.character(a_string)
is.numeric(a_string)
as.character(2) # type conversion  (a.k.a. casting)
as.numeric("2")
as.numeric("Ciao")
```

### Types and variables {-}

A variable storing a value of a given type is said to have the same type. However, variables in `R` don't have an assigned type themselves. That means that a variable can be assigned a numeric value first and then changed to a character value. 

```{r, echo=TRUE}
a_variable <- 1.41
a_variable
is.numeric(a_variable)
a_variable <- "Hello world!"
a_variable
is.character(a_variable)
is.numeric(a_variable)
```

To be more precise, many programming languages require to *declare* a variable. That is, to state the type of a variable before it can be used. Variable declaration was particularly common in older programming languages such as `C` and `Java`. `R` does not require to declare variables types.



## More on vectors and factors {-}

### Vectors {-}

The operator `:` can be used to create integer vectors, starting from the number specified before the operator to the number specified after the operator. 

```{r, echo=TRUE}
# Create a vector containing integers between 2 and 4
two_to_four <- 2:4
two_to_four
# Retrieve cities between the second and the fourth
east_midlands_cities <- c("Derby", "Leicester", "Lincoln", "Nottingham")
east_midlands_cities[two_to_four]
# As the second element of two_to_four is 3...
two_to_four[2]
# the following command will retrieve the third city
east_midlands_cities[two_to_four[2]]
# Create a vector with cities from the previous vector
selected_cities <- c(east_midlands_cities[1], east_midlands_cities[3:4])
```


The functions `seq` and `rep` can also be used to create vectors, as illustrated below.

```{r, echo=TRUE}
seq(1, 10, by = 0.5)
seq(1, 10, length.out = 6)
rep("Ciao", 4)
```


The logical operators `any` and `all` can be used to test conditions on the vector. The former returns `TRUE` if at least one element satisfies the statement and the second returns `TRUE` if all elements satisfy the condition

```{r, echo=TRUE}
any(east_midlands_cities == "Leicester")
my_sequence <- seq(1, 10, length.out = 7)
my_sequence
any(my_sequence > 5)
all(my_sequence > 5)
```



### Factors {-}

A **factor** is a data type similar to a vector. However, the values contained in a factor can only be selected from a set of **levels**.

```{r, echo=TRUE}
houses_vector <- c("Bungalow", "Flat", "Flat",
  "Detached", "Flat", "Terrace", "Terrace")
houses_vector
houses_factor <- factor(c("Bungalow", "Flat", "Flat",
  "Detached", "Flat", "Terrace", "Terrace"))
houses_factor
```

The function **table** can be used to obtain a tabulated count for each level.

```{r, echo=TRUE}
houses_factor <- factor(c("Bungalow", "Flat", "Flat",
  "Detached", "Flat", "Terrace", "Terrace"))
houses_factor
table(houses_factor)
```

A specific set of levels can be specified when creating a factor by providing a **levels** argument.

```{r, echo=TRUE}
houses_factor_spec <- factor(
  c("People Carrier", "Flat", "Flat", "Hatchback",
      "Flat", "Terrace", "Terrace"),
  levels = c("Bungalow", "Flat", "Detached",
       "Semi", "Terrace"))

table(houses_factor_spec)
```

In statistics terminology, (unordered) factors are **categorical** (i.e., binary or nominal) variables. Levels are not ordered.

```{r, echo=TRUE}
income_nominal <- factor(
  c("High", "High", "Low", "Low", "Low",
      "Medium", "Low", "Medium"),
  levels = c("Low", "Medium", "High"))
```

The *greater than* operator is not meaningful on the `income_nominal` factor defined above.

```{r, echo=TRUE}
income_nominal > "Low"
```

In statistics terminology, ordered factors are **ordinal** variables. Levels are ordered.

```{r, echo=TRUE}
income_ordered <- ordered(
  c("High", "High", "Low", "Low", "Low",
      "Medium", "Low", "Medium"),
  levels = c("Low", "Medium", "High"))

income_ordered > "Low"

sort(income_ordered)
```




---

<small>by [Stefano De Sabbata](https://sdesabbata.github.io/) -- text licensed under the [CC BY-SA 4.0](https://creativecommons.org/licenses/by-sa/4.0/), contains public sector information licensed under the [Open Government Licence v3.0](http://www.nationalarchives.gov.uk/doc/open-government-licence), code licensed under the [GNU GPL v3.0](https://www.gnu.org/licenses/gpl-3.0.html).</small>
