# Solutions {-}

<br/><small><a href="javascript:if(window.print)window.print()">Print this chapter</a></small>

## Solutions 101 {-}

Note that you can add sections to your code, such as the "Intro" section below by selecting Code > Insert Section... from the top menu in RStudio. Those will then be visible on a right panel in  the RStudio integrated editor if you select Code > Show Document Outline from the top menu. Code sections can help to visually organise your code and the Document Outline allows you to select a section and jump to it in the code.

The code below deletes all variable and function currently defined in the environment, which is generally good practice, to avoid your script interacting with objects left over from previous sessions.

```{r, echo=TRUE, eval=FALSE}
rm(list = ls())
```

Load the necessary libraries.

```{r, echo=TRUE, eval=FALSE}
# Load the tidyverse
library(tidyverse)

# Load magrittr
# necessary for options 3 and 4 
# of my answer to Question 101.2.4
library(magrittr)
```



### Solutions 101.1 {-}

**Question 101.1.1:** Write a piece of code using the pipe operator that takes as input the number `1632`, calculates the logarithm to the base 10, takes the highest integer number lower than the calculated value (lower round), and verifies whether it is an integer.

```{r, echo=TRUE, eval=FALSE}
1632 %>%
  # calculate the logarithm to the base 10
  log10() %>%
  # highest integer number lower than the value
  floor() %>%
  # check whether it is an integer
  is.integer()
  # The answer is FALSE
  # as the value is still of type numeric
  # rather than type integer
```

```{r, echo=TRUE, eval=FALSE}
# The code below replicates the procedure above
# but checking the value data type
1632 %>%
  log10() %>%
  floor() %>%
  class()
```

```{r, echo=TRUE, eval=FALSE}
# The code below replicates the procedure above
# but using as.integer to transfor the value
# to an integer type
1632 %>%
  log10() %>%
  floor() %>%
  as.integer() %>%
  class()
```

```{r, echo=TRUE, eval=FALSE}
# As above but checking whether the value is an integer
1632 %>%
  log10() %>%
  floor() %>%
  as.integer() %>%
  is.integer()
```

**Question 101.1.2:** Write a piece of code using the pipe operator that takes as input the number `1632`, calculates the square root, takes the lowest integer number higher than the calculated value (higher round), and verifies whether it is an integer.

```{r, echo=TRUE, eval=FALSE}
1632 %>%
  # calculate the square root
  sqrt() %>%
  # lowest integer number higher than the value
  ceiling() %>%
  # check whether it is an integer
  is.integer()
```

**Question 101.1.3:** Write a piece of code using the pipe operator that takes as input the string `"1632"`, transforms it into a number, and checks whether the result is *Not a Number*.

```{r, echo=TRUE, eval=FALSE}
"1632" %>%
  # transform it into a number
  as.numeric() %>%
  # check whether the result is Not a Number
  is.nan()
```

**Question 101.1.4:** Write a piece of code using the pipe operator that takes as input the string `"-16.32"`, transforms it into a number, takes the absolute value and truncates it, and finally checks whether the result is *Not Available*.

```{r, echo=TRUE, eval=FALSE}
"-16.32" %>%
  # transform it into a number
  as.numeric() %>%
  # take the absolute value
  abs() %>%
  # truncate
  trunc() %>%
  # check whether the result is Not Available
  is.na()
```

**Question 101.1.5:** Rewrite a piece of code below by substituting the last line with the function `mean()`. What kind of result do you obtain? What does it represent?

```{r, echo=TRUE, eval=FALSE, fig.height=3}
iris %>% 
  pull(Petal.Length) %>% 
  mean()
```

The output is a numeric value representing the aritmetic mean (average) of all the petal lengths of flowers in the `iris` dataset.

**Question 101.1.6:** Further edit the code created for *Question 101.1.6* by substituting `Petal.Length` with `Petal.Width` first and `Species` then? What kind of results do you obtain? What do they mean?


```{r, echo=TRUE, eval=FALSE, fig.height=3}
iris %>% 
  pull(Petal.Width) %>% 
  mean()

iris %>% 
  pull(Species) %>% 
  mean()
```

## Solutions 102 {-}

### Solutions 102.1 {-}

Extend the code in the script `practical-102_my-script-002.R` to include the code necessary to solve the questions below.

```{r, echo=TRUE, warning=FALSE, message=FALSE}
library(tidyverse)
library(nycflights13)
library(knitr)
```


**Question 101.1.1:** Write a piece of code using the pipe operator and the `dplyr` library to generate a table showing the air time and the carrier, but only for flights starting from the JFK airport. As in the examples above, use `slice_head` and `kable` to output a nicely-formatted table containing only the first 10 rows.

```{r, echo=TRUE}
# Start from the entire dataset
flights %>%
  # Retain only the necessary columns
  select(origin, carrier, air_time) %>%
  # Retain only rows representing flights from JFK
  filter(origin == "JFK") %>%
  slice_head(n = 5) %>%
  kable()
```

**Question 102.1.2:** Write a piece of code using the pipe operator and the `dplyr` library to generate a table showing the arrival delay and the overall air time, but only for flights of October 12th. As in the examples above, use `slice_head` and `kable` to output a nicely-formatted table containing only the first 10 rows.

```{r, echo=TRUE}
# Start from the entire dataset
flights %>%
  # Retain only the necessary columns
  select(year:day, arr_delay, air_time) %>%
  # Retain only rows where month equals 10 and day equals 12
  filter(month == 10 & day == 12) %>%
  slice_head(n = 5) %>%
  kable()
```

**Question 102.1.3:** Write a piece of code using the pipe operator and the `dplyr` library to generate a table showing the arrival delay, origin and destination, but only for flight leaving between 11am and 2pm. As in the examples above, use `slice_head` and `kable` to output a nicely-formatted table containing only the first 10 rows.

```{r, echo=TRUE}
# Start from the entire dataset
flights %>%
  # Retain only the necessary columns
  select(origin, dest, dep_time, arr_delay) %>%
  # Retain only rows where departure time is greater than or equal to 1100
  # and departure time is also less than or equal to 1100
  filter(dep_time >= 1100 & dep_time <= 1400) %>%
  slice_head(n = 5) %>%
  kable()
```


## Solutions 103 {-}

### Solutions 103.1 {-}

Create an RMarkdown document in RStudio, using *Exercise 103* as title and *PDF* as output. Delete all the contents except the first five lines which compose the heading. Save the document as `practical-103_exercises.Rmd`. Add the libraries and code necessary to read the data from the `2011_OAC_Raw_uVariables_Leicester.csv` file. Create a first section of the document (e.g., adding a second heading *Exercise 103.1*) and add your answers to the questions below.

In order to answer the questions below, inspect the **look-up** table `2011_OAC_Raw_uVariables_Lookup.csv` (e.g., using Microsoft Excel) to identify the columns necessary to complete the task.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)

leicester_2011OAC <- 
  read_csv(
    paste0(
      Sys.getenv("RGDS_HOME"), 
      "/data/", 
      "2011_OAC_Raw_uVariables_Leicester.csv"
    )
  )
```


**Question 103.1.1:** Identify the five variables which are part of the variable subdomain *Housing Type* and write the code necessary to compute the total number of household spaces in Leicester for each housing type.

```{r}
leicester_2011OAC %>% 
  summarise(
    u086_tot = sum(u086),
    u087_tot = sum(u087),
    u088_tot = sum(u088),
    u089_tot = sum(u089),
    u090_tot = sum(u090)
  ) %>% 
  kable()
```

**Question 103.1.2:** Write the code necessary to compute the total number of household spaces in Leicester for each housing type grouped by 2011 OAC supergroup.

```{r}
leicester_2011OAC %>% 
  group_by(supgrpname) %>% 
  summarise(
    u086_tot = sum(u086),
    u087_tot = sum(u087),
    u088_tot = sum(u088),
    u089_tot = sum(u089),
    u090_tot = sum(u090)
  ) %>% 
  kable()
```

**Question 103.1.3:** Write the code necessary to compute the percentage of household spaces (i.e., over to the total number of household spaces) in Leicester for each housing type grouped by 2011 OAC supergroup.

```{r}
leicester_2011OAC %>% 
  group_by(supgrpname) %>% 
  summarise(
    u086_tot = sum(u086),
    u087_tot = sum(u087),
    u088_tot = sum(u088),
    u089_tot = sum(u089),
    u090_tot = sum(u090)
  ) %>% 
  mutate(
    tot_hspaces = 
      u086_tot + u087_tot +
      u088_tot + u089_tot +
      u090_tot
  ) %>% 
  mutate(
    u086_perc = (u086_tot / tot_hspaces) * 100,
    u087_perc = (u087_tot / tot_hspaces) * 100,
    u088_perc = (u088_tot / tot_hspaces) * 100,
    u089_perc = (u089_tot / tot_hspaces) * 100,
    u090_perc = (u090_tot / tot_hspaces) * 100
  ) %>% 
  select(
    supgrpname,
    u086_perc, u087_perc, u088_perc,
    u089_perc, u090_perc
  ) %>% 
  kable(digits = c(0, 2, 2, 2, 2, 2))
```

**Question 103.1.4:** Modify the code written for *Question 103.1.3*, using [the verb `rename`]() to change the column names of the columns containing the percentages to names that resemble the related housing type (e.g., `perc_of_detached`).


## Solutions 104 {-}

### Solutions 104.1 {-}

Extend the code in the script `Data_Wrangling_Example.R` (see code below) to include the code necessary to solve the questions below. Use the full list of variable names from the 2011 UK Census used to generate the 2011 OAC thatcan be found in the file `2011_OAC_Raw_uVariables_Lookup.csv` to indetify which columns to use to complete the tasks. 

```{r, echo=TRUE, eval=FALSE}
# Data_Wrangling_Example.R 

# Load the tidyverse
library(tidyverse)

# Load 2011 OAC data
leicester_2011OAC <- 
  read_csv("data/2011_OAC_Raw_uVariables_Leicester.csv")

# Load Indexes of Multiple deprivation data
leicester_IMD2015 <- 
  read_csv("data/IndexesMultipleDeprivation2015_Leicester.csv")

leicester_IMD2015_decile_wide <- leicester_IMD2015 %>%
  # Select only Socres
  filter(Measurement == "Decile") %>%
  # Trim names of IndicesOfDeprivation
  mutate(
    IndicesOfDeprivation = str_replace_all(IndicesOfDeprivation, "\\s", "")
  ) %>%
  mutate(
    IndicesOfDeprivation = str_replace_all(IndicesOfDeprivation, "[:punct:]", "")
  ) %>%
  mutate(
    IndicesOfDeprivation = str_replace_all(IndicesOfDeprivation, "\\(", "")
  ) %>%
  mutate(
    IndicesOfDeprivation = str_replace_all(IndicesOfDeprivation, "\\)", "")
  ) %>%
  # Spread
  pivot_wider(
    names_from = IndicesOfDeprivation,
    values_from = Value
  ) %>%
  # Drop columns
  select(-DateCode, -Measurement, -Units)

# Join
leicester_2011OAC_IMD2015 <- 
  leicester_2011OAC %>%
  inner_join(
    leicester_IMD2015_decile_wide, 
    by = c("LSOA11CD" = "FeatureCode")
  )
```

**Question 104.1.1:** Write a piece of code using the pipe operator and the `dplyr` library to generate a table showing the percentage of EU citizens over total population, calculated grouping OAs by the related decile of the Index of Multiple Deprivations, but only accounting for areas classified as Cosmopolitans or Ethnicity Central or Multicultural Metropolitans.

```{r, echo=TRUE, eval=FALSE}
leicester_2011OAC_IMD2015 %>%
  filter(supgrpname %in% c("Cosmopolitans", "Ethnicity Central", "Multicultural Metropolitans")) %>%
  group_by(IndexofMultipleDeprivationIMD) %>%
  summarise(
    adults_not_empl_perc = (sum(u044 + u045) / sum(Total_Population)) * 100
  ) %>%
  kable()
```

**Question 104.1.2:** Write a piece of code using the pipe operator and the `dplyr` library to generate a table showing the percentage of EU citizens over total population, calculated grouping OAs by the related supergroup in the 2011 OAC, but only accounting for areas in the top 5 deciles of the Index of Multiple Deprivations.

```{r, echo=TRUE, eval=FALSE}
leicester_2011OAC_IMD2015 %>%
  filter(IndexofMultipleDeprivationIMD <= 5) %>%
  group_by(supgrpname) %>%
  summarise(
    eu_perc = (sum(u043 + u044 + u045) / sum(Total_Population)) * 100
  ) %>%
  kable()
```


**Question 104.1.3:** Write a piece of code using the pipe operator and the `dplyr` library to generate a table showing the percentage of people aged 65 and above, calculated grouping OAs by the related supergroup in the 2011 OAC and decile of the Index of Multiple Deprivations, and ordering the table by the calculated value in a descending order.

```{r, echo=TRUE, eval=FALSE}
leicester_2011OAC_IMD2015 %>%
  filter(IndexofMultipleDeprivationIMD <= 5) %>%
  group_by(supgrpname, IndexofMultipleDeprivationIMD) %>%
  summarise(
    aged_65_above = (sum(u016 + u017 + u018 + u019) / sum(Total_Population)) * 100
  ) %>%
  arrange(-aged_65_above) %>%
  kable()
```

Extend the code in the script `Data_Wrangling_Example.R` to include the code necessary to solve the questions below.

**Question 104.1.4:** Write a piece of code using the pipe operator and the `dplyr` and `tidyr` libraries to generate a long format of the `leicester_2011OAC_IMD2015` table only including the values (census variables) used in *Question 104.1.3*.

```{r, echo=TRUE, eval=FALSE}
long_table <- leicester_2011OAC_IMD2015 %>%
  select(OA11CD, supgrpname, IndexofMultipleDeprivationIMD, u016, u017, u018, u019, Total_Population) %>%
  pivot_longer(
    # Can't combine character values (e.g. supgrpname)
    # with numeric value (e.g, Total_Population) thus
    # pivot only numeric columns
    cols = u016:Total_Population,
    names_to = "attribute",
    values_to = "value"
  ) 

long_table %>%
  slice_head(n = 5) %>%
  kable()

long_table_alt <- leicester_2011OAC_IMD2015 %>%
  select(OA11CD, supgrpcode, IndexofMultipleDeprivationIMD, u016, u017, u018, u019, Total_Population) %>%
  pivot_longer(
    # Otherwise, use supgrpcode instead of supgrpname
    cols = -OA11CD,
    names_to = "attribute",
    values_to = "value"
  ) 

long_table_alt %>%
  slice_head(n = 7) %>%
  kable()
```


**Question 104.1.5:** Write a piece of code using the pipe operator and the `dplyr` and `tidyr` libraries to generate a table similar to the one generated for *Question 104.1.4*, but showing the values as percentages over total population.

```{r, echo=TRUE, eval=FALSE}
perc_long_table <- leicester_2011OAC_IMD2015 %>%
  select(OA11CD, supgrpname, IndexofMultipleDeprivationIMD, u016, u017, u018, u019, Total_Population) %>%
  mutate(
    perc_u016 = (u016 / Total_Population) * 100, 
    perc_u017 = (u017 / Total_Population) * 100, 
    perc_u018 = (u018 / Total_Population) * 100, 
    perc_u019 = (u019 / Total_Population) * 100
  ) %>%
  select(OA11CD, supgrpname, IndexofMultipleDeprivationIMD, perc_u016, perc_u017, perc_u018, perc_u019) %>%
  pivot_longer(
    # Can't combine character values (e.g. supgrpname)
    # with numeric value (e.g, Total_Population) thus
    # pivot only numeric columns
    cols = perc_u016:perc_u019,
    names_to = "attribute",
    values_to = "value"
  ) 

perc_long_table %>%
  slice_head(n = 5) %>%
  kable()
```



## Solutions 201 {-}

### Solutions 201.1 {-}

Open the `Leicester_population` project used in previous chapters and extend the *"Exploring deprivation indices in Leicester"* document to include the code necessary to solve the questions below. Use the full list of variable names from the 2011 UK Census used to generate the 2011 OAC that can be found in the file `2011_OAC_Raw_uVariables_Lookup.csv` to indetify which columns to use to complete the tasks. 

**Question 201.1.1:** Write a piece of code to create a chart showing the percentage of EU citizens over total population for each decile of the Index of Multiple Deprivations

```{r, echo=TRUE, eval=FALSE}
leicester_2011OAC_IMD2015 %>% 
  group_by(IndexofMultipleDeprivationIMD) %>%
  summarise(
    eu_perc = (sum(u043 + u044 + u045) / sum(Total_Population)) * 100
  ) %>%
  ggplot(
    aes(
      x = eu_perc,
      # Note that it is necessary to convert the index
      # to a factor, otherwise it is interpreted as a number
      y = as_factor(IndexofMultipleDeprivationIMD)
    )
  ) +
  geom_col() +
  theme_bw()
```

Alternatively.

```{r, echo=TRUE, eval=FALSE}
leicester_2011OAC_IMD2015 %>% 
  group_by(IndexofMultipleDeprivationIMD) %>%
  summarise(
    eu_perc = (sum(u043 + u044 + u045) / sum(Total_Population)) * 100
  ) %>%
  mutate(
    reminder = 100 - eu_perc
  ) %>%
  pivot_longer(
    cols = -IndexofMultipleDeprivationIMD,
    names_to = "country_of_origin",
    values_to = "percentage"
  ) %>% 
  ggplot(
    aes(
      x = percentage,
      # Note that it is necessary to convert the index
      # to a factor, otherwise it is interpreted as a number
      y = as_factor(IndexofMultipleDeprivationIMD),
      fill = country_of_origin
    )
  ) +
  geom_col() +
  theme_bw()
```

**Question 201.1.2:** Write a piece of code to create a chart showing the relationship between the percentage of EU citizens over total population with the related score of the Index of Multiple Deprivations, and illustrating also the 2011 OAC class of each OA.

```{r, echo=TRUE, eval=FALSE}
# Use a similar code as used in Exercise 104
# but filtering in the Scores rather than the deciles
leicester_IMD2015_score_wide <- leicester_IMD2015 %>%
  # Select only Socres
  filter(Measurement == "Score") %>%
  # Trim names of IndicesOfDeprivation
  mutate(
    IndicesOfDeprivation = str_replace_all(IndicesOfDeprivation, "\\s", "")
  ) %>%
  mutate(
    IndicesOfDeprivation = str_replace_all(IndicesOfDeprivation, "[:punct:]", "")
  ) %>%
  mutate(
    IndicesOfDeprivation = str_replace_all(IndicesOfDeprivation, "\\(", "")
  ) %>%
  mutate(
    IndicesOfDeprivation = str_replace_all(IndicesOfDeprivation, "\\)", "")
  ) %>%
  # Spread
  pivot_wider(
    names_from = IndicesOfDeprivation,
    values_from = Value
  ) %>%
  # Drop columns
  select(-DateCode, -Measurement, -Units)

# Join
leicester_2011OAC_IMD2015_score <- 
  leicester_2011OAC %>%
  inner_join(
    leicester_IMD2015_score_wide, 
    by = c("LSOA11CD" = "FeatureCode")
  )

leicester_2011OAC_IMD2015_score %>% 
  mutate(
    eu_perc = ((u043 + u044 + u045) / Total_Population) * 100
  ) %>%
  ggplot(
    aes(
      x = eu_perc,
      y = IndexofMultipleDeprivationIMD,
      colour = fct_reorder(supgrpname, supgrpcode)
    )
  ) +
  geom_point() +
  scale_fill_manual(
    values = c("#e41a1c", "#f781bf", "#ff7f00", "#a65628", "#984ea3", "#377eb8", "#ffff33")
  ) +
  theme_bw()
```

**Question 201.1.3:** Write a piece of code to create a chart showing the relationship between the percentage of people aged 65 and above with the related score of the Income Deprivation, and illustrating also the 2011 OAC class of each OA.

```{r, echo=TRUE, eval=FALSE}
leicester_2011OAC_IMD2015_score %>% 
  mutate(
    aged_65_above = ((u016 + u017 + u018 + u019) / Total_Population) * 100
  ) %>%
  ggplot(
    aes(
      x = aged_65_above,
      y = LivingEnvironmentDeprivationDomain,
      colour = fct_reorder(supgrpname, supgrpcode)
    )
  ) +
  geom_point() +
  scale_fill_manual(
    values = c("#e41a1c", "#f781bf", "#ff7f00", "#a65628", "#984ea3", "#377eb8", "#ffff33")
  ) +
  theme_bw()
```


**Question 201.1.4:** What does the graph produced for *Question 201.1.3* mean? Write up to 100 words explaining what conclusions can be drawn from the graph -- remember that [*"the larger the score, the more deprived the area"*](https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/853811/IoD2019_FAQ_v4.pdf).

**Question 201.1.5:** Identify the index of multiple deprivation that most closely relate to the percentage of people per OA whose *"day-to-day activities limited a lot or a little"* based on the *"Standardised Illness Ratio"*.



## Solutions 202 {-}

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(pastecs)
library(car)

leicester_2011OAC <- read_csv(paste0(Sys.getenv("RGDS_HOME"), "/data/", "2011_OAC_Raw_uVariables_Leicester.csv"))

leic_2011OAC_20to24 <- leicester_2011OAC %>%
  mutate(
    perc_age_20_to_24 = (u011 / Total_Population) * 100,
    supgrpname = 
      dplyr::recode(
        supgrpname, 
          `Suburbanites` = "SU",
          `Cosmopolitans` = "CP",
          `Multicultural Metropolitans` = "MM",
          `Ethnicity Central` = "EC",
          `Constrained City Dwellers` = "CD",
          `Hard-Pressed Living` = "HP",
          `Urbanites` = "UR"
    )
  ) %>%
  select(OA11CD, supgrpname, perc_age_20_to_24)
```

### Solutions 202.1 {-}

Create a new RMarkdown document, and add the code necessary to recreate the table `leic_2011OAC_20to24` used in the example above. Use the code below to re-shape the table `leic_2011OAC_20to24` by pivoting the `perc_age_20_to_24` column wider into multiple columns using `supgrpname` as new column names. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
leic_2011OAC_20to24_supgrp <- leic_2011OAC_20to24 %>%
  pivot_wider(
    names_from = supgrpname,
    values_from = perc_age_20_to_24
  )
```

That manipulation creates one column per supergroup, containing the `perc_age_20_to_24` if the OA is part of that supergroup, or an `NA` value if the OA is not part of the supergroup. The transformation is illustrated in the two tables below. The first shows an extract from the original `leic_2011OAC_20to24` dataset, followed by the wide version `leic_2011OAC_20to24_supgrp`.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
leic_2011OAC_20to24 %>%
  slice_min(OA11CD, n = 10) %>%
  kable(digits = 3)

leic_2011OAC_20to24_supgrp %>%
  slice_min(OA11CD, n = 10) %>%
  kable(digits = 3)
```


**Question 202.1.1:** The code below uses the newly created `leic_2011OAC_20to24_supgrp` table to calculate the descriptive statistics calculated for the variable `leic_2011OAC_20to24` for each supergroup. Is `leic_2011OAC_20to24` normally distributed in any of the subgroups? If yes, which supergroups and based on which values do you justify that claim? (Write up to 200 words)

```{r, echo=TRUE, message=FALSE, warning=FALSE}
leic_2011OAC_20to24_supgrp %>%
  select(-OA11CD) %>%
  stat.desc(norm = TRUE) %>%
  kable(digits = 3)
```

We can set a *p < .01* threshold, which is reasonable for the number of cases in the dataset (hundreds, at least for some of the 2011OAC supergroups). We can claim that `leic_2011OAC_20to24` is normally distributed in the supergroups Suburbanites (SU),	Cosmopolitans (CP), Constrained City Dwellers (CD) and  Hard-Pressed Living (HP), as the `normtest.p` value is above `0.01`, which allows us to reject the null hypothesis. The variable `leic_2011OAC_20to24` seems instead not to be normally distributed for Multicultural Metropolitans (MM), Ethnicity Central (EC) and Urbanites (UR). That is also further illustrated by the graphs below.

```{r echo=TRUE, fig.height=5, fig.width=5, message=FALSE, warning=FALSE}
leic_2011OAC_20to24 %>% 
  filter(supgrpname %in% c("SU", "CP", "CD", "HP")) %>% 
  ggplot(
    aes(
      x = perc_age_20_to_24
    )
  ) + 
  geom_histogram(
    aes(
      y = ..density..
    )
  ) +
  facet_wrap(
    vars(supgrpname),
    ncol = 2,
    scales = "free"
  ) +
  ggtitle("Normally distributed") +
  theme_bw()
```

```{r echo=TRUE, fig.height=5, fig.width=5, message=FALSE, warning=FALSE}
leic_2011OAC_20to24 %>% 
  filter(supgrpname %in% c("MM", "EC", "UR")) %>% 
  ggplot(
    aes(
      x = perc_age_20_to_24
    )
  ) + 
  geom_histogram(
    aes(
      y = ..density..
    )
  ) +
  facet_wrap(
    vars(supgrpname),
    ncol = 2,
    scales = "free"
  ) +
  ggtitle("Not normally distributed") +
  theme_bw()
```


**Question 202.1.2:** Write the code necessary to test again the normality of `leic_2011OAC_20to24` for the supergroups where the analysis conducted for Question 202.1.1 indicated they are normal, using the function `shapiro.test`, and draw the respective Q-Q plot.

Example for Hard-Pressed Living (HP).

```{r}
leic_2011OAC_20to24 %>%
  filter(supgrpname == "HP") %>% 
  pull(perc_age_20_to_24) %>%
  shapiro.test()
```

```{r}
leic_2011OAC_20to24 %>%
  filter(supgrpname == "HP") %>% 
  ggplot(
    aes(
      sample = perc_age_20_to_24
    )
  ) +
  stat_qq() +
  stat_qq_line()
```



**Question 202.1.3:** Observe the output of the Levene’s test executed below. What does the result tell you about the variance of `perc_age_20_to_24` in supergroups?

Note that the `leveneTest` was not designed to work with a Tidyverse approach. As such, the code below uses [the `.` argument placeholder](https://magrittr.tidyverse.org/#the-argument-placeholder) to specify that the input table `leic_2011OAC_20to24` which is coming down from the pipe should be used as argument for the `data` parameter.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
leic_2011OAC_20to24 %>% 
  leveneTest(
    perc_age_20_to_24 ~ supgrpname, 
    data = .
  )
```

If significant, the Levene’s test indicates that the variance is different in different levels. As before, we can set a *p < .01* threshold. In the output above, the *p* value (`Pr(>F)`) is much lower than our threshold, indicating that the test is significance. Thus, the `perc_age_20_to_24` has different variance for different 2011OAc supergroups.


## Solutions 203

### Solutions 203.1

Create a new RMarkdown document, and add the code necessary to oad the `2011_OAC_Raw_uVariables_Leicester.csv` dataset.

```{r, echo=FALSE, message=FALSE, warning=FALSE,}
leicester_2011OAC <- read_csv(paste0(Sys.getenv("RGDS_HOME"), "/data/", "2011_OAC_Raw_uVariables_Leicester.csv"))
```

**Question 203.1.1:** Check whether the values of mean age (`u020`) are normally distributed, and whether they can be transformed to a normally distributed set using logarithmic or inverse hyperbolic sine functions.

**Question 203.1.2:** Check whether the values of mean age (`u020`) are normally distributed when looking at the different 2011OAC supergroups separately. Check whether they can be transformed to a normally distributed set using logarithmic or inverse hyperbolic sine functions.

**Question 203.1.3:** Is the distribution of mean age (`u020`) different in different 2011OAC supergroups in Leicester?

### Solutions 203.2

**Question 203.2.1:** As mentioned above, when discussing movement in cities, there is an assumption that people living in the city centre live in flats and work or cycle to work, whereas people living in the suburbs live in whole houses and commute via car. Study the correlation between the presence of flats (`u089`) and people commuting to work on foot, bicycle or other similar means (`u122`) in the same OAs. Consider whether the values might need to be normalised or otherwised transformed before starting the testing procedure. 

**Question 203.2.2:** Another interesting issue to explore is the relationship between car ownership and the use of public transport. Study the correlation between the presence of households owning 2 or more cars or vans (`u118`) and people commuting to work via public transport (`u120`) or on foot, bicycle or other similar means (`u122`) in the same OAs. Consider whether the values might need to be normalised or otherwised transformed before starting the testing procedure. 


<!--

### Solutions 104.2 {-}

Answer the question below, consulting the [`stringr` library reference (stringr.tidyverse.org/reference)](https://stringr.tidyverse.org/reference/index.html) as necessary

**Question 104.2.1:** Write a piece of code using the pipe operator and the `stringr` library that takes as input the string `"I like programming in R"`, and transforms it all in uppercase.

```{r, echo=TRUE, eval=FALSE}
"I like programming in R" %>%
  # transforms it all in uppercase
  str_to_upper()
```


**Question 104.2.2:** Write a piece of code using the pipe operator and the `stringr` library that takes as input the string `"I like programming in R"`, and truncates it, leaving only 10 characters.

```{r, echo=TRUE, eval=FALSE}
"I like programming in R" %>%
  # truncate leaving only 10 characters
  str_trunc(10)
```


**Question 104.2.3:** Write a piece of code using the pipe operator and the `stringr` library that takes as input the string `"I like programming in R"`, and truncates it, leaving only 10 characters and using no ellipsis.

```{r, echo=TRUE, eval=FALSE}
"I like programming in R" %>%
  # truncate
  str_trunc(
    # leaving only 10 characters
    10,
    # no ellipsis
    ellipsis = ""
  )
```


**Question 104.2.4:** Write a piece of code using the pipe operator and the `stringr` library that takes as input the string `"I like programming in R"`, and manipulates to leave only the string `"I like R"`.

```{r, echo=TRUE, eval=FALSE}
# Option 1
"I like programming in R" %>%
  # using remove
  str_remove("programming in ")

# Option 2
"I like programming in R" %>%
  # using replace
  str_replace("programming in ", "")

# Option 3
"I like programming in R" %>%
  # create two sub-strings
  str_sub(
    # the first starting from the first character
    # the second starting from the last character
    start = c(1, -1),
    # the first ending at the sixth character
    # the secondending at the last character
    end = c(6, -1)
  ) %>% 
  # join the two strings into one
  # using a single space to separate them
  str_c(collapse = " ")

# Option 4
"I like programming in R" %>%
  # split the string at each space
  str_split(" ") %>% 
  # that will create a list with one element
  # that is the vector of strings resulting
  # from the split
  # to extract that element
  # use extract2 from magrittr
  # equivalent to [[1]]
  # https://magrittr.tidyverse.org/reference/aliases.html
  extract2(1) %>%
  # extract the first, second and fifth elements
  # from the vector
  # use extract2 from magrittr
  # equivalent to [c(1, 2, 5)]
  # https://magrittr.tidyverse.org/reference/aliases.html
  extract(c(1, 2, 5)) %>%
  # join the two strings into one
  # using a single space to separate them
  str_c(collapse = " ")
```


## Solutions 114 {-}

```{r, echo=TRUE, eval=FALSE}
# Load the tidyverse
library(tidyverse)
```



### Solutions 114.1 {-}

**Question 114.1.1:** Use the modulo operator `%%` to create a conditional statement that prints `"Even"` if a number is even and `"Odd"` if a number is odd.

```{r, echo=TRUE, eval=FALSE}
number_for_test <- 7

# If a number modulo 2 is 0 then the nuber is even
if (number_for_test %% 2 == 0) {
  cat("Even\n")
# otherwise it is odd
} else {
  cat("Odd\n")
}
```


**Question 114.1.2:** Encapsulate the conditional statement written for *Question 114.1.1* into a `for` loop that executes the conditional statement for all numbers from 1 to 10.

```{r, echo=TRUE, eval=FALSE}
for (iterator in 1:10) {
  if (iterator %% 2 == 0) {
    cat("Even\n")
  } else {
    cat("Odd\n")
  }
  
}
# After the loop delete the iterator
# just to be on the safe side :)
rm(iterator)
```


**Question 114.1.3:** Encapsulate the conditional statement written for *Question 114.1.1* into a `for` loop that prints the name of cities in odd positions (i.e., first, third, fifth) in the vector `c("Birmingham", "Derby", "Leicester", "Lincoln", "Nottingham", "Wolverhampton")`.

```{r, echo=TRUE, eval=FALSE}
cities <- c("Birmingham", "Derby", "Leicester", "Lincoln", "Nottingham", "Wolverhampton")

# for each index in the vecotr cities
# i.e., 1 to length of the vector
for (iterator in 1:length(cities)) {
  # Testing for odd number intead of even
  # thus checking if the result is 1 rather than 0
  if (iterator %% 2 == 1) {
    cat(cities[iterator], "\n")
  }
  
}
# After the loop delete the iterator
# just to be on the safe side :)
rm(iterator)
```

**Question 114.1.4:** Write the code necessary to print the name of the cities in the vector `c("Birmingham", "Derby", "Leicester", "Lincoln", "Nottingham", "Wolverhampton")` as many times as their position in the vector (i.e., once for the first city, two times for the second, and so on and so forth).

```{r, echo=TRUE, eval=FALSE}
# for each index in the vecotr cities
# i.e., 1 to length of the vector
for (iterator in 1:length(cities)) {
  # need a second iterator!
  # to print the name of the city as many 
  # times as 1 to the current value if the iterator above
  for (second_iterator in 1:iterator){
    # Let's keep the city name on the sam line
    # just addid a space between occurencies
    cat(cities[iterator], " ")
  }
  # Once we are done printing one city name
  # then go to new line
  cat("\n")
}
# After the loop delete both iterators
# just to be on the safe side :)
rm(iterator)
rm(second_iterator)
```



### Solutions 114.2 {-}

**Question 114.2.1:** Write a function that calculates the areas of a circle, taking the radius as the first parameter.

```{r, echo=TRUE, eval=FALSE}
area_of_a_circle <- function (circle_radius) {
  # the area of a circle is equal to
  # pi times radius squared
  circle_area <- pi * (circle_radius^2)
  # return the value
  circle_area
}

# Test
area_of_a_circle(3)
```

**Question 114.2.2:** Write a function that calculates the volume of a cylinder, taking the radius of the base as the first parameter and the height as the second parameter. The function should call the function defined above and multiply the returned value by the height to calculate the result.

```{r, echo=TRUE, eval=FALSE}
volume_of_a_cilinder <- function(base_radius, cilinder_height) {
  # call the function defined above to calculate the area
  base_area <- area_of_a_circle(base_radius)
  # then calculate volume
  cilinder_volume <- base_area * cilinder_height
  # return the value
  cilinder_volume
}

# Test
volume_of_a_cilinder(3, 2)
```



**Question 114.2.3:** Write a function with two parameters, a vector of numbers and a vector of characters (text). The function should check that the input has the correct data type. If all the numbers in the first vector are greater than zero, return the elements of the second vector from the first to the length of the first vector. 

```{r, echo=TRUE, eval=FALSE}
silly_function <- function (num_vector, chr_vector) {
  # Check vector types
  # Note that you can use logic operators
  # to combine conditional statements
  # and the function all
  # to check if the condition is true for all the elements of the vector
  if ( all(is.numeric(num_vector)) & all(is.character(chr_vector)) ) {
    # Check if all numbers greater than zero
    if (all(num_vector > 0)){
      # Subset the vector to return
      chr_vector[1:length(num_vector)]
    } else {
      cat("That's low...\n")
    }
  } else {
    cat("Those values seem highly inappropriate!\n")
  }
}

# Test 1
silly_function(1:3, cities)

# Test 3
silly_function(-1:1, cities)

# Test 3
silly_function(cities, 1:3)
```


-->


---

<small>by [Stefano De Sabbata](https://sdesabbata.github.io/) -- text licensed under the [CC BY-SA 4.0](https://creativecommons.org/licenses/by-sa/4.0/), contains public sector information licensed under the [Open Government Licence v3.0](http://www.nationalarchives.gov.uk/doc/open-government-licence), code licensed under the [GNU GPL v3.0](https://www.gnu.org/licenses/gpl-3.0.html).</small>